\section{Articulated rigid-body dynamics}\label{sec:articulated}

An articulated mechanism---a tree of $n+1$ rigid bodies connected by joints,
with body~$0$ as the floating base---provides a concrete test of the framework
above.
The configuration space is $Q = SE(3) \times \bT^n$, and every step of the classical recursive
dynamics algorithms is an operation on $\mathfrak{se}(3)$ or its dual
\cite{Featherstone,MurrayLiSastry}.

\medskip\noindent\textbf{(A) Configuration space and product of exponentials.}
Label the bodies $0, 1, \ldots, n$ along the kinematic tree.
Body $i$ has joint angle $\theta_i \in \bT$ and screw axis
$S_i \in \mathfrak{se}(3)$ (in body frame).
The pose of body $i$ relative to the world frame is given by the product of
exponentials:
\[
T_i(q)
= g_0 \, \bar{M}_1 \, e^{S_1 \theta_1} \cdots \bar{M}_i \, e^{S_i \theta_i},
\]
where $g_0 \in SE(3)$ is the base pose and
$\bar{M}_i \in SE(3)$ is the reference-configuration offset of joint $i$.

\medskip\noindent\textbf{(B) Recursive velocity propagation---the adjoint
representation in action.}
The body velocity of link $i$ is
$\mathcal{V}_i^b = T_i^{-1}\dot{T}_i \in \mathfrak{se}(3)$, and it
satisfies the recursion
\[
\mathcal{V}_i^b
= \mathrm{Ad}_{(e^{S_i\theta_i}\bar{M}_i)^{-1}}\,\mathcal{V}_{i-1}^b
  + S_i\,\dot{\theta}_i.
\]
The recursion is built entirely from the \textbf{adjoint representation}
$\mathrm{Ad}_g : \mathfrak{se}(3) \to \mathfrak{se}(3)$---the same object
governing the Peter--Weyl decomposition of $L^2(SE(3))$.

\medskip\noindent\textbf{(C) Spatial inertia and the Euler--Poincar\'e
equation.}
Each body $i$ carries a spatial inertia
$\cM_i \in \mathrm{Sym}^+(\mathfrak{se}(3)^*)$: a $6 \times 6$
positive-definite matrix in Pl\"ucker coordinates encoding mass, center of
mass, and rotational inertia.
The Newton--Euler equation for body $i$ in body frame is
\[
\mathcal{F}_i
= \cM_i\,\dot{\mathcal{V}}_i
  + \mathrm{ad}^*_{\mathcal{V}_i}\!(\cM_i\,\mathcal{V}_i),
\]
where $\mathrm{ad}^*$ is the coadjoint action of $\mathfrak{se}(3)$ on
$\mathfrak{se}(3)^*$.
This is exactly the \textbf{Euler--Poincar\'e equation} on $\mathfrak{se}(3)$
for a single rigid body.
For a free rigid body ($\mathcal{F} = 0$), writing
$\mu = \cM\,\mathcal{V} \in \mathfrak{se}(3)^*$ gives
$\dot{\mu} = \mathrm{ad}^*_{\mathcal{V}}\mu$; the rotational component
recovers \textbf{Euler's equation} $\dot{\Pi} = \Pi \times \omega$.

\medskip\noindent\textbf{(D) Recursive Newton--Euler Algorithm
(RNEA)---$O(n)$ inverse dynamics.}
\begin{itemize}[nosep]
\item \emph{Forward pass} (root $\to$ leaves): propagate velocities
  $\mathcal{V}_i$ and accelerations $\dot{\mathcal{V}}_i$ using $\mathrm{Ad}$.
\item \emph{Backward pass} (leaves $\to$ root): compute body wrenches
  $\mathcal{F}_i$ via Euler--Poincar\'e, propagate to parents using the
  coadjoint $\mathrm{Ad}^*$, and project onto joint axes:
  $\tau_i = S_i^T \mathcal{F}_i$.
\end{itemize}
Complexity: $O(n)$---one Euler--Poincar\'e evaluation and one $\mathrm{Ad}$
transform per body.

\noindent In Lagrangian coordinates \cite{CarpentierMansard}, the inverse dynamics reads
\begin{equation}\label{eq:lagrangian-eom}
\tau = M(q)\,\ddot{q} + C(q,\dot{q})\,\dot{q} + g(q)
  - \sum_i J_i(q)^T f_i^{\mathrm{ext}},
\end{equation}
where $M(q) \in \R^{n \times n}$ is the joint-space inertia matrix,
$C(q,\dot{q})\,\dot{q}$ collects Coriolis and centrifugal terms,
$g(q)$ is the gravity vector, and $J_i(q)$ is the frame Jacobian at
which external force $f_i^{\mathrm{ext}}$ is applied.
RNEA evaluates the right-hand side in $O(n)$; its analytical
derivatives $\partial\tau/\partial q$, $\partial\tau/\partial\dot{q}$
are likewise $O(n)$ \cite{CarpentierMansard}.

\medskip\noindent\textbf{(E) Articulated Body Algorithm (ABA)---$O(n)$
forward dynamics.}
The ABA computes the forward dynamics---inverting \eqref{eq:lagrangian-eom}
to obtain $\ddot{q}$ given $(\tau, f^{\mathrm{ext}})$---without forming
$M$ explicitly.
The key object is the \emph{articulated-body inertia} $\cM_i^A$, the
effective inertia of body $i$ and all its descendants accounting for joint
freedom.
The recursion (leaves $\to$ root) is
\[
\cM_i^A = \cM_i + \!\!\sum_{j \in \mathrm{children}(i)}
  \mathrm{Ad}_{T_{ji}}^*
  \!\left(\cM_j^A
    - \cM_j^A S_j \bigl(S_j^T \cM_j^A S_j\bigr)^{\!-1} S_j^T \cM_j^A
  \right) \mathrm{Ad}_{T_{ji}}.
\]
The inner expression
$\cM^A - \cM^A S(S^T\cM^A S)^{-1} S^T \cM^A$ is a \textbf{Schur
complement}---the projection of the inertia onto the space orthogonal to the
joint axis.
This is a Riccati recursion on $\mathrm{Sym}^+(\mathfrak{se}(3)^*)$.
Complexity: $O(n)$.

\medskip\noindent\textbf{(E$'$) Contact forces: recover then eliminate.}
When the mechanism interacts with the environment, unilateral contacts
impose constraints $\phi_k(q) \geq 0$ with associated Jacobians
$J_k^c = \partial\phi_k/\partial q$.
The contact forces $\lambda_k \geq 0$ satisfy the complementarity
condition $0 \leq \lambda_k \perp \phi_k(q) \geq 0$ \cite{MenagerCarpentier2025}.
In the Lagrangian form \eqref{eq:lagrangian-eom}, they enter as
additional external forces $f_k^{\mathrm{ext}} = \lambda_k\,\hat{n}_k$
at the contact points.

\medskip\noindent\emph{Recover.}
Substituting the forward dynamics into the contact constraint
$\ddot{\phi}_k \geq 0$ yields the \textbf{Delassus system}:
\[
G\,\lambda = J^c\, M(q)^{-1}\bigl(\tau - C\dot{q} - g\bigr)
  + \dot{J}^c\,\dot{q},
\qquad G = J^c\, M(q)^{-1} (J^c)^T.
\]
The Delassus operator $G$ is a Schur complement of the mass matrix
onto the contact space---the same algebraic structure as the
articulated-body inertia projection in part~(E).

\medskip\noindent\emph{Eliminate.}
Once $\lambda$ is recovered, substitute back:
$\ddot{q} = M^{-1}\bigl(\tau - C\dot{q} - g + (J^c)^T\lambda\bigr)$.
The contact forces are eliminated from the forward dynamics; what
remains is branchless arithmetic on $\mathfrak{se}(3)$.

\medskip\noindent\emph{Barrier smoothing.}
The complementarity condition $0 \leq \lambda \perp \phi \geq 0$ is a
hard branch.
Replacing it with a log-barrier potential
$V_{\mathrm{barrier}}(\phi) = -\varepsilon\log\phi$ yields a smooth
contact force $\lambda_\varepsilon = \varepsilon / \phi$ that
approximates the complementarity as $\varepsilon \to 0$.
(For exact treatment without relaxation, the complementarity can be
handled as a quadratic program with complementarity
constraints~\cite{MenagerCarpentier2025}.)
The MPPI estimator of \cref{prop:forward-only} samples through the barrier-smoothed
landscape without branch.

\medskip\noindent\emph{Continuous foot height from gait phase.}
The gait space is $\bT^3$: three phase offsets
$\varphi = (\varphi_{\mathrm{FL}}, \varphi_{\mathrm{RL}}, \varphi_{\mathrm{RR}})$
relative to the reference leg (FR at phase~$0$).
Each $\varphi_k \in [0,1)$ encodes the fraction of a stride period by which
leg~$k$ leads the reference.
Define the foot height schedule
\begin{equation}\label{eq:foot-height}
h_k(t;\varphi) = h_{\max}\,\Bigl(
  \tfrac{1}{2} - \tfrac{1}{2}\cos\bigl(
    2\pi\,\bigl[(t/T_s + \varphi_k) \bmod 1 - \beta\bigr]_+ / (1-\beta)
  \bigr)
\Bigr),
\end{equation}
where $T_s$ is the stride period, $\beta$ the duty factor, and $[\cdot]_+$
is the positive part.
The function $h_k$ is smooth in $\varphi$ (it is a composition of smooth
periodic functions) and captures the swing trajectory: $h_k = 0$ during
stance ($\text{cycle position} < \beta$) and $h_k > 0$ during swing.
The barrier force from the previous paragraph becomes
$\lambda_k(\varepsilon) = \varepsilon / h_k(t;\varphi)$:
large when the foot is near the ground ($h_k \to 0^+$), decaying as
the foot lifts.
No binary contact mask is needed---the force generation is continuous and
differentiable in both $\varphi$ and $\varepsilon$.
In the cost functional (steps~4--5 below), $h_k$ is regularized to
$h_k + \varepsilon$ to avoid the singularity at $h_k = 0$ during
stance (\cref{rem:regularized-barrier}).

\medskip\noindent\emph{Impedance realization.}
The barrier scale $\sigma_k = \lambda_k / (\lambda_k + 1)$ modulates a
joint-level impedance controller:
\begin{equation}\label{eq:impedance}
\tau_k = K_p(\sigma_k)\,(q_k^{\mathrm{des}} - q_k)
       - K_d(\sigma_k)\,\dot{q}_k,
\end{equation}
where the gains interpolate between stance and swing:
\[
K_p(\sigma) = K_p^{\mathrm{swing}}
  + (K_p^{\mathrm{stance}} - K_p^{\mathrm{swing}})\,\sigma, \qquad
K_d(\sigma) = K_d^{\mathrm{swing}}
  + (K_d^{\mathrm{stance}} - K_d^{\mathrm{swing}})\,\sigma.
\]
Stance legs ($\sigma_k \approx 1$) are stiff: the PD controller presses the
foot against the ground and the contact solver provides ground reaction forces
naturally.
Swing legs ($\sigma_k \approx 0$) compliantly track a foot trajectory derived
from $h_k(t;\varphi)$.
The target joint angles $q_k^{\mathrm{des}}$ blend between standing and
swing-peak configurations proportionally to the normalized foot height
$h_k / h_{\max}$.

This is an impedance controller in the classical sense: the PD creates a
virtual spring--damper at each joint, and the barrier scale $\sigma_k$
continuously modulates the mechanical impedance between rigid support
(stance) and compliant tracking (swing).
No explicit force-to-torque mapping (Jacobian transpose) is needed---the
contact physics is delegated entirely to the rigid-body simulator.
The barrier parameter $\varepsilon$ thus controls not only the contact force
distribution and the MPPI temperature, but also the impedance profile of
the physical controller.

\begin{remark}[Regularized barrier and self-consistent contact force]
\label{rem:regularized-barrier}
The standard log-barrier $-\varepsilon\log h_k$ is singular at $h_k = 0$
(stance phase, where $h_k \equiv 0$ by \eqref{eq:foot-height}).
The regularized barrier
\[
V_k^{\mathrm{reg}} = -\varepsilon\log(h_k + \varepsilon)
\]
is finite everywhere and its gradient yields the barrier scale directly:
$-\partial V_k^{\mathrm{reg}}/\partial h_k = \varepsilon/(h_k + \varepsilon) = \sigma_k$.
The identity
$-\varepsilon\log(h_k + \varepsilon) = -\varepsilon\log\varepsilon + \varepsilon\log\sigma_k$
shows that the first term is constant across gait samples (hence drops from
Boltzmann weights), while the second term $\varepsilon\log\sigma_k$ is $O(1)$
and does not amplify cost variance---ensuring that the effective sample size
remains bounded away from~$1$.
\end{remark}

\medskip\noindent\emph{PID extension and gravity compensation.}
The PD law \eqref{eq:impedance} has zero steady-state error only when the
gravitational torque vanishes---a condition violated at the thigh and calf
joints of a standing quadruped.
Add an integral term and feedforward gravity compensation:
\begin{equation}\label{eq:pid-impedance}
\tau_k = K_p(\sigma_k)\,e_k
       + K_i(\sigma_k)\,z_k
       - K_d(\sigma_k)\,\dot{q}_k
       + \sigma_k\,\tau_k^{\mathrm{grav}},
\end{equation}
where $e_k = q_k^{\mathrm{des}} - q_k$ is the position error,
$z_k(t) = \int_0^t e_k(s)\,ds$ is the integrator state, and
$\tau_k^{\mathrm{grav}}$ is the gravity compensation torque from inverse
dynamics.
The integral gain interpolates as
$K_i(\sigma) = K_i^{\mathrm{stance}}\,\sigma$: active only in stance
($\sigma \approx 1$), dormant in swing ($\sigma \approx 0$).
This mirrors industrial force controllers, where PI action eliminates
the steady-state gravity error that pure PD cannot.

\medskip\noindent\emph{B-spline generalization.}
The sinusoidal schedule \eqref{eq:foot-height} is a special case: two
effective control points (ground, peak) with a fixed trigonometric
interpolant.
More generally, represent each $h_k(t)$ as a periodic cubic B-spline with
$n_k$ control points $c_{k,1}, \ldots, c_{k,n_k} \geq 0$:
\begin{equation}\label{eq:bspline-height}
h_k(t) = \max\Bigl(0,\;\sum_{j=1}^{n_k} c_{k,j}\,B_j(t \bmod T_s)\Bigr),
\end{equation}
where $B_j$ are periodic cubic B-spline basis functions with knots at
$\{0, T_s/n_k, \ldots, (n_k{-}1)T_s/n_k\}$.
The non-negativity clamp ensures $h_k \geq 0$; the barrier
$\lambda_k = \varepsilon / h_k$ then derives the contact mode as before.

Sampling noise at the $n_k$ knot points and interpolating is equivalent to
MPPI with structured covariance
$\Sigma_u = B\,\Sigma_c\,B^T$, where $B$ is the $H \times n_k$ basis
matrix.
This restricts the search to smooth (C$^2$) trajectories---a frequency
filter that excludes unphysical high-frequency oscillations from the sample
space.
The dimension of the gait space increases from $3$ (phase offsets) to
$4n_k$ (spline coefficients), but remains far below $4H$ (pointwise),
achieving the ``complexity reduction for free'' observed empirically by
Alvarez-Padilla et al.

\medskip\noindent\emph{Joint-space renormalization.}
Define normalized coordinates centered on the standing configuration
$q_0 \in \R^n$:
\begin{equation}\label{eq:renormalization}
\tilde{q}_i = \frac{q_i - q_{0,i}}{r_i}, \qquad
r_i = \tfrac{1}{2}(q_i^{\max} - q_i^{\min}).
\end{equation}
The standing pose maps to $\tilde{q} = 0$, and the impedance target
\eqref{eq:impedance} for stance is simply $q_k^{\mathrm{des}} = 0$ in
normalized coordinates.
The scale $r_i$ is the half-range of joint~$i$; the PD gains become
uniform across joints with disparate physical ranges (e.g.\ hip
$[\pm 60^\circ]$ vs.\ thigh $[-90^\circ, +200^\circ]$), eliminating the
need for per-joint gain tuning.

\smallskip
\noindent\emph{Pitfall: non-uniform physical gains.}
In normalized coordinates, a uniform gain $K_p$ maps to physical gains
$K_p^{\mathrm{phys},i} = K_p / r_i$.
Since the half-ranges differ substantially---e.g.\ thigh
$r_{\mathrm{thigh}} \approx 2.53\;\mathrm{rad}$ vs.\ calf
$r_{\mathrm{calf}} \approx 1.42\;\mathrm{rad}$---the physical
stiffness at the thigh is $\sim\!1.8\times$ weaker than at the calf.
The thigh, which bears the dominant gravitational load, is the weakest
joint---exactly the wrong hierarchy.
The integral term in \eqref{eq:pid-impedance} compensates: it accumulates
the gravity-induced steady-state error that the proportional term alone
cannot eliminate with uniform normalized gains.

\medskip\noindent\emph{Joint-limit barrier.}
Each joint~$i$ has normalized limits $a_i < 0 < b_i$ (asymmetric when
$q_{0,i}$ is not the midpoint of $[q_i^{\min}, q_i^{\max}]$).
The joint-limit barrier potential is
\begin{equation}\label{eq:joint-barrier}
V_{\mathrm{joint}}(\tilde{q})
= -\varepsilon \sum_{i=1}^{n}
  \bigl[\log(\tilde{q}_i - a_i) + \log(b_i - \tilde{q}_i)\bigr],
\end{equation}
with gradient (repulsive torque in normalized space):
\[
\frac{\partial V_{\mathrm{joint}}}{\partial \tilde{q}_i}
= -\varepsilon\Bigl[
  \frac{1}{\tilde{q}_i - a_i} - \frac{1}{b_i - \tilde{q}_i}
\Bigr].
\]
For joints with $q_0 = q_{\mathrm{mid}}$ (symmetric limits, $a = -1$,
$b = 1$), the barrier reduces to
$V = -\varepsilon\log(1 - \tilde{q}^2)$.
The same $\varepsilon$ that controls contact sharpness, MPPI temperature,
entropic OT regularization, and impedance modulation now also controls
joint-limit softness---a \emph{quintuple} unification of a single
barrier/temperature parameter.

\medskip\noindent\emph{Anti-windup as anti-winding.}
The integrator state $z_k(t) = \int_0^t e_k\,ds$ in \eqref{eq:pid-impedance}
lives on $\R$---the universal cover of the circle.
The physical torque, however, is constrained:
$|\tau_k| \leq \tau_{\max,k}$.
When the actuator saturates, further error accumulation in $z_k$ has no
physical effect: the integrator \emph{winds} on $\R$ while the output
is pinned at $\pm\tau_{\max}$.
This is integrator windup---literally, a winding phenomenon.

Anti-windup freezes the integrator at saturation:
\begin{equation}\label{eq:anti-windup}
\dot{z}_k =
\begin{cases}
e_k & \text{if } |\tau_k| < \tau_{\max,k}, \\
0   & \text{if } |\tau_k| = \tau_{\max,k}.
\end{cases}
\end{equation}
This is the projection from the universal cover $\R$ to the realizable
set $\mathcal{Z}_k = \{z : |\tau_k(z)| \leq \tau_{\max,k}\}$---it prevents
fictitious winding-number accumulation.

The parallel with the joint-limit barrier is exact:
\begin{center}
\begin{tabular}{lll}
& \textbf{State space} & \textbf{Actuator space} \\
\hline
Constraint & $a_i < \tilde{q}_i < b_i$ & $|\tau_k| \leq \tau_{\max,k}$ \\
Smooth method & $-\varepsilon\log(\tilde{q}-a)$ (barrier) &
  $K_i \int e\,dt$ saturates smoothly \\
Hard method & hard wall at $a, b$ & clamp $\dot{z}=0$ at $\tau_{\max}$
\end{tabular}
\end{center}
The log-barrier keeps $\tilde{q}$ in the interior (smooth anti-winding in
state space); anti-windup keeps $z$ in the realizable set (hard
anti-winding in actuator space).
As $\varepsilon \to 0$, the barrier approaches the hard wall---the two
mechanisms converge to the same boundary behavior.

\medskip\noindent\emph{Time-varying feedback: Riccati inner solve.}
The impedance controller \eqref{eq:pid-impedance} uses hand-tuned gains
$K_p, K_d, K_i$.
More generally, given a gait schedule $h_k(t;\varphi)$ and the induced
barrier scales $\sigma_k(t)$, the centroidal dynamics
\[
\dot{x} = Ax + B(\sigma(t))\,u + c, \qquad
B(\sigma)_{ij} = \sigma_k(t)\,B^0_{ij}
\]
is \emph{exactly} linear (not linearized) with time-varying $B(t)$.
The state $x = (p, v, \phi, \omega, \theta) \in \R^{13}$ collects
center-of-mass position, velocity, orientation, angular velocity, and stride
phase; the control $u \in \R^{12}$ is the four-foot force vector.
The kinematic part $A_{\mathrm{kin}}$ encodes the double-integrator structure
($\dot{p} = v$, $\dot{\phi} = \omega$); the affine term $c$ encodes gravity
and the phase clock.

\medskip\noindent\emph{Viscous ground and the contact projector.}
Each foot $k$ contacts a surface with unit outward normal
$\hat{n}_k \in S^2$.
The \textbf{contact projector}
\begin{equation}\label{eq:contact-projector}
  \Pi_k = I_3 - \hat{n}_k \otimes \hat{n}_k
\end{equation}
is the orthogonal projection onto the tangent plane at the contact point:
$\Pi_k v$ extracts the component of velocity parallel to the surface.
A foot in contact ($\sigma_k \approx 1$) generates viscous friction
proportional to the tangential velocity,
\begin{equation}\label{eq:viscous-friction}
  F_{\mathrm{fric},k} = -\sigma_k\,\eta\,\Pi_k\,v,
\end{equation}
where $\eta > 0$ is the viscous ground coefficient (a material property).
The per-foot weighted sum enters the acceleration rows of $A$:
\begin{equation}\label{eq:A-viscous}
  A(t)\big|_{3{:}6,\,3{:}6}
  = -\frac{\eta}{m}\sum_{k=1}^{4}\sigma_k(t)\,\Pi_k,
\end{equation}
so that $A(t) = A_{\mathrm{kin}} + A_{\mathrm{fric}}(t)$ with the
friction block \eqref{eq:A-viscous} in the velocity--acceleration coupling.

\smallskip\noindent\emph{Flat-ground recovery.}
When the terrain is flat with vertical gravity
$\mathbf{g} = (0,0,-g)$, every contact normal is
$\hat{n}_k = \hat{e}_z$ and
$\Pi_k = \mathrm{diag}(1,1,0)$ for all $k$.
The sum collapses to
$(\sum_k \sigma_k)\,\Pi_{\mathrm{Null}(\mathbf{g})}$---friction
acts in the horizontal plane only, as expected.
The system remains \emph{exactly} linear in $x$: the coefficients
of $A(t)$ depend on $\sigma_k(t)$ and the normals $\hat{n}_k$,
not on $x$.
The Riccati recursion \eqref{eq:riccati-tvlqr} applies with
$A_d(t) = I + \Delta t\,A(t)$ at each backstep.
The barrier $-\varepsilon\log(h_k + \varepsilon)$ prevents the foot
from ``sinking''---simultaneously a complementarity relaxation and
a buoyancy bound.

\medskip\noindent\emph{Support $\oplus$ Coverage decomposition.}
At each contact foot $k$, the force space $\R^3$ decomposes as
\begin{equation}\label{eq:support-coverage}
\R^3 = \mathrm{Range}(\hat{n}_k) \;\oplus\; \mathrm{Image}(\Pi_k).
\end{equation}
The \textbf{support} component $\mathrm{Range}(\hat{n}_k)$ carries the
normal reaction force: through $B(\sigma)$, it generates the vertical
acceleration $\dot{v}_z = \sum_k \sigma_k f_{k,z}/m$ that must equal $g$
for weight support.
The \textbf{coverage} component $\mathrm{Image}(\Pi_k)$ carries the
tangential friction: through $A(t)$, it resists horizontal velocity and
provides traction for locomotion.

The finite-horizon Riccati \eqref{eq:riccati-tvlqr} with terminal cost
$P_H = Q_f$ suffers \emph{feedforward decay}: the gravity-compensating
feedforward $k_t$ weakens toward the horizon boundary.
The remedy: define time-averaged discrete dynamics
$\bar{A}_d = I + \Delta t\,\bar{A}$,
$\bar{B}_d = \Delta t\,\bar{B}$
where $\bar{A} = H^{-1}\sum_t A(t)$, $\bar{B} = H^{-1}\sum_t B(t)$.
The \textbf{steady-state DARE}
\begin{equation}\label{eq:dare-terminal}
P_\infty = Q + \bar{A}_d^\top P_\infty \bar{A}_d
  - \bar{A}_d^\top P_\infty \bar{B}_d
    (R + \bar{B}_d^\top P_\infty \bar{B}_d)^{-1}
    \bar{B}_d^\top P_\infty \bar{A}_d
\end{equation}
encodes the infinite-horizon cost-to-go.
Using $P_H = P_\infty$ as terminal cost ensures that $k_t$ provides
adequate gravity compensation at every backstep---the Riccati
``covers'' the full horizon.

Even with $P_\infty$, the Riccati may fall short of exact weight support
due to time-varying $\sigma_k(t)$.
The \textbf{support floor} enforces
$\sum_k \sigma_k(t)\,f_{k,z}(t) \geq mg$
after each forward step: if the effective vertical force is below
threshold, the deficit is distributed uniformly across active feet.
This is a projection onto the support half-space---the
$\mathrm{Range}(\hat{n}_k)$ component of
\eqref{eq:support-coverage}---and guarantees weight support regardless
of Riccati approximation quality.

\medskip\noindent\emph{Contact no-slip and the cyclic constraint.}
A stance foot planted on the ground must not slide: the foot-ground
relative velocity is zero.
In the centroidal model, foot positions are fixed in the body frame,
so the foot velocity equals the center-of-mass velocity $v$.
The \textbf{contact no-slip cost}
\begin{equation}\label{eq:no-slip}
J_{\mathrm{slip}}
= w_{\mathrm{slip}} \int_0^T \!\Bigl(\sum_k \sigma_k(t)\Bigr)
  \bigl(v_x(t)^2 + v_y(t)^2\bigr)\,dt
\end{equation}
penalizes tangential velocity during contact: the more feet in stance,
the higher the penalty.
This creates the fundamental tension that drives gait cycling:
tracking demands $v_x = v^*$, while no-slip demands $v = 0$ during
stance.
The resolution: alternate contacts so that stance duration is minimized
while maintaining thrust---a trot pays half the slip cost of full stance.

The no-slip cost makes $Q$ time-varying:
$Q(t) = Q + w_{\mathrm{slip}} \bigl(\sum_k \sigma_k(t)\bigr) E_{v_x, v_y}$,
where $E$ selects the horizontal velocity components.
The Riccati recursion \eqref{eq:riccati-tvlqr} and Joseph form
\eqref{eq:joseph-form} carry through with $Q \to Q(t)$;
the affine recursion for $p_t$ retains the original $Q$ because the
no-slip cost penalizes \emph{absolute} velocity (foot-ground slip),
not deviation from the reference.

\smallskip\noindent\emph{Integral controllability and the backflip test.}
Any pointwise contact constraint would forbid flight phases---yet a
backflip, with $B(t) = 0$ during flight, is controllable over the
full cycle.
The controllability Gramian
\[
W(0, T_s) = \int_0^{T_s}
  \Phi(0,t)\, B(\sigma(t))\, R^{-1} B(\sigma(t))^\top \Phi(0,t)^\top\, dt
\]
is the integral measure: $W \succ 0$ iff the system is controllable
over one stride, regardless of instantaneous contact loss
\cite{Raibert}.
The no-slip cost \eqref{eq:no-slip} is a practical realization of this
integral perspective: it evaluates the \emph{total} contact sliding over
a cycle, permitting flight ($\sigma_k = 0 \Rightarrow$ zero penalty)
while penalizing prolonged stance.
The support floor (\S\ref{sec:articulated}, forward simulate step)
handles weight support pointwise; the no-slip cost handles gait cycling
as an integral constraint.
Together they complete the Support $\oplus$ Coverage decomposition:
support is pointwise in $\mathrm{Range}(\hat{n}_k)$, coverage is
integral in $\mathrm{Image}(\Pi_k)$.

\begin{remark}[Three reference fields for the contact projector]
\label{rem:reference-fields}
The normal $\hat{n}_k$ admits three canonical choices:
\begin{enumerate}[nosep]
\item \emph{Gravity field}: $\hat{n}_k = -\mathbf{g}/|\mathbf{g}|$
  (global vertical).
  Flat-ground case; friction acts in the horizontal plane.
\item \emph{Contact field}: $\hat{n}_k = \hat{n}_k(x)$
  (surface normal at foot $k$'s contact point).
  The physical truth for uneven terrain: the tangent plane
  rotates with the local surface gradient.
\item \emph{Task field}: $\hat{n}_k = \hat{n}_{\mathrm{task}}$
  (an equilibrium direction defined by the task, not geometry).
  This determines which velocity components are ``free'' for
  locomotion and which are ``resisted''---functionally
  defining the locomotion subspace.
\end{enumerate}
Choices~(1) and~(2) coincide on flat ground.
Choice~(3) connects to the skew direction of gait asymmetry:
the projector $\Pi_k$ determines which velocity components accumulate
net displacement through periodic contact.
\end{remark}

\begin{remark}[Viscous ground as buoyancy]\label{rem:buoyancy}
A foot in ground contact is analogous to a hull in water.
Normal force (along $\hat{n}_k$) is buoyancy---weight
support controlled by submersion depth via $\sigma_k$.
Tangential force (in the contact plane $\Pi_k$) is viscous drag---the
medium resists relative motion, providing the friction that enables
locomotion.
The barrier scale $\sigma_k = \varepsilon/(h_k + \varepsilon)$ controls
``submersion'': deeper contact (lower $h_k$) $\Rightarrow$ higher
$\sigma_k$ $\Rightarrow$ more friction $\Rightarrow$ more available thrust.
As $\varepsilon \to 0$, the viscous model converges to binary stick/slip,
recovering Coulomb-like contact.
\end{remark}

Euler discretization gives $A_d(t) = I + \Delta t\, A(t)$,
$B_d(t) = \Delta t\, B(t)$.

The discrete Riccati recursion
\begin{equation}\label{eq:riccati-tvlqr}
K_t = (R + B_d^\top P_{t+1} B_d)^{-1} B_d^\top P_{t+1} A_d, \qquad
P_t = Q + A_d^\top P_{t+1}(A_d - B_d K_t)
\end{equation}
where $A_d = A_d(t)$ and $B_d = B_d(t)$ are both time-varying through
$\sigma_k(t)$,
yields optimal time-varying feedback $K(t) \in \R^{12 \times 13}$ and
feedforward $k(t) \in \R^{12}$ for each sampled gait.

\smallskip\noindent\emph{Joseph form and Gauss--Newton exactness.}
The $P_t$ update in \eqref{eq:riccati-tvlqr} involves a subtraction
$A_d^\top P_{t+1} A_d - K_t^\top S\, K_t$ that can destroy
positive-definiteness through floating-point cancellation.
Writing $F_t = A_d - B_d K_t$ (closed-loop dynamics), the algebraically
equivalent \textbf{Joseph form} \cite{BucyJoseph}
\begin{equation}\label{eq:joseph-form}
P_t = Q + F_t^\top\, P_{t+1}\, F_t + K_t^\top R\, K_t
\end{equation}
is a sum of three positive semi-definite matrices---manifestly
$P_t \succeq 0$ at every backstep, with no subtraction.
For nonlinear dynamics, the value-function Hessian includes
second-order terms $V_x \cdot f_{xx}$ that can spoil positive
definiteness; the \emph{Gauss--Newton approximation}
\cite{TassaErezTodorov,MastalliCrocoddyl} drops these terms,
recovering the iLQR/DDP recursion used in Crocoddyl.
In the centroidal model, the dynamics is \emph{exactly} linear
($f_{xx} = 0$), so the Gauss--Newton approximation is
exact---the Joseph form \eqref{eq:joseph-form} is the true
Hessian, not an approximation.
The optimal control is $u_t = k_t - K_t(x_t - x_t^{\mathrm{ref}})$:
$K_t$ corrects deviations from the reference trajectory, $k_t$ accounts
for gravity and the nominal contact pattern.

The MPPI outer loop samples gait schedules $h_k$ (B-spline coefficients)
and Boltzmann-reweights the \emph{policy cost}---not individual controls,
but the cost of the full closed-loop trajectory under the Riccati-derived
policy.
This is \emph{policy-level} MPPI: Monte Carlo over contact modes,
closed-form for feedback gains.
The Riccati compression of \S\ref{sec:riccati} applies exactly within
each gait sample; the non-Riccati part (gait mode selection) pays the
$O(N^{-1/2})$ CLT rate.

\medskip\noindent\emph{Dual $\varepsilon$ unification.}
The barrier potential $V_{\mathrm{barrier}} = -\varepsilon\log h_k$ and the
Hopf--Cole linearization $\psi = e^{-V/(2\varepsilon)}$ (\S\ref{sec:mppi})
share the \emph{same} parameter $\varepsilon$, and this is not a coincidence.
Both are log-transforms that convert a constrained nonlinear problem into a
linear one:
\begin{itemize}[nosep]
\item The barrier converts the complementarity constraint
  $0 \leq \lambda \perp h \geq 0$ into the unconstrained potential
  $V = -\varepsilon\log h$;
\item Hopf--Cole converts the nonlinear HJB
  $\partial_t V + q - \tfrac{1}{2}\|\nabla V\|^2 = 0$ into the linear
  backward heat equation for $\psi$.
\end{itemize}
The MPPI cost for a trajectory with gait phase $\varphi$ therefore includes
the barrier contribution:
\[
J(\varphi) = \int_0^T \!\bigl[
  q(g_t) + \tfrac{1}{2}\|u_t\|^2
  - \varepsilon\textstyle\sum_k \log\bigl(h_k(t;\varphi) + \varepsilon\bigr)
\bigr]\,dt + \Phi(g_T).
\]
The Boltzmann weight $w(\varphi) \propto e^{-J(\varphi)/\varepsilon}$
samples through the barrier-smoothed landscape: as $\varepsilon \to 0$,
forces sharpen to complementarity \emph{and} the MPPI distribution
concentrates on the optimal gait---simultaneously.
The dual ascent on $\varepsilon$ (\cref{sec:cycle}) controls both the
contact sharpness and the exploration--exploitation tradeoff in a single
scalar.

\medskip\noindent\textbf{(E$''$) Optimal transport interpretation.}
The Boltzmann reweighting of \S\ref{sec:mppi},
$w(\varphi) \propto e^{-J(\varphi)/\varepsilon}$,
has the structure of an entropic optimal transport coupling.
In Cuturi's framework \cite{Cuturi2013}, the entropic regularization of the
Wasserstein distance yields the coupling
$\pi_\varepsilon(x,y) = a(x)\,b(y)\,e^{-c(x,y)/\varepsilon}$,
where the Sinkhorn potentials $a, b$ enforce the marginal constraints.
In the MPPI setting:
\begin{itemize}[nosep]
\item the \emph{source} measure is the student-$t$ prior on $\bT^3$
  (gait phase space);
\item the \emph{target} is the desired velocity distribution
  ($\delta_{v^*}$ for tracking);
\item the \emph{cost} $c(\varphi, v^*) = J(\varphi)$ is the
  barrier-smoothed trajectory cost;
\item the \emph{regularization} $\varepsilon$ is the same dual parameter
  controlling contact sharpness and exploration.
\end{itemize}

The dual ascent on $\varepsilon$ in \cref{sec:cycle} is a variant of the
Sinkhorn algorithm: it alternates between evaluating the transport cost
(forward simulation) and adjusting the regularization to satisfy an entropy
constraint.
The compression rate
$H_{\mathrm{target}} \leftarrow \gamma\,H_{\mathrm{target}}$
tightens the constraint over iterations, analogous to the annealing
$\varepsilon \to 0$ in entropic OT \cite{Villani}.

\emph{Locomotion as constrained transport.}
A walking robot transports its center-of-mass distribution through a
contact-constrained $SE(3)$ landscape.
The gait phase $\varphi \in \bT^3$ determines \emph{how} the mass is
transported---which feet push when---and the height barrier
$V_{\mathrm{height}} = -\varepsilon\bigl(\log(p_z - z_{\min})
  + \log(z_{\max} - p_z)\bigr)$
constrains the transport plan to a feasible corridor.
The ``skew'' direction is the gait asymmetry:
the robot does not follow the geodesic in $SE(3)$ but takes a cyclic path
that accumulates net displacement through periodic contact.
The contact projector $\Pi_k$ determines which velocity components
are subject to friction and hence available for thrust; the skew
direction steers net displacement within the image of
$\sum_k \sigma_k \Pi_k$.

The dual $\varepsilon$ is now an \emph{octuple} unification:
\begin{enumerate}[nosep]
\item contact sharpness ($\lambda_k = \varepsilon/h_k$, barrier force),
\item exploration temperature ($w \propto e^{-J/\varepsilon}$,
      Boltzmann weight),
\item entropic regularization ($\pi_\varepsilon \propto e^{-c/\varepsilon}$,
      Sinkhorn coupling),
\item impedance modulation ($K_p(\sigma_k)$, stance/swing interpolation),
\item joint-limit softness ($V_{\mathrm{joint}} = -\varepsilon\log(\tilde{q}-a)
      - \varepsilon\log(b-\tilde{q})$, virtual wall),
\item integrator anti-winding ($K_i(\sigma_k) = K_i^{\mathrm{stance}}\sigma_k$:
      the integral gain activates with $\sigma_k$, coupling windup
      dynamics to the barrier scale),
\item viscous ground coupling ($F_{\mathrm{fric},k} = -\sigma_k\,\eta\,\Pi_k\,v$:
      tangential friction projected onto the per-foot contact plane
      and modulated by the barrier scale, making $A(t)$ time-varying
      through the same $\sigma_k$ that controls $B(t)$),
\item contact no-slip ($J_{\mathrm{slip}} = w_{\mathrm{slip}} \sum_k
      \sigma_k \|v_{xy}\|^2$: tangential sliding penalized in proportion
      to contact intensity, making $Q(t)$ time-varying through the same
      $\sigma_k$ that controls $A(t)$ and $B(t)$).
\end{enumerate}
All eight converge as $\varepsilon \to 0$: contact forces sharpen to
complementarity, the MPPI distribution concentrates on the optimal gait,
the entropic OT coupling converges to the Monge map,
impedance gains become binary stance/swing,
joint limits become hard constraints,
the integrator anti-windup becomes a hard saturation clamp,
tangential friction sharpens to binary stick/slip,
and the no-slip cost becomes a hard no-sliding constraint during stance.

\medskip\noindent\emph{Cost decomposition along the $\varepsilon$-path.}
Decompose the total cost as $J = J_{\mathrm{track}} + J_{\mathrm{barrier}}$,
where $J_{\mathrm{track}}$ collects the velocity and height tracking terms
and $J_{\mathrm{barrier}}$ collects all log-barrier contributions
(contact, height limits, orientation, joint limits).
Since every barrier term is proportional to $\varepsilon$,
\[
J_{\mathrm{barrier}}(\varphi;\varepsilon)
= \varepsilon \int_0^T \!\bigl[
  \textstyle\sum_k \Psi_k(g_t, \varphi)
\bigr]\,dt
= O(\varepsilon),
\]
the total cost $J$ is \emph{not} expected to decrease monotonically across
MPPI iterations: during the early dual-ascent phase, $\varepsilon$ increases
(to satisfy the entropy constraint $S \geq H_{\mathrm{target}}$), inflating
$J_{\mathrm{barrier}}$ even when $J_{\mathrm{track}}$ is flat or improving.
The cost path traces a transient hump before $\varepsilon$ begins to anneal.

This is the barrier method working as designed, not a convergence failure.
Monotonicity holds for $J_{\mathrm{track}}$ at fixed $\varepsilon$, and
for $J$ once the dual variable stabilizes.
The correct convergence diagnostic is therefore to monitor the two components
separately: $J_{\mathrm{track}}$ should decrease (the optimizer is finding
better gaits) while $J_{\mathrm{barrier}}$ follows $\varepsilon$ (the barrier
is tightening toward complementarity).

\medskip\noindent\emph{Three-level hierarchy and the Kalman dual.}
The complete control architecture separates three timescales:
\begin{enumerate}[nosep]
\item \emph{Policy level} (per-stride): MPPI samples gait schedules
  $h_k$ on $\bT^3$ (or B-spline coefficients in $\R^{4n_k}_{\geq 0}$),
  Boltzmann-reweights on trajectory cost---the combinatorial search over
  contact modes.
\item \emph{Action level} (per-timestep): two sub-mechanisms.
  \emph{Coverage}: TV-LQR derives optimal
  $K(t), k(t)$ via backward Riccati \eqref{eq:riccati-tvlqr} with
  DARE terminal cost $P_\infty$ \eqref{eq:dare-terminal}---the
  continuous optimization of feedback gains over the full horizon.
  \emph{Support}: the forward simulator enforces
  $\sum_k \sigma_k f_{k,z} \geq mg$ at each step---a projection
  onto the weight-support half-space in $\mathrm{Range}(\hat{n}_k)$.
  The centroidal foot forces $u = k_t - K_t(x - x^{\mathrm{ref}})$
  remain in the planning domain (for cost evaluation);
  physical joint torques come from the impedance controller
  \eqref{eq:pid-impedance}, modulated by $\sigma_k$.
\item \emph{Estimation level} (continuous): the Kalman--Bucy filter,
  the dual Riccati equation
  \[
  \dot{\Sigma} = A\Sigma + \Sigma A^\top
    - \Sigma C^\top W^{-1}C\Sigma + V,
  \]
  fuses IMU, joint encoders, and foot contact sensors into a state
  estimate $\hat{x}(t)$.
\end{enumerate}
Levels 2 and 3 solve the same matrix Riccati ODE with different data:
the control Riccati uses $(A, B, Q, R)$; the estimation Riccati uses
$(A, C, V, W)$ where $C$ is the sensor model and $V, W$ are process
and measurement noise covariances.
The barrier parameter $\varepsilon$ connects all three: MPPI temperature
(level~1), Riccati cost weights (level~2), and the process-noise
scale in the Kalman filter (level~3).

\medskip\noindent\emph{The optimal cycle algorithm.}
The preceding paragraphs assemble the components of the
barrier-smoothed MPPI controller.
We consolidate them into a single specification,
referencing each equation as it appears.

\smallskip
\noindent\textsc{Input.}
Robot model $(M, \mathcal{I}, r_k, q^{\min}, q^{\max},
\tau_{\max})$;
task $(v^*, z^*, H, \Delta t)$.

\smallskip
\noindent\textsc{Precompute} (constant throughout).
\begin{itemize}[nosep]
\item $A_{\mathrm{kin}} \in \R^{13 \times 13}$: double integrator
  ($\dot{p} = v$, $\dot{\phi} = \omega$).
\item $c \in \R^{13}$: gravity ($c_5 = -g$) and
  phase clock ($c_{12} = f_{\mathrm{stride}}$).
\item $c_d = \Delta t\, c$.
\item Contact normals $\hat{n}_k \in S^2$, $k = 1, \ldots, 4$
  (default: $\hat{e}_z$ for flat ground);
  projectors $\Pi_k = I_3 - \hat{n}_k \otimes \hat{n}_k$.
\item Reference trajectory
  $x^{\mathrm{ref}}_t$.
\end{itemize}

\smallskip
\noindent\textsc{Initialize.}
Gait phase $\varphi \in \bT^3$ (e.g.\ trot);
$\varepsilon \leftarrow \varepsilon_0$;
$H_{\mathrm{target}} \leftarrow \gamma_0 \log N$.

\smallskip
\noindent\textsc{Outer loop} (re-cycle, once per stride $T_s$).
\begin{enumerate}[label=\textup{\arabic*.},nosep]
\item \textbf{Sample.}
  $\varphi^{(i)} = \varphi + \delta\varphi^{(i)}$,
  \; $\delta\varphi^{(i)} \sim \cN(0, \Sigma_\varphi)$,
  \; $i = 1, \ldots, N$.

\item \textbf{Barrier bridge.}
  For each $i$ and $t = 0, \ldots, H{-}1$:
  foot height
  $h_k$ via \eqref{eq:foot-height};
  force scale
  $\sigma_k = \varepsilon / (h_k + \varepsilon)$;
  state matrix
  $A_d(t) = I + \Delta t\, A(t)$ via \eqref{eq:A-viscous};
  input matrix
  $B_d(t) = \Delta t\, B(\sigma(t))$.
  Terminal cost $P_\infty^{(i)}$ via DARE
  \eqref{eq:dare-terminal} from time-averaged
  $\bar{A}_d^{(i)}, \bar{B}_d^{(i)}$.

\item \textbf{Backward Riccati.}
  Initialize $P_H = P_\infty^{(i)}$ (coverage).
  Running cost $Q(t) = Q + w_{\mathrm{slip}}(\sum_k \sigma_k(t))\,E_{v_x,v_y}$
  (no-slip).
  For $t = H{-}1, \ldots, 0$, the recursion
  \eqref{eq:riccati-tvlqr} with $Q \to Q(t)$ yields
  $K_t^{(i)} \in \R^{12 \times 13}$,
  $k_t^{(i)} \in \R^{12}$.

\item \textbf{Forward simulate.}
  $u_t = k_t - K_t(x_t - x^{\mathrm{ref}}_t)$;
  \emph{support floor}: project onto
  $\sum_k \sigma_k f_{k,z} \geq mg$;
  then $x_{t+1} = A_d(t) x_t + B_d(t) u_t + c_d$.
  Accumulate
  \[
  J^{(i)} \;=\; \textstyle\sum_t \bigl[
    \|x_t - x^{\mathrm{ref}}_t\|_Q^2
    + \|u_t\|_R^2
    - \varepsilon \sum_k \log(h_k + \varepsilon)
  \bigr]\Delta t \;+\; \Phi(x_H).
  \]

\item \textbf{Boltzmann reweight.}
  $w^{(i)} = e^{-J^{(i)}/\varepsilon}
  \big/ {\textstyle\sum_j} e^{-J^{(j)}/\varepsilon}$.

\item \textbf{Update.}
  $\varphi \leftarrow
  \mathrm{arg}\!\bigl(\sum_i w^{(i)}
  e^{2\pi i \varphi^{(i)}}\bigr)/(2\pi)$
  \quad (circular mean on $\bT^3$).

\item \textbf{Dual ascent.}
  $S \leftarrow -\sum_i w^{(i)} \log w^{(i)}$.
  If $S < H_{\mathrm{target}}$:
  $\varepsilon \leftarrow
  \alpha_{\mathrm{up}}\,\varepsilon$;
  else
  $\varepsilon \leftarrow
  \alpha_{\mathrm{down}}\,\varepsilon$.
  Then
  $H_{\mathrm{target}} \leftarrow
  \gamma\, H_{\mathrm{target}}$.
\end{enumerate}

\smallskip
\noindent\textsc{Inner loop} (per timestep,
impedance realization).
\begin{enumerate}[label=\textup{\arabic*.},nosep,start=8]
\item \textbf{Impedance realization.}
  The barrier scale $\sigma_k$ from step~1 modulates the
  PID impedance controller \eqref{eq:pid-impedance}:
  stance legs ($\sigma_k \approx 1$) are stiff, swing legs
  ($\sigma_k \approx 0$) compliantly track $h_k$.
  The rigid-body simulator provides ground reaction forces
  through contact---no Jacobian transpose mapping is needed.
  Joint-limit barrier \eqref{eq:joint-barrier} and
  anti-windup \eqref{eq:anti-windup} apply.
\end{enumerate}

\begin{remark}[Arithmetic completeness of the algorithm]
Every operation in steps~1--7 is a finite composition of
$\exp$, $\log$, and linear algebra
(\cref{prop:forward-only}).
Step~3 is exact---not sampled: the centroidal dynamics is
affine in $x$ with time-varying $A(t)$ and $B(t)$, so the Riccati
recursion gives closed-form optimal feedback within each
gait sample (\S\ref{sec:riccati}).
MPPI pays $O(N^{-1/2})$ only for gait mode selection
over $\bT^3$.
Each outer iteration increases the effective $\nu$
and decreases $H(f)$
(\cref{prop:student-convergence}):
the re-cycle converges to the cycle.
\end{remark}

\medskip\noindent\textbf{(F) Lie--Poisson duality: velocity
$\leftrightarrow$ momentum.}
The Legendre transform $\mathbb{FL}: \mathfrak{g} \to \mathfrak{g}^*$ via the
composite inertia tensor maps body velocities to spatial momenta
\cite{MarsdenRatiu}.
Euler--Poincar\'e on $\bigoplus_i \mathfrak{se}(3)$ (velocity variables) is
dual to Lie--Poisson on $\bigoplus_i \mathfrak{se}(3)^*$ (momentum
variables):
RNEA computes the Euler--Poincar\'e side; ABA computes the Lie--Poisson side.
The inertia tensor
$\cM_i : \mathfrak{se}(3) \to \mathfrak{se}(3)^*$ is the bridge, exactly as
in the classical duality $\mathfrak{g} \leftrightarrow \mathfrak{g}^*$ of
geometric mechanics.

\medskip\noindent\textbf{(G) Connection to the spectral and
stochastic framework.}
Three connections tie the recursive dynamics to the preceding results (\cref{sec:cycle,sec:spectral}):

\begin{enumerate}[label=(\roman*)]
\item \emph{Representation-theoretic.}
The $\mathrm{Ad}$ and $\mathrm{Ad}^*$ in Featherstone's recursions are the
adjoint and coadjoint representations of $SE(3)$.
In the Peter--Weyl/Plancherel decomposition of $L^2(SE(3))$, these same
representations govern how Fourier modes transform under group action.
The tree topology constrains them to act only between parent--child
pairs---``nearest-neighbor coupling'' that enforces sparsity in the
Peter--Weyl expansion.

\item \emph{Riccati structure.}
The ABA recursion is a discrete matrix Riccati equation on
$\mathfrak{se}(3)^*$.
The per-block Riccati of the HJB (\S\ref{sec:hjb}, for left-invariant quadratic cost)
is the spectral-domain counterpart: ABA factors the Riccati in the spatial
domain (tree topology), while Peter--Weyl factors it in the spectral domain
(representation blocks).

\item \emph{Stochastic extension and MPPI.}
Adding noise to joint torques or base forces gives an SDE on
$SE(3) \times \bT^n$.
The Hopf--Cole linearization (\S\ref{sec:mppi}) converts the HJB for the
articulated system to a Feynman--Kac expectation.
With barrier-smoothed contact forces
$\lambda_k = \varepsilon/h_k(t;\varphi)$ (preceding paragraphs),
the landscape is smooth in both the gait phase
$\varphi \in \bT^3$ and the diffusion parameter $\varepsilon$:
no contact mode switching is needed.
Each forward sample of the SDE can be computed in $O(n)$ per time step using
the recursive structure---the tree factorization is preserved under
randomization.
The Monte Carlo estimator converges at the CLT rate $O(N^{-1/2})$,
independently of $n$ and $\dim G = 6 + n$:
Featherstone provides the $O(n)$ forward simulator; the CLT provides the
dimension-free convergence guarantee.
This is the MPPI framework for articulated rigid-body control.
\end{enumerate}

