\section{Forward-only reduction}\label{sec:forward-only}

\begin{proposition}[Forward-only reduction via the exponential bridge]
\label{prop:forward-only}
Consider the MPPI estimator for the controlled diffusion on a Lie group~$G$
described in \S\ref{sec:mppi}.
\begin{enumerate}[label=\textup{(\roman*)}]
\item \textbf{Arithmetic completeness.}
Every operation in the estimator is a finite composition of three primitives:
\begin{itemize}[nosep]
\item $\exp_G : \mathfrak{g} \to G$ \textup{(}Lie group exponential\textup{)},
\item $\log_G : G \to \mathfrak{g}$ \textup{(}Lie group logarithm\textup{)},
\item linear algebra \textup{(}matrix multiply-accumulate on
      $\mathfrak{g} \cong \R^d$\textup{)}.
\end{itemize}
No backward differentiation through the dynamics and no other transcendental
operations are required.
\item \textbf{Dimension-free convergence.}
The estimator converges at $O(N^{-1/2})$, independently of $\dim G$.
\end{enumerate}
\end{proposition}

\begin{proof}[Proof \textup{(}constructive enumeration\textup{)}]
We enumerate the four computational stages and verify that each uses only the
three primitives.

\medskip\noindent\emph{SDE step.}
$g_{t+h} = g_t \cdot \exp_G\!\bigl(h\,u + \sqrt{2\varepsilon h}\;\xi\bigr)$,
where $\xi \in \mathfrak{g}$ is Gaussian noise.
This requires one~$\exp_G$ and one matrix multiply-accumulate (MMA) in
$\mathfrak{g}$.

\medskip\noindent\emph{Cost evaluation.}
For a tracking cost $q(g) = \bigl\|\log_G(g^{-1}g_{\mathrm{target}})\bigr\|^2$:
one~$\log_G$ and one MMA (squared norm in~$\mathfrak{g}$).

\medskip\noindent\emph{Exponential reweighting.}
$w^{(i)} = \exp\!\bigl(-\mathrm{cost}^{(i)}/(2\varepsilon)\bigr)$:
one scalar exponential (the $\dim G = 1$ case of $\exp_G$).

\medskip\noindent\emph{Weighted average.}
$u^* = 2\varepsilon\,\sum_i w^{(i)}\xi^{(i)}\big/\sum_i w^{(i)}$:
MMA in~$\mathfrak{g}$.

\medskip\noindent
The weights $w^{(i)}$ are bounded (bounded cost, fixed~$\varepsilon$), so the
weighted average is a bounded-weight Monte Carlo estimator. The CLT
(\cref{thm:add-clt,thm:mult-clt}) gives convergence at $O(N^{-1/2})$,
independently of~$\dim G$.
\end{proof}

