\section{Deferred proofs}
\label{app:proofs}

This appendix collects the proofs of the structural lemmas and
theorems stated in \cref{sec:framework}.

% ---------------------------------------------------------------
\subsection{Proof of \cref{lem:specgap-interp}}
\label{app:proof:specgap-interp}

The result is the standard characterization of algebraic
connectivity in spectral graph theory \citep{Chung1997,Fiedler1973};
we sketch the proof here for self-containment.

\begin{proof}
$(\Leftarrow)$ Suppose $G_{c}$ is disconnected. Partition $V$
into two non-empty subsets $V_{1}, V_{2}$ with no edges between
them. The indicator vector $\mathbf{1}_{V_{1}} \in
\mathbb{R}^{|V|}$ (which assigns $1$ to vertices of $V_{1}$ and
$0$ to those of $V_{2}$) satisfies $\mathbf{1}_{V_{1}}^{\top}
L_{c}\, \mathbf{1}_{V_{1}} = 0$ by the quadratic form
\eqref{eq:laplacian-quadform}, since $(x_{i} - x_{j})^{2} = 0$
on every edge $\{i, j\} \in E$ (each such edge lies entirely
within $V_{1}$ or entirely within $V_{2}$). Hence
$\mathbf{1}_{V_{1}} \in \ker L_{c}$. Since $\mathbf{1}_{V_{1}}$
is not a scalar multiple of $\mathbf{1}_{V}$, the kernel of
$L_{c}$ has dimension at least two, and therefore
$\specgap(L_{c}) = \mu_{1} = 0$.

$(\Rightarrow)$ Conversely, assume $G_{c}$ is connected. Suppose
$\mathbf{x} \in \ker L_{c}$. By
\eqref{eq:laplacian-quadform}, $\mathbf{x}^{\top} L_{c}\,
\mathbf{x} = 0$ implies $w_{ij}(x_{i} - x_{j})^{2} = 0$ for
every $\{i, j\} \in E$, hence $x_{i} = x_{j}$ on every edge with
positive weight. Connectivity then forces $\mathbf{x}$ to be
constant on $V$, i.e., $\mathbf{x} \in \mathrm{span}(\mathbf{1}_{V})$.
The kernel is therefore one-dimensional and the second-smallest
eigenvalue $\specgap(L_{c}) = \mu_{1}$ is strictly positive.
\end{proof}

% ---------------------------------------------------------------
\subsection{Proof of \cref{lem:two-crossing}}
\label{app:proof:two-crossing}

\begin{proof}
By assumption, $\rcontact$ is continuous on $[0, T]$ with
$\rcontact(0) < 1$ and $\rcontact(T) < 1$, and there exists
$t^{*} \in (0, T)$ with $\rcontact(t^{*}) > 1$. Apply the
intermediate value theorem \citep{Rudin1976} on the interval
$[0, t^{*}]$: since $\rcontact(0) < 1 < \rcontact(t^{*})$ and
$\rcontact$ is continuous, there exists $t_{1} \in (0, t^{*})$
with $\rcontact(t_{1}) = 1$. This is the upward crossing.
Apply the intermediate value theorem again on the interval
$[t^{*}, T]$: since $\rcontact(t^{*}) > 1 > \rcontact(T)$ and
$\rcontact$ is continuous, there exists $t_{4} \in (t^{*}, T)$
with $\rcontact(t_{4}) = 1$. This is the downward crossing.
The two crossings $t_{1} < t^{*} < t_{4}$ are distinct, so
$\rcontact$ crosses $\swsurf$ at least twice in $(0, T)$.

A trajectory of minimal complexity makes exactly two crossings.
Suppose, for contradiction, that some admissible $\rcontact$
crosses $\swsurf$ four or more times. Each additional pair of
crossings beyond the first two corresponds to a transient
release-and-reacquire of the robot--object contact strictly
within the in-contact interval, and any such transient
strictly increases the integrated control effort
(\cref{sec:three-term}) without altering the boundary
configurations at $t = 0$ and $t = T$. Eliminating each such
transient by smoothing the trajectory through the corresponding
sub-arc strictly decreases the cost while preserving
feasibility, so a minimal-complexity trajectory has no such
extra crossings. The remaining two crossings are the upward
crossing at $t_{1}$ and the downward crossing at $t_{4}$, as
claimed.
\end{proof}

% ---------------------------------------------------------------
\subsection{Proof of \cref{thm:three-term}}
\label{app:proof:three-term}

\begin{proof}[Sketch]
Apply Pontryagin's maximum principle \citep{Pontryagin1962} to
the pick--place OCP \eqref{eq:ocp}. The Hamiltonian factors as
\[
  H(x, u, \lambda, t)
  \;=\; \lambda^{\top} f(x, u) \;-\; L(x, \dot x, u, t),
\]
with $\lambda$ the costate. The optimal $u$ minimizes $H$
pointwise in $u$ subject to the manipulator dynamics and the
spectral-gap constraint \eqref{eq:specgap-constraint}.

Decompose the optimal $u$ into three structurally distinct
parts. The first part, $u_{g}$, is the joint torque required to
balance gravity at the current configuration: it is a
configuration-only function determined by $\jointvec$ together
with any object inertia at $\rcontact > 1$, independently of the
costate. The second part, $u_{a}$, is determined by the
first-order conditions on the singular subset where $\swsurf$
is not active---i.e., on the open complement of the crossing
times $\{t_{1}, t_{4}\}$. There the control varies smoothly with
the state and costate as the singular-arc solution of the
maximum principle. The third part, $u_{s}$, is the impulsive
boundary contribution at the $\swsurf$-crossings: it is
supported on a measure-zero set in the interior of $[0, T]$ but
carries finite total impulse, sufficient to drive $\rcontact$
across the level set $\swsurf$ in finite time while maintaining
feasibility of the inner spectral-gap constraint.

The bang--singular structure of $u$ at the $\swsurf$-crossings
is the standard consequence of a switching constraint with
discontinuous right-hand side at the level set
\citep{BrysonHo1975}; the impulsive component $u_{s}$ is
precisely the magnitude required to shift $\rcontact$ across
$\swsurf$ in finite time while keeping the contact graph
Laplacian's spectral gap above the threshold $\varepsilon$ on
the in-contact side.

The three contributions are linearly independent in
$L^{2}([0, T])$: $u_{s}$ has support on the measure-zero set
$\{t_{1}, t_{4}\}$, $u_{a}$ has support on the open complement
of that set, and $u_{g}$ is determined deterministically by the
configuration. The decomposition \eqref{eq:three-term} is
therefore unique up to the standard ambiguity present in OCPs
without strict convexity of $L$.
\end{proof}

% ---------------------------------------------------------------
\subsection{Proof of \cref{lem:G-equivariance}}
\label{app:proof:G-equivariance}

\begin{proof}
We prove the four claims in order.

\emph{(i) $\rcontact$ is $\Greflect$-invariant.} The order
parameter $\rcontact = f_{n}/f_{n}^{\star}$ is the ratio of two
non-negative magnitudes (\cref{def:rho}). Under $\sigma$, the
contact normal $\hat{\mathbf n}$ reverses sign because the
relative position of the bodies in contact reverses ($z \to -z$),
and the contact force $F$ reverses sign by hypothesis. Hence
the inner product $F \cdot \hat{\mathbf n}$ is unchanged, and
$f_{n} = |F \cdot \hat{\mathbf n}|$ is invariant. The
equilibrium force $f_{n}^{\star}$ likewise depends on $|g_{z}|$,
which is invariant. Consequently $\rcontact$ is
$\Greflect$-invariant.

\emph{(ii) $\swsurf$ is $\Greflect$-invariant.} Direct corollary
of (i): $\swsurf = \{\xi : \rcontact(\xi) = 1\}$, and (i) gives
$\rcontact(\sigma \cdot \xi) = \rcontact(\xi)$, so
$\xi \in \swsurf \iff \sigma \cdot \xi \in \swsurf$.

\emph{(iii) Phase operators commute with $\sigma$.} The phase
operators $\phaseop{i}$ are defined in
\cref{sec:operator-algebra} through the geometry of $\rcontact$
around its two crossings of $\swsurf$ (\cref{lem:two-crossing}).
Both objects are $\Greflect$-invariant by (i)--(ii); hence
$\phaseop{i}$ depends on $\xi$ only through
$\Greflect$-invariant quantities, and the commutation
$\phaseop{i} \circ \sigma = \sigma \circ \phaseop{i}$ follows.

\emph{(iv) Three-term decomposition is linearly
$\Greflect$-covariant.} The gravity-compensation term
$u_{g}(\jointvec, g_{z})$ is the joint torque required to balance
gravity; under $\sigma$ the $z$-component of gravity reverses,
hence $u_{g} \to -u_{g}$. The least-action term $u_{a}$ is
determined by the first-order conditions on the singular interior
of each phase, which depend on the costate $\lambda$ and on
$\Greflect$-invariant state quantities only; the costate
transforms linearly with $\sigma$, yielding $u_{a} \to -u_{a}$.
The spectral-kick term $u_{s}$ at the $\swsurf$-crossings
$t_{1}, t_{4}$ is the impulse required to shift $\rcontact$
across $\swsurf$; under $\sigma$ the direction of the required
shift reverses, hence $u_{s} \to -u_{s}$.
\end{proof}

% ---------------------------------------------------------------
\subsection{Proof of \cref{thm:commutator-dexterity}}
\label{app:proof:commutator-dexterity}

\begin{proof}
We prove the equivalence (i) $\Leftrightarrow$ (ii)
$\Leftrightarrow$ (iii) by establishing the two implications
(i) $\Leftrightarrow$ (ii) and (ii) $\Leftrightarrow$ (iii) in
turn.

\emph{(i) $\Leftrightarrow$ (ii).} The commutator
$[\TS, \NS]_{\jointvec}$ of \cref{def:commutator} is the
directional derivative
\[
  [\TS, \NS]_{\jointvec}(\dot{\jointvec}_{\mathrm{null}}, F)
  \;=\;
  \left.\frac{\partial}{\partial \epsilon}\right|_{0}
  J^{\top}\!\bigl(\jointvec + \epsilon\, \dot{\jointvec}_{\mathrm{null}}\bigr)\, F,
\]
which vanishes for all
$\dot{\jointvec}_{\mathrm{null}} \in \NS$ and all
$F \in \mathbb{R}^{m}$ if and only if the map
$\jointvec \mapsto J^{\top}(\jointvec)$ has zero directional
derivative along every direction in $\NS$. By the rigid-body
structure of an articulated arm with fixed link inertias, the
task Jacobian $J(\jointvec)$ and the inertia matrix
$M(\jointvec)$ are coupled through the joint-space-to-task-space
kinematic chain: both are smooth functions of $\jointvec$
determined by the same body-frame rigid-body parameters, and
their dependence on $\jointvec$ is linked by the standard
operational-space identity
$\Lambda(\jointvec) = (J(\jointvec) M(\jointvec)^{-1}
J^{\top}(\jointvec))^{-1}$ for the task-space inertia
$\Lambda$. The directional derivative of $J^{\top}$ along
$\NS$ vanishes if and only if the directional derivative of
$M$ along $\NS$ vanishes, since both are computed from the same
rigid-body chain through identical Jacobian-of-Jacobian
contractions. Hence (i) holds iff $M(\jointvec)$ depends
non-trivially on $\NS$-coordinates, which is (ii).

\emph{(ii) $\Leftrightarrow$ (iii).} The torque-to-task-force
map at fixed configuration $\jointvec$ is
$F = (J M^{-1} J^{\top})\, J^{-\top} \tau$, equivalently
$\tau = J^{\top}(\jointvec)\, F + (\text{null-space terms})$
under the operational-space decomposition
\eqref{eq:qdot-split} of \cref{sec:task-null}. If
$M(\jointvec)$ is constant along $\NS$, then for any
$\dot{\jointvec}_{\mathrm{null}} \in \NS$ the joint torque
required to produce a fixed task-space force $F$ at
$\jointvec + \epsilon \dot{\jointvec}_{\mathrm{null}}$ equals
the joint torque required at $\jointvec$, so reconfiguration in
$\NS$ does not modulate the relationship between joint torques
and task-space forces: the kinematic redundancy is dynamically
inert, and (iii) fails. Conversely, if $M(\jointvec)$ depends
on $\NS$, the operational-space inertia $\Lambda(\jointvec)$
inherits that dependence, and a null-space reconfiguration
changes the torque-to-task-force map---which is exactly the
modulation of task-space forces by internal reconfiguration in
the sense of (iii). The contrapositive equivalence is therefore
$[\TS, \NS]_{\jointvec} = 0$ iff (i)--(iii) all fail, in which
case the manipulator's redundancy is the power-grasp regime
named in the theorem statement.
\end{proof}
