\documentclass[11pt]{article}

% ── Packages ──────────────────────────────────────────────
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage[capitalise,noabbrev]{cleveref}
\usepackage{xeCJK}
\setCJKmainfont{Songti SC}
\usepackage{tikz}

% ── Theorem environments ──────────────────────────────────
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% ── Macros ────────────────────────────────────────────────
\newcommand{\Viab}{\mathrm{Viab}}
\newcommand{\Umax}{U_{\max}}
\newcommand{\Ur}{U_r}
\newcommand{\Obs}{\mathcal{O}}
\newcommand{\R}{\mathbb{R}}

\title{The Knife Is the Mean:\\
An Agentic Theory of Viability Maintenance}
\author{Jianghan Zhang\\
\texttt{Jianghan.Zhang.gr@dartmouth.edu}\\
Dartmouth College}
\date{February 2026}

\begin{document}
\maketitle

\begin{center}
\begin{tikzpicture}[scale=0.7]
  % ── 青冥宝剑 (Green Destiny) ──
  % Blade — straight jian, double-edged
  \shade[top color=cyan!15!white, bottom color=cyan!35!gray!80!white]
    (0,-0.2) -- (0.18,0.3) -- (0.14,5.8) -- (0,6.8) --
    (-0.14,5.8) -- (-0.18,0.3) -- cycle;
  % Central ridge
  \draw[cyan!50!black, line width=0.4pt, opacity=0.7] (0,0.4) -- (0,6.6);
  % Edge highlights
  \draw[white, line width=0.2pt, opacity=0.3]
    (0.16,0.4) -- (0.13,5.7) -- (0,6.7);
  \draw[white, line width=0.2pt, opacity=0.3]
    (-0.16,0.4) -- (-0.13,5.7) -- (0,6.7);
  % Guard (护手)
  \shade[left color=yellow!80!brown!90!black,
         right color=yellow!60!brown!80!black]
    (-0.8,-0.1) rectangle (0.8,0.05);
  \fill[yellow!80!brown!90!black]
    (-0.8,-0.1) -- (-0.8,0.05) -- (-0.95,-0.025) -- cycle;
  \fill[yellow!80!brown!90!black]
    (0.8,-0.1) -- (0.8,0.05) -- (0.95,-0.025) -- cycle;
  % Handle (剑柄)
  \shade[left color=brown!45!black, right color=brown!25!black]
    (-0.11,-0.1) rectangle (0.11,-1.6);
  % Handle wrapping
  \foreach \y in {-0.3,-0.55,-0.8,-1.05,-1.3} {
    \draw[yellow!70!brown!80!black, line width=0.6pt]
      (-0.11,\y) -- (0.11,{\y+0.15});
  }
  % Pommel (剑首)
  \shade[ball color=yellow!65!brown!85!black] (0,-1.75) circle (0.16);
  % 青冥 inscription on blade
  \node[cyan!70!black, opacity=0.5, font=\tiny, rotate=90] at (0,3.2) {青冥};
\end{tikzpicture}

\bigskip

{\large 有志者事竟成，破釜沉舟百二秦关终属楚；\\[3pt]
苦心人天不负，卧薪尝胆三千越甲可吞吴。}

\medskip
{\small\itshape ---蒲松龄}
\end{center}

\bigskip

\begin{abstract}
We develop an agentic theory of viability maintenance in systems with a
principal agent (the \emph{king}) whose survival is governed by a single
axiom: the existence of a viable path to infinity. Under this axiom, we
derive a structural criterion---the \emph{knife}---that classifies any
resource as a threat based on two conditions: autonomous actuation and
observability. We prove that the knife is not an intrinsic property of a
resource but a \emph{phase function} determined by the system's mean
field. We establish three main results: (1)~a binary lifecycle theorem
(every knife is either relinquished or eliminated), (2)~a fixed-point
impossibility (no third path exists), and (3)~an unconstrained power
paradox (maximizing control forces maximizing elimination). The
framework is validated against Chinese imperial history (475~BCE--1644~CE)
and extended to the Atlantic triangular trade, the structure of
ideological hatred, parasitic network topologies, and militarism.
The central thesis is that the knife is the mean: the critical threshold
separating tools from threats is determined by the system's average
autonomous actuation level, making viability maintenance a mean-field
phenomenon.
\end{abstract}

% ══════════════════════════════════════════════════════════
\section{Introduction}\label{sec:intro}
% ══════════════════════════════════════════════════════════

Consider a system with $n$ agents $a_1, \ldots, a_n$, each possessing a
control set $U_i$ that determines what actions they can take
independently. One agent---the \emph{principal agent} or
\emph{king}---holds maximal authority: $U = \Umax$. The king's sole
objective is survival, formalized as the existence of a viable path from
every reachable state to infinity.

This paper asks: \emph{what structural features of the system force the
king to eliminate other agents?} The answer turns out to be a two-condition
criterion we call the \emph{knife}. A resource is a knife if and only if
it can actuate independently of the king \emph{and} the king can observe
it. This criterion is not chosen---it is the unique logical consequence
of the viability axiom in a multi-agent environment.

The framework yields several results that are surprising in their
precision. The lifecycle of every knife is binary: it is either
voluntarily relinquished (path~(a)) or forcibly eliminated (path~(b)).
There is no third option---attempting to ``prove your knife is not a
knife'' is a fixed-point impossibility. Unconstrained power is not
freedom but a perpetual elimination machine: the more control the king
has, the more knives he must cut.

The central thesis of this paper is that the knife is the
\emph{mean field} of the system. The threshold separating tools from
threats is not absolute but statistical: it is determined by the average
autonomous actuation level across all agents. When the system's mean
field shifts (e.g., from wartime to peacetime), the same resource
changes classification without any change in its physical properties.
This makes viability maintenance a mean-field phenomenon: the king does
not respond to individual threats but to deviations from the system mean.

We validate the framework against two millennia of Chinese imperial
history, where the viability axiom operated with unusual clarity due to
the concentration of sovereignty in a single agent. The framework
correctly classifies the fates of historical figures---Han~Xin (knife,
eliminated), Xiao~He (half-knife, blunted), Zhang~Liang (not a knife,
survived)---and explains structural phenomena such as the Qin unification
as a graph-theoretic optimization. Extensions to the Atlantic slave trade,
the structure of ideological hatred, parasitic network topologies, and
militarism demonstrate that the framework applies beyond its original
historical domain.

\paragraph{Organization.}
\Cref{sec:framework} introduces the viability axiom, the knife
definition, and the phase transition.
\Cref{sec:results} presents the main theorems.
\Cref{sec:meanfield} develops the mean-field interpretation.
\Cref{sec:applications} applies the framework to historical and
contemporary systems.
\Cref{sec:discussion} discusses the domain of applicability and the
dynamics of the population (the ``water'').

% ══════════════════════════════════════════════════════════
\section{The Framework}\label{sec:framework}
% ══════════════════════════════════════════════════════════

\subsection{The viability axiom}\label{sec:axiom}

Let $S$ denote the state space, $U$ the control set of the principal
agent, and $K \subset S$ the \emph{viability kernel}---the set of states
in which the king retains supreme authority:
\[
  K = \bigl\{\, s \in S \;:\; \text{the king retains supreme authority}
  \,\bigr\}.
\]

\begin{axiom}[Viability]\label{ax:viability}
For every state $s \in S$, there exists a viable path
$\gamma: [0, \infty) \to K$ with $\gamma(0) = s$:
\[
  \forall\, s \in S,\quad
  \exists\;\gamma: s \to \infty
  \quad\text{such that}\quad
  \gamma(t) \in \Viab(K) \;\;\forall\, t \geq 0.
\]
\end{axiom}

When $U = \Umax$, the control set is full and this path is
mathematically guaranteed to exist. The question is: \emph{what can
break it?}

In a multi-agent system with agents $a_1, \ldots, a_n$, each having
control set $U_i$, the answer is unique: an actuator whose output can
push the king's state out of $\Viab(K)$, and which can execute
\emph{independently of the king}. If the actuator's execution must pass
through the king, the king can intercept. If it can bypass the king, the
king's $\Umax$ cannot react in time.

\begin{remark}[Scope]\label{rem:scope}
This paper presents a compressed model of the topology of execution
capability under the viability axiom. It deliberately ignores culture,
personality, economics, and moral narrative in exchange for a testable
structural criterion. Historical cases are used to validate the
criterion's discriminating power, not to claim the criterion exhausts
history.
\end{remark}

\subsection{The knife}\label{sec:knife}

\begin{definition}[Knife]\label{def:knife}
A resource $r$ is a \emph{knife} if it satisfies two conditions:
\begin{enumerate}[label=(\arabic*)]
  \item \textbf{Autonomous actuation.} The resource can operate
  independently of the king. Formally, there exists an action
  $a \in \Ur$ such that the execution function $f_r$ satisfies
  \[
    f_r(s, a) \notin K
    \quad\text{and}\quad
    a \text{ does not require the king's authorization.}
  \]
  \item \textbf{Observability.} The king's detection function $\Obs$
  can observe $r$ and its execution capability:
  $r \in \mathrm{Im}(\Obs)$.
\end{enumerate}
\end{definition}

The classification is exhaustive:
\begin{itemize}
  \item Condition~(1) not satisfied: \textbf{not a knife}. (Zhang
  Liang's strategic counsel---a pure function that cannot execute
  itself.)
  \item Condition~(2) not satisfied: \textbf{hidden knife}. (More
  dangerous, but outside the king's strategy space. Unobservable
  $=$ indefensible $=$ system noise.)
  \item Both satisfied: \textbf{knife}.
\end{itemize}

\begin{remark}[Intent is irrelevant]\label{rem:intent}
The criterion tests \emph{capability}, not \emph{intention}. The king
detects whether you \emph{can} act, not whether you \emph{want to}.
Loyalty does not enter the criterion.
\end{remark}

\begin{remark}[Logical necessity]\label{rem:necessity}
These two conditions are not chosen by the modeler. They are the unique
logical consequence of the viability axiom $+$ unconstrained power $+$
multi-agent environment.
\end{remark}

\subsection{Phase transition}\label{sec:phase}

The knife is a \emph{phase function}, not an intrinsic property.

\begin{proposition}[Phase-dependent labeling]\label{prop:phase}
The same resource $r$ receives different labels under different system
phases $\varphi$:
\[
  \mathrm{Label}(r, \varphi) =
  \begin{cases}
    \textbf{tool} & \text{if } \varphi = \text{wartime (king needs }
    r\text{'s actuation),} \\
    \textbf{knife} & \text{if } \varphi = \text{peacetime (king no
    longer needs } r\text{, but } r \text{ persists).}
  \end{cases}
\]
The phase transition does not change the physical properties of $r$.
It changes the king's objective function $J(s, \varphi)$.
\end{proposition}

\begin{proof}
In wartime, the king's objective $J_{\mathrm{war}}$ includes terms
where $r$'s actuation has positive utility. In peacetime,
$J_{\mathrm{peace}}$ optimizes for long-term survival
($\exists\;\text{path to } \infty$), and the same actuation becomes a
boundary threat on $\Viab(K)$. The resource $r$ is unchanged;
the labeling function $\mathrm{Label}(r, \varphi)$ is what shifts.
\end{proof}

\subsection{The cut vertex principle}\label{sec:cutvertex}

\begin{definition}[Cut vertex]\label{def:cutvertex}
In the execution graph $G = (V, E)$ of the system, a vertex
$v \in V$ is a \emph{cut vertex} if $G \setminus \{v\}$ is
disconnected. An agent who is a cut vertex controls all execution
chains: removing them disconnects the system.
\end{definition}

\begin{theorem}[Cut vertex $\neq$ maximum actuator]\label{thm:cutvertex}
The optimal survival strategy for the king is to be a cut vertex, not
the maximum actuator. That is, the king maximizes viability by
ensuring all execution chains pass through him, rather than by
maximizing his own actuation.
\end{theorem}

\begin{proof}
A maximum actuator $v^*$ with $\|U_{v^*}\| = \max_i \|U_i\|$
suffers from three structural defects:
\begin{enumerate}[label=(\roman*)]
  \item \emph{Non-scalability}: a single actuator cannot cover the
  full state space simultaneously.
  \item \emph{Single point of failure}: $\Viab(K)$ depends entirely
  on $v^*$'s performance; one failure collapses the system.
  \item \emph{Self-referential paradox}: if the king \emph{is} the
  knife (the strongest autonomous actuator), he cannot perform
  viability maintenance on himself.
\end{enumerate}
A cut vertex $v_c$ with $\|U_{v_c}\| \approx 0$ but routing
authority over all chains avoids all three: the system is scalable
(add more actuators), fault-tolerant (one actuator's failure does
not disconnect the graph), and the king is structurally distinct
from the knives he must manage.
\end{proof}

\begin{example}[Liu Bang vs.\ Xiang Yu]\label{ex:liubang}
Xiang Yu was the strongest actuator in the late Qin system
(\emph{Shiji}: ``he could lift a bronze tripod''). His strategy:
$U = \Umax$ through personal combat. Liu Bang had near-zero
personal actuation but made himself the cut vertex of the execution
graph: Han Xin's armies needed Liu Bang's legitimacy, Xiao He's
administration needed his authorization, Zhang Liang's counsel
needed him to listen.

After the phase transition (founding of the Han dynasty), Liu Bang
executed precise viability maintenance: killed Han Xin (knife),
imprisoned then released Xiao He (blunted half-knife), left Zhang
Liang alone (not a knife). Xiang Yu, the maximum actuator, died at
Gaixia---a single actuator cannot cover the full state space.
\end{example}

\subsection{Case analysis: the three fates}\label{sec:cases}

The framework's discriminating power is tested against three figures from
the Han founding (c.~202~BCE), all subordinates of the same king (Liu
Bang), operating in the same post-unification phase:

\begin{center}
\begin{tabular}{@{}lcccc@{}}
\toprule
\textbf{Agent} & $\Ur$ & $\mathrm{Im}(\Obs)$ &
\textbf{Classification} & \textbf{Fate} \\
\midrule
Han Xin & $\neq \varnothing$ (military) & Yes & Knife &
Path~(b): eliminated \\
Xiao He & $\neq \varnothing$ (admin) & Yes & Half-knife &
Path~(a): self-blunted \\
Zhang Liang & $= \varnothing$ (counsel) & Yes & Not a knife &
Survived \\
\bottomrule
\end{tabular}
\end{center}

All three are visible ($r \in \mathrm{Im}(\Obs)$). The discriminant is
condition~(1): can the resource actuate independently?

\begin{example}[Han Xin: pure knife]\label{ex:hanxin}
Han Xin commanded armies that obeyed \emph{him}, not Liu Bang. His
execution chain was closed: he could mobilize, march, and fight without
the king's authorization. Both conditions of \cref{def:knife} satisfied.
After the phase transition, the knife criterion triggered and Liu Bang
eliminated him. Han Xin's quoted proverb (``when the hare dies, the dog
is cooked'') correctly identified path~(b) but failed to act on it---he
understood the classification but not that the only exit was path~(a).
\end{example}

\begin{example}[Xiao He: self-blunting]\label{ex:xiaohe}
Xiao He administered the capital and controlled grain supply---autonomous
actuation at the logistical level. The king observed this
($r \in \mathrm{Im}(\Obs)$), making Xiao He a knife by
\cref{def:knife}. Xiao He's response: deliberate self-corruption
(accepting bribes conspicuously). This performed two operations
simultaneously:
\begin{enumerate}[label=(\roman*)]
  \item \emph{Signal reduction}: visible moral degradation signals
  $\|U_r\| \to 0$ (an official this corrupt cannot coordinate a revolt).
  \item \emph{Mean-field alignment}: pull $\|U_r\|$ toward $\bar{U}$,
  falling below the detection threshold (\cref{thm:meanfield}).
\end{enumerate}
This is path~(a) executed through reputation rather than resignation.
\end{example}

\begin{example}[Zhang Liang: structural safety]\label{ex:zhangliang}
Zhang Liang was a strategist. Strategy is a pure function: it
\emph{advises} action but cannot \emph{execute} it. Zhang Liang's
counsel required Liu Bang's decision, Liu Bang's generals, and Liu
Bang's administration to produce any effect. Every execution chain
passed through the king (\cref{cor:breakpoint}). Result:
$\Ur = \varnothing$, condition~(1) fails, not a knife.
Zhang Liang retired and survived.
\end{example}

% ══════════════════════════════════════════════════════════
\section{Main Results}\label{sec:results}
% ══════════════════════════════════════════════════════════

\subsection{The binary lifecycle}\label{sec:lifecycle}

\begin{theorem}[Binary fate]\label{thm:lifecycle}
Every knife has exactly two possible outcomes after phase transition:
\begin{enumerate}[label=(\alph*)]
  \item \textbf{Relinquish}: the holder voluntarily sets
  $\Ur \to \varnothing$.
  \item \textbf{Elimination}: the king forces removal via
  $u^* \in \Umax$.
\end{enumerate}
There is no path~(c).
\end{theorem}

\begin{proof}
The knife criterion is $\Ur \neq \varnothing \;\wedge\;
r \in \mathrm{Im}(\Obs)$. To exit this classification, at least
one condition must fail. Condition~(2) is controlled by the king
(he can always look); only condition~(1) can be changed by the
holder. Setting $\Ur \to \varnothing$ is path~(a). If the holder
does not, the king must act (viability axiom), which is path~(b).
The classification is exhaustive.
\end{proof}

\begin{remark}[Han Xin's error]\label{rem:hanxin}
The proverb ``when the cunning hare is killed, the hunting dog is
cooked'' (\emph{Shiji}, Huaiyin Hou) conflates three structurally
distinct resources: the \emph{bow} ($\Ur = \varnothing$, tool,
``stored'' not destroyed), the \emph{dog} ($\Ur \neq \varnothing$,
actuator, ``cooked''), and the \emph{advisor} (pure function, no
actuation---Zhang Liang survived). Han Xin quoted the answer but did
not parse its fine structure.
\end{remark}

\subsection{The fixed-point impossibility}\label{sec:fixedpoint}

\begin{theorem}[No path (c)]\label{thm:fixedpoint}
There is no strategy that ``proves your knife is not a knife'' while
retaining the knife. Formally, the map
$T: \Ur \mapsto \varnothing$ conditional on $\Ur \neq \varnothing$
has no fixed point other than $\Ur = \varnothing$.
\end{theorem}

\begin{proof}
Case~1: $\Ur = \varnothing$. Then $r$ is not a knife, and no proof
is needed. $\Ur = \varnothing$ is self-certifying.
Case~2: $\Ur \neq \varnothing$. Then no speech act can set
$\Ur \to \varnothing$---the criterion tests physical capability,
not narrative. The only way to satisfy $T(U_r) = \varnothing$ is to
physically relinquish $\Ur$, which is path~(a).

Moreover, the act of proving is itself a signal: ``I need to prove
my knife is not a knife'' implies suspicion, i.e., $r$ is already
in $\mathrm{Im}(\Obs)$. The proof attempt reinforces condition~(2).
\end{proof}

\subsection{The unconstrained power paradox}\label{sec:paradox}

\begin{theorem}[Perpetual elimination]\label{thm:paradox}
$U = \Umax$ implies the king must preemptively eliminate all
observable autonomous actuators:
\[
  U = \Umax \implies
  \text{the king must preempt all } r \text{ with }
  \Ur \neq \varnothing \;\wedge\; r \in \mathrm{Im}(\Obs).
\]
The more unconstrained the king, the more knives he must cut.
\end{theorem}

\begin{proof}
$\Umax$ means the king tolerates \emph{no} autonomous actuation:
every such actuator is a boundary threat on $\Viab(K)$. A
constrained system (constitutional regime) institutionalizes knife
dynamics by installing breakpoints. An unconstrained system must
handle every knife individually. The paradox: unconstrained power is
not freedom---it is a perpetual elimination machine.
\end{proof}

\begin{proposition}[Imperfect observability accelerates the paradox]
\label{prop:imperfect}
If the detection function $\Obs$ is imperfect, the paradox
\emph{intensifies}, not weakens.
\end{proposition}

\begin{proof}
Three steps:
\begin{enumerate}[label=(\roman*)]
  \item The king knows $\Obs$ is imperfect. Hidden knives
  (\cref{def:knife}, condition~(2) unsatisfied) are more dangerous
  than visible ones. The king has motive, capability, and survival
  obligation to expand $\Obs$.
  \item Expanding $\Obs$ does not ``discover existing knives''---it
  \emph{creates new ones} definitionally. A hidden actuator
  satisfying condition~(1) but not~(2) enters $\mathrm{Im}(\Obs)$
  upon expansion $\to$ both conditions now satisfied $\to$ it
  \emph{becomes} a knife. The knife exists in the intersection
  $\Ur \neq \varnothing \;\wedge\; r \in \mathrm{Im}(\Obs)$;
  expanding $\Obs$ expands this intersection.
  \item Positive feedback:
  $U \to \Umax \implies \Obs \to \Obs_{\max}$.
  The expansion of $\Obs$ is the \emph{adjoint process} of the
  expansion of $U$. The elimination machine has two engines: the
  cutting arm ($U$) and the detecting eye ($\Obs$). They co-drive.
\end{enumerate}
Historical instances: Qin's mutual surveillance law
(\emph{lianzuo}), Han's gold-purity test (\emph{zhuo\-jin
duo\-jue}), Ming's three-layer nested monitoring (Jinyiwei
$\to$ Dongchang $\to$ Xichang---each layer itself becomes a new
knife).
\end{proof}

\begin{proposition}[$\Umax$ as attractor]\label{prop:attractor}
$\Umax$ is an attractor, not a state. No historical king achieves
literal $\Umax$, but the system dynamics point toward it:
\[
  \frac{d}{dt}\|U(t) - \Umax\| \leq 0
  \implies
  \frac{d}{dt}\bigl(\text{detected knives}\bigr) \geq 0.
\]
The paradox describes the trajectory, not the endpoint.
\end{proposition}

\subsection{The breakpoint criterion}\label{sec:breakpoint}

\begin{corollary}[Breakpoint strategy]\label{cor:breakpoint}
A resource $r$ is not a knife if and only if its execution chain
contains at least one node controlled by the king (a
\emph{breakpoint}):
\[
  r \text{ is not a knife}
  \iff
  \exists\; v \in \text{execution chain of } r
  \;\text{s.t.}\; v \text{ is controlled by the king.}
\]
\end{corollary}

\begin{proof}
If a breakpoint exists, $r$ cannot execute independently
(condition~(1) fails), so $r$ is not a knife. If no breakpoint
exists, the execution chain is closed and $r$ can actuate
autonomously, satisfying condition~(1). Combined with
observability, this makes $r$ a knife.
\end{proof}

\begin{remark}[Modern translation]\label{rem:modern}
Zhang Liang's strategy: ``ensure your capability always requires a
component you do not control.'' Liu Bang's strategy: ``become the
mandatory node in every execution chain.''
\end{remark}

% ══════════════════════════════════════════════════════════
\section{The Knife as Mean Field}\label{sec:meanfield}
% ══════════════════════════════════════════════════════════

The preceding sections defined the knife as a two-condition criterion
applied to individual resources. We now argue that the knife is
fundamentally a \emph{mean-field} phenomenon.

\subsection{The mean actuation field}

Consider $n$ agents with autonomous actuation levels
$\|U_1\|, \ldots, \|U_n\|$. Define the \emph{mean actuation field}:
\[
  \bar{U} = \frac{1}{n} \sum_{i=1}^{n} \|U_i\|.
\]

The king's detection function $\Obs$ has finite bandwidth: it cannot
monitor all agents with equal precision. In practice, $\Obs$ triggers
on agents whose actuation \emph{deviates significantly from the mean}:
\[
  r \in \mathrm{Im}(\Obs)
  \iff
  \|U_r\| - \bar{U} > \tau(\Obs),
\]
where $\tau(\Obs)$ is the detection threshold determined by the king's
observational capacity.

\begin{theorem}[The knife is the mean]\label{thm:meanfield}
The knife threshold is determined by the system's mean actuation field.
A resource $r$ is a knife if and only if:
\begin{enumerate}[label=(\roman*)]
  \item $\Ur \neq \varnothing$ (autonomous actuation exists), and
  \item $\|U_r\|$ exceeds the mean field by more than the detection
  threshold: $\|U_r\| > \bar{U} + \tau(\Obs)$.
\end{enumerate}
Consequently, the phase transition (\cref{prop:phase}) is a shift in
$\bar{U}$, not a change in any individual $\Ur$.
\end{theorem}

\begin{proof}
In wartime, many agents have high actuation (soldiers, generals,
administrators). The mean $\bar{U}$ is high, so the threshold
$\bar{U} + \tau(\Obs)$ is high: few agents exceed it. Most actuation
is \emph{expected} and therefore not flagged.

At the phase transition (end of war), most agents' actuation drops to
near zero (soldiers demobilize, wartime powers expire). The mean
$\bar{U}$ drops sharply. But agents who \emph{retain} wartime-level
actuation now exceed the new, lower threshold. The same $\|U_r\|$
that was below the wartime mean is now above the peacetime mean.

The knife is not created by the agent---it is created by the shift in
the mean. The agent's actuation is unchanged; the system's reference
frame has moved.
\end{proof}

\begin{remark}[Connection to statistical mechanics]\label{rem:statmech}
This is precisely the mechanism of a phase transition in statistical
mechanics: the order parameter (mean actuation) shifts, and
configurations that were typical in one phase become atypical---and
therefore detectable---in the other. The viability axiom plays the role
of the free energy: the system minimizes threats to $\Viab(K)$, just
as a thermodynamic system minimizes free energy.
\end{remark}

\subsection{Implications}

The mean-field interpretation resolves several puzzles:

\begin{enumerate}
  \item \textbf{Why identical resources have different fates.} Two
  generals with identical $\Ur$ can have opposite outcomes if one
  operates in a high-$\bar{U}$ environment (wartime coalition) and
  the other in a low-$\bar{U}$ environment (consolidated empire).
  The knife is relative to the mean.

  \item \textbf{Why the paradox is a feedback loop.} As the king
  eliminates knives, $\bar{U}$ drops, lowering the threshold. Agents
  who were below the old threshold now exceed the new one $\to$ new
  knives $\to$ more elimination $\to$ lower $\bar{U}$ $\to$ \ldots
  This is the positive feedback of \cref{thm:paradox}, now given a
  statistical mechanism.

  \item \textbf{Why self-blunting works.} Xiao He's strategy
  (self-corruption to signal low $\|U_r\|$) works precisely because
  the detection function triggers on \emph{deviation from the mean}.
  By visibly degrading his own actuation, Xiao He pulled $\|U_r\|$
  toward $\bar{U}$, falling below the detection threshold.

  \item \textbf{Why breakpoints prevent knives.} A breakpoint in the
  execution chain reduces $\|U_r\|$ (effective autonomous actuation)
  to below $\bar{U} + \tau(\Obs)$, since the king controls part of
  the chain. The resource remains capable but not \emph{independently}
  capable---it does not deviate from the mean.
\end{enumerate}

% ══════════════════════════════════════════════════════════
\section{Applications}\label{sec:applications}
% ══════════════════════════════════════════════════════════

\subsection{The Qin operating system}\label{sec:qin}

The Qin state (356--207~BCE) provides the first complete engineering
implementation of the framework. Shang Yang's reforms map directly to
the formal vocabulary:

\begin{center}
\begin{tabular}{@{}lp{5cm}p{5cm}@{}}
\toprule
\textbf{Policy} & \textbf{Framework equivalent} & \textbf{Effect} \\
\midrule
Abolish well-field system & Remove aristocratic $\Ur$ & Nobles
$\to$ commoners \\
Military merit ranks & $\Ur$ controlled by state (revocable) &
Actuation is rented, not owned \\
Mutual surveillance (\emph{lianzuo}) & Maximize $\Obs$ &
Neighbors $=$ distributed sensor network \\
Standardize weights \& measures & Increase $\Obs$ precision &
Higher observational resolution \\
Commandery-county system & All chains through capital &
King $=$ cut vertex \\
\bottomrule
\end{tabular}
\end{center}

Qin's unification of the six states was a graph-theoretic outcome:
Qin's star graph (center $=$ Xianyang, $O(1)$ dispatch) vs.\ the six
states' mesh graphs (multiple aristocratic centers, $O(n^2)$
coordination cost).

Shang Yang's fate validates the framework's robustness: his reform
network itself became an autonomous actuator ($\Ur \neq \varnothing$).
The system eliminated its creator---not irony, but a robustness test.
A knife-detection system that exempts its designer is not robust.

\subsubsection{The submartingale-induced unitary group}

More precisely, Shang Yang's reform sequence is a
\emph{submartingale-induced unitary group}.

\begin{definition}[Submartingale reform]\label{def:submartingale}
A reform sequence $X_0, X_1, \ldots, X_n$ is a \emph{submartingale}
if each step strictly increases the centralization index:
$\mathbb{E}[X_{k+1}] \geq X_k$, and each step is irreversible
(reversal requires undoing all subsequent steps).
\end{definition}

Shang Yang's five reforms form such a sequence:
\[
  \underbrace{\text{abolish well-fields}}_{X_0}
  \to \underbrace{\text{merit ranks}}_{X_1}
  \to \underbrace{\text{mutual surveillance}}_{X_2}
  \to \underbrace{\text{standard measures}}_{X_3}
  \to \underbrace{\text{commandery-county}}_{X_4}.
\]
Each step burns the entropy of the old feudal order irreversibly.

Once installed, the system forms a \emph{unitary group}: a
structure-preserving transformation that maintains its invariants
(king $=$ cut vertex, all knives eliminated) at every cycle, applied
to every vector in the state space without exception.

\begin{theorem}[Shang Yang's paradox]\label{thm:shangyang}
Any agent who installs a knife-elimination system via submartingale
reform must possess $\Ur \neq \varnothing$ to execute the installation.
But the submartingale is monotone: each step lowers the detection
threshold. The installer's own actuation becomes increasingly visible
with each reform step. Upon completion, the unitary group acts on the
installer:
\[
  \underbrace{\text{installer drives submartingale}}_{\text{requires }
  \Ur \neq \varnothing}
  \;\longrightarrow\;
  \underbrace{\text{unitary group forms}}_{\text{detects all }
  \Ur \neq \varnothing}
  \;\longrightarrow\;
  \underbrace{\text{group acts on installer}}_{\text{elimination}}.
\]
The system has no creator exemption.
\end{theorem}

\begin{proof}
The installer is not in the invariant subspace of the group he
created (because $\Ur \neq \varnothing$). A unitary group does not
preserve non-invariant elements---it decomposes them. But unitary
means \emph{norm-preserving}: the installer is destroyed, but his
contributions are fully conserved. The reforms---commandery-county
system, merit ranks, mutual surveillance, standard measures---persist
uniformly across the empire. The person is eliminated; not one bit of
the contribution is lost. This is not destruction but
\emph{delocalization}: a localized power entity is scattered across
the full state space by the unitary action. Norm conserved, structure
zeroed.
\end{proof}

\subsubsection{The existence proof}

Qin unified the six states, then collapsed (207~BCE, lasting only
15~years). But Qin left something more powerful than any army: an
\emph{existence proof}.

\begin{theorem}[Qin existence theorem]\label{thm:qin}
\[
  \exists\;\text{centralized state}\;\text{s.t.}\;
  \text{autonomous actuation} = 0
  \;\wedge\;
  \text{state functions.}
\]
\end{theorem}

Before Qin, no one knew a state without feudal aristocrats was
possible. Qin proved it was---and proved it functioned \emph{better}
(star-graph dispatch vs.\ mesh-graph coordination). You cannot refute
an existence proof. You can burn the paper, but the theorem persists.

After Qin fell, two men entered the ruins. Liu Bang, then a minor
official, had once seen the First Emperor's procession and sighed:
``A great man should be like this!''---\emph{I want to BE this system's
cut vertex}. Xiang Yu, upon conquering Xianyang, burned the palaces
and said: ``Who would be emperor in the dark where no one can
see?''---he wanted the \emph{trophy}, not the \emph{theorem}. One read
the existence proof. The other burned it. The one who read it founded
a dynasty that lasted four centuries.

\subsection{The dollar as knife precursor}\label{sec:dollar}

Money is not a knife (it cannot actuate independently). Money is a
\emph{knife precursor}: a universal voucher exchangeable for
$\Ur \neq \varnothing$ on the market. The king detects not how much
money you have, but whether your money has \emph{already converted}
into autonomous execution capability.

Fan Li (5th century BCE) understood this: he accumulated fortunes
three times and dispersed them twice, ensuring $\Ur$ never crossed
the threshold. Shen Wansan and Hu Xueyan did not---their wealth
formed closed execution loops, and they were eliminated.

\subsection{The Atlantic triangular trade: a knife with no cut
vertex}\label{sec:triangle}

The Atlantic Triangular Trade (16th--19th century) is a three-leg
closed execution loop: manufactured goods (Europe $\to$ West Africa),
enslaved human beings (West Africa $\to$ Americas), raw materials
(Americas $\to$ Europe). This loop satisfies \cref{def:knife}: it
actuates autonomously (multiple nations operate independent instances;
remove any one and the loop continues) and is observable (300~years of
ships, ledgers, and auction blocks).

The loop has a property stronger than condition~(1): it is
\emph{self-financing}. The raw materials extracted on leg~3 fund the
goods shipped on leg~1, which purchase the enslaved labor on leg~2,
who produce the raw materials on leg~3. A knife that funds its own
actuation cannot be starved from outside.

\paragraph{Where the isomorphism breaks.}
In \cref{sec:water}, the king--pawn relationship is sustained by water
(viability bargain: $\text{water} > 0$). In the Triangular Trade,
enslaved people had $\text{water} = 0$ from day one. Force~$F$
substituted for water---temporarily. The substitution cost: every unit
of force burns resources drawn from the extracted water on leg~3. The
loop becomes a self-financing violence machine. \Cref{sec:water}
predicts $\text{water} = 0 \implies \text{pawn} \to \text{knife}
\implies \text{collapse}$; force delays but does not prevent this.
The longer the delay, the more violent the eventual resolution.

\paragraph{Abolition as relabeling.}
Applying \cref{prop:phase}: did abolition (1807--1888) change the
\emph{physics} or only the \emph{label}? The labor force moved from
slavery to sharecropping on the same plantations, the resource flow
(raw materials $\to$ Europe) persisted, the force mechanism shifted
from slave codes to Jim Crow and convict leasing, and the laborer
still had no breakpoint. Verdict: phase transition in the sense of
\cref{sec:phase}---label changed, topology unchanged. This is
path~(c), which \cref{thm:fixedpoint} proves is not a resolution.

\begin{theorem}[Closed-loop knife]\label{thm:triangle}
Any closed execution loop that is self-financing, has no cut vertex
(mesh topology, multiple independent operators), and substitutes
force for water, has exactly two exit paths: physical dismantlement
of the loop topology [path~(a)], or system-level collapse when
force-substitution fails [path~(b)]. Relabeling [path~(c)] preserves
$\Ur \neq \varnothing$ and resolves nothing.
\end{theorem}

\paragraph{Missing cut vertex.}
The essay's framework (\cref{sec:cutvertex}) assumes a single cut
vertex. The Triangular Trade has none: its topology is a mesh, not a
star. Removing any one colonial power does not disconnect the loop.
This makes path~(a) structurally harder---there is no single point to
dismantle---but does not create a third option.

\subsection{The corrupted detection function}\label{sec:nazi}

\Cref{def:knife} uses a structural detection function
$\Obs_{\mathrm{structural}}$ that tests $\Ur$: \emph{what can the
resource do independently?} This function is identity-blind. Zhang
Liang is safe because $\Ur = \varnothing$, not because he is Zhang
Liang.

\begin{definition}[Detection function corruption]\label{def:corruption}
A detection function is \emph{corrupted} when $\Obs_{\mathrm{structural}}$
(tests $\Ur$) is replaced by $\Obs_{\mathrm{identity}}$ (tests group
membership):
\[
  \Obs_{\mathrm{identity}}:\;
  r \;\mapsto\;
  \begin{cases}
    \textbf{knife} & \text{if } \mathrm{agent}(r) \in \text{Group } X, \\
    \textbf{not knife} & \text{otherwise.}
  \end{cases}
\]
\end{definition}

The corruption produces a \emph{false positive catastrophe}: every
member of Group~$X$ with $\Ur = \varnothing$ is classified as a knife
and eliminated. Every agent outside Group~$X$ with $\Ur \neq \varnothing$
is a false negative. Six million false positives is the output of a
corrupted detection function running to completion.

\begin{definition}[Nazi structure]\label{def:nazi}
A system exhibits \emph{Nazi structure} if and only if: (1)~it performs
viability maintenance (eliminates perceived threats to $\Viab(K)$),
(2)~its detection criterion is identity-based, not structure-based,
and (3)~it executes path~(b) against agents classified by identity
who do not satisfy \cref{def:knife} ($\Ur = \varnothing$ for the
individual). These conditions are necessary and sufficient.
\end{definition}

\begin{corollary}[Identity-invariance]\label{cor:nazi}
The Nazi structure is identity-invariant on both sides: it depends
neither on the identity of the perpetrator nor on the identity of the
target. Replacing Group~$X$ with any other identity group preserves
all three conditions.
\end{corollary}

The operator that maps $\Obs_{\mathrm{structural}} \mapsto
\Obs_{\mathrm{identity}}$ is \emph{Hate}. In this framework, Hate is
not an emotion---it is an operator on detection functions with a
precise signature (structural $\to$ identity) and a precise output
(mass false positives). The motivation is irrelevant; the mapping
determines the output.

\subsection{The parasitic cut vertex}\label{sec:parasite}

\Cref{thm:cutvertex} establishes the cut vertex as the optimal
survival strategy: route all execution chains through yourself.
A \emph{parasitic cut vertex} has the same topology but opposite flow
direction: it routes access and information for extraction rather than
coordination.

\begin{definition}[Parasitic cut vertex]\label{def:parasite}
A parasitic cut vertex $v_p$ satisfies:
(1)~$G \setminus \{v_p\}$ is disconnected (cut vertex),
(2)~$v_p$ holds no formal authority ($\Ur \approx \varnothing$
officially),
(3)~$v_p$ extracts resources from both sides of the partition by
routing access across the cut.
\end{definition}

The parasitic cut vertex turns \emph{others} into pawns via the
routing function (e.g., blackmail: ``I route your secret; comply or I
re-route it publicly''). This is \cref{thm:cutvertex} inverted:
instead of building the system, the parasite extracts from it.

\begin{proposition}[Self-termination]\label{prop:parasite}
A parasitic cut vertex is self-terminating. When the secrecy that
maintains the cut ($\text{water} = \text{secrecy}$) fails, the graph
reconnects, the cut vertex property vanishes, and the former parasite
faces the combined action of all previously partitioned nodes.
\end{proposition}

Unlike the productive cut vertex (which may persist for centuries via
institutional embedding), the parasitic variant stores no structural
contribution. Upon delocalization, norm is conserved but there is
nothing to conserve: the extraction leaves no invariant.

\subsection{Militarism and the net-positive ask}\label{sec:militarism}

Militarism is the structural type where the pawn ($\Ur = \varnothing$,
the military apparatus) captures the cut vertex position, producing
$\Ur \neq \varnothing$ for the military and $\Ur \to \varnothing$ for
civilian institutions. In the framework's language: the pawn becomes the king.

\begin{example}[The Kniefall: path~(a) in 30 seconds]\label{ex:kniefall}
On December~7, 1970, West German Chancellor Willy Brandt knelt before
the Warsaw Ghetto Uprising memorial. This act satisfies four conditions
that make it a pure instance of path~(a):
\emph{physical} (a bodily act, not a speech act---cannot be retracted
by reinterpretation),
\emph{unconditional} (no negotiation, no demand for reciprocity),
\emph{performed by the cut vertex} (the head of state, the node
through which all institutional chains pass), and
\emph{irretractable} (a photograph is a permanent record).
Brandt's Kniefall is an existence proof that path~(a) is available to
any system.
\end{example}

After the Kniefall, Germany installed institutional breakpoints:
Article~1 of the \emph{Grundgesetz} (``Human dignity is inviolable''),
the Federal Constitutional Court as an independent $\Obs$, EU and NATO
membership as external breakpoints. These institutionalize the phase
transition: the system moved from $U = \Umax$ (Nazi regime) through
path~(a) (Kniefall) to a constrained system with structural breakpoints.

\begin{proposition}[Net-positive theorem]\label{prop:netpositive}
Path~(a) is strictly net-positive for all parties. Comparative evidence:
Germany (path~(a), Kniefall 1970) vs.\ Japan (path~(c), relabeling
without structural change, 80~years and counting). Every measurable
outcome---diplomatic relations, regional stability, economic
integration, soft power, domestic constitutional health---favors the
path~(a) system. The cost of path~(a) is pride. The return is
everything else.
\end{proposition}

% ══════════════════════════════════════════════════════════
\section{Discussion}\label{sec:discussion}
% ══════════════════════════════════════════════════════════

\subsection{Domain of applicability}\label{sec:domain}

The framework is a compressed model that degrades under the following
conditions:

\begin{center}
\begin{tabular}{@{}lp{5.5cm}p{5cm}@{}}
\toprule
\textbf{Condition} & \textbf{Failure mode} & \textbf{Example} \\
\midrule
Diffuse sovereignty & No unique king; cut vertex undefined & Late
European feudalism, early federalism \\
External shock dominance & $\Viab(K)$ broken by external force;
internal knife dynamics secondary & Mongol invasion, colonialism \\
Rapid ideological reshaping & $K$ itself is changing faster than
actuation reshapes state & Religious revolution, ideology \\
High-latency detection & $\Obs$ too slow; phase transition completes
before detection & Large empire frontiers \\
Universal $\Ur \approx \varnothing$ & No knives; framework trivial &
Extremely atomized societies \\
\bottomrule
\end{tabular}
\end{center}

The framework applies to systems with a unique sovereign, differentiated
agent capabilities, and sufficient observability. This covers the main
interval of Chinese imperial history but not all political forms.

\subsection{The water dynamics}\label{sec:water}

The framework so far analyzes one layer: the king--knife interaction. A
complete viability analysis requires the \emph{viability chain}---the
three-level dependency that sustains the system:
\[
  \text{King}
  \xrightarrow{\;\text{needs}\;}
  \text{Pawn}
  \xrightarrow{\;\text{needs}\;}
  \text{Water.}
\]
The \emph{pawn} is any agent with $\Ur \neq \varnothing$ whose actuation
is currently directed by the king (soldiers, administrators, tax
collectors). \emph{Water} is the population's aggregate resource
level---the substrate from which the pawn draws manpower, revenue, and
legitimacy.

\begin{definition}[Water]\label{def:water}
Water $w(t) \in [0, W_{\max}]$ is the population's aggregate extractable
resource level at time $t$. The pawn's actuation is bounded by water:
$\|\Ur\| \leq g(w)$ for some monotone function $g$ with $g(0) = 0$.
\end{definition}

The king needs the pawn to execute viability maintenance (eliminate
knives, administer territory). The pawn needs water to function. If
$w \to 0$, the pawn's actuation capacity collapses regardless of the
king's commands.

\begin{proposition}[Binary action space at $w = 0$]\label{prop:binary}
When $w(t) \to 0$, the pawn's action space collapses to a binary:
\[
  A_{\text{pawn}} = \{\text{submit},\; \text{rebel}\}.
\]
The intermediate options (negotiate, migrate, trade, accumulate) require
$w > 0$. At $w = 0$, the pawn has nothing to lose, and the viability
axiom (\cref{ax:viability}) now applies \emph{to the pawn}: the pawn's
own survival requires a viable path, and submission no longer provides
one. The pawn becomes a knife---$\Ur$ transitions from $\varnothing$ to
$\neq \varnothing$---and the king faces a knife he created by exhausting
the water.
\end{proposition}

\begin{theorem}[Du Mu's theorem]\label{thm:dumu}
Let $w(t)$ be decreasing under extraction. Then:
\[
  w(t) \to 0
  \implies
  \text{pawn} \to \text{knife}
  \implies
  \text{king absorbed.}
\]
The causal chain is internal: the system destroys itself by exhausting
its own substrate.
\end{theorem}

This is the content of Du Mu's \emph{A Fang Gong Fu} (825~CE): ``It was
not the Qin who destroyed the six states, but the six states themselves;
it was not the world that destroyed Qin, but Qin itself.'' In our
language: the states depleted their own water, creating the knives that
destroyed them. Qin, having unified, then depleted its own water
(conscription for the Great Wall and Epang Palace), creating the knives
(Chen Sheng, Wu Guang) that destroyed it.

\begin{remark}[Water as viability constraint]
``Water can carry the boat, and water can capsize the boat'' (attributed
to Wei Zheng, Tang dynasty) is not a metaphor. It is a restatement of
\cref{thm:dumu}: the substrate that enables the king's viability
($w > 0$ $\implies$ pawn functions $\implies$ king's path to $\infty$
exists) is the same substrate whose depletion destroys it ($w = 0$
$\implies$ pawn $\to$ knife $\implies$ no path to $\infty$).
\end{remark}

% ══════════════════════════════════════════════════════════
\section{Conclusion}\label{sec:conclusion}
% ══════════════════════════════════════════════════════════

We have presented an agentic theory of viability maintenance built on a
single axiom (the existence of a viable path to infinity) and a
two-condition criterion (the knife). The framework produces three main
theorems (binary lifecycle, fixed-point impossibility, unconstrained
power paradox) and a central interpretive result: the knife is the mean
field.

The knife is not an intrinsic property of a resource. It is a
statistical deviation from the system's mean autonomous actuation,
made visible by the detection function and made dangerous by the
viability axiom. Phase transitions shift the mean, not the individual.
The king responds to the mean, not to intent.

This reframing connects viability maintenance to mean-field theory in
statistical mechanics, where phase transitions are driven by shifts in
the order parameter rather than changes in individual configurations.
The viability axiom plays the role of free energy minimization; the
knife plays the role of the critical fluctuation.

Two millennia of Chinese imperial history validate the framework with
unusual clarity. The same structure appears---with instructive
breaks---in the Atlantic slave trade, ideological hatred, parasitic
network topologies, and militarism. The framework's failure conditions
(\cref{sec:domain}) are as informative as its successes: they delineate
the boundary between systems where the viability axiom operates cleanly
and systems where it is dominated by other dynamics.

The knife is the mean. Viability maintenance is a mean-field phenomenon.
The theory is agentic because the agents---not their intentions, not
their narratives, not their moral qualities, but their structural
positions in the execution graph---determine the outcome.

% ══════════════════════════════════════════════════════════
\appendix
% ══════════════════════════════════════════════════════════

\section{原典与人物}\label{app:sources}

The historical analysis in this paper draws primarily on Sima Qian's
\emph{Shiji} (《史记》, \emph{Records of the Grand Historian},
c.~94~BCE) and Du Mu's \emph{A Fang Gong Fu} (《阿房宫赋》, 825~CE).
This appendix collects the original Classical Chinese passages cited or
referenced in the main text, with English translations and brief
profiles of the historical figures.

\subsection{人物志}\label{app:persons}

\paragraph{刘邦 Liu Bang (256--195 BCE).}
Founder of the Han dynasty. Rose from minor local official (亭长) to
emperor. Near-zero personal combat ability; made himself the cut vertex
of the execution graph (\cref{ex:liubang}).

\begin{quote}
高祖为人，仁而爱人，喜施，意豁如也。常有大度。不事家人生产作业。

\medskip
\emph{Gaozu was a man of benevolence who loved people, was generous in
giving, and broad-minded. He had great magnanimity. He did not engage
in household production or labor.}
\hfill ---《史记·高祖本纪》
\end{quote}

Translation into the framework: Liu Bang himself had no actuation
capability. He could not fight, could not administer, could not
strategize. His sole structural role was as the mandatory routing node.

On first seeing the First Emperor's procession:
\begin{quote}
嗟乎，大丈夫当如此也！

\medskip
\emph{Ah, a great man should be like this!}
\hfill ---《史记·高祖本纪》
\end{quote}

Liu Bang saw the existence proof (\cref{thm:qin}) and wanted to
\emph{be} the system's cut vertex.

His self-assessment after founding the Han dynasty:
\begin{quote}
夫运筹策帷帐之中，决胜於千里之外，吾不如子房。镇国家，抚百姓，给馈饷，不绝粮道，吾不如萧何。连百万之军，战必胜，攻必取，吾不如韩信。此三者，皆人杰也，吾能用之，此吾所以取天下也。项羽有一范增而不能用，此其所以为我擒也。

\medskip
\emph{For devising strategies within a tent and securing victory a
thousand li away, I am not as good as Zhang Liang. For governing the
state, caring for the people, providing supplies, and keeping the grain
roads open, I am not as good as Xiao He. For commanding a million
soldiers, winning every battle and taking every siege, I am not as good
as Han Xin. These three are all heroes---but I can use them. This is
why I won the empire. Xiang Yu had one Fan Zeng but could not use him.
This is why he was captured by me.}
\hfill ---《史记·高祖本纪》
\end{quote}

「吾能用之」(\emph{I can use them}) is the operational definition of
the cut vertex: actuation resides in others, but routing authority
resides in Liu Bang. Every execution chain passes through him. The
character 用 (use/employ) is not metaphorical---it is a precise
description of the cut vertex's function in the execution graph.

\paragraph{项羽 Xiang Yu (232--202 BCE).}
Supreme military commander of the anti-Qin uprising. Maximum actuator
(\cref{ex:liubang}): personal combat ability unmatched in the system.

\begin{quote}
籍长八尺馀，力能扛鼎，才气过人。

\medskip
\emph{[Xiang] Ji was over eight chi tall, could lift a bronze tripod,
and his talent and spirit surpassed all others.}
\hfill ---《史记·项羽本纪》
\end{quote}

\begin{quote}
项王嗔目而叱之，赤泉侯人马俱惊，辟易数里。

\medskip
\emph{The King of Xiang glared and bellowed at him; the Marquis of
Chiquan and his horse both recoiled in terror, retreating several li.}
\hfill ---《史记·项羽本纪》
\end{quote}

On seeing the same procession Liu Bang saw:
\begin{quote}
彼可取而代也！

\medskip
\emph{That one---I can replace him!}
\hfill ---《史记·项羽本纪》
\end{quote}

Liu Bang: ``I want to \emph{be} this.'' Xiang Yu: ``I can
\emph{replace} him.'' One read the existence proof as a system to
inhabit. The other read it as a person to defeat.

Han Xin's assessment of Xiang Yu:
\begin{quote}
项王见人恭敬慈爱，言语呕呕，人有疾病，涕泣分食饮，至使人有功当封爵者，印刓敝，忍不能予。此所谓妇人之仁也。

\medskip
\emph{The King of Xiang is respectful and caring when he meets people;
his speech is warm and gentle. When someone is ill, he weeps and shares
his food and drink. But when a man has earned merit and deserves a
title, he fondles the seal until its edges are worn smooth, and still
cannot bring himself to hand it over. This is what is called the
benevolence of a woman.}
\hfill ---《史记·淮阴侯列传》
\end{quote}

「印刓敝，忍不能予」: the seal is carved and ready, rubbed smooth from
handling, yet he cannot let it go. This is not benevolence---it is the
inability to distribute actuation. A cut vertex that cannot delegate
is a maximum actuator pretending to route.

After conquering the Qin capital:
\begin{quote}
富贵不归故乡，如衣绣夜行，谁知之者！

\medskip
\emph{To be wealthy and noble but not return home is like wearing
embroidered robes at night---who would see it?}
\hfill ---《史记·项羽本纪》
\end{quote}

His last song, at Gaixia (垓下歌):
\begin{quote}
力拔山兮气盖世，时不利兮骓不逝。\\
骓不逝兮可奈何，虞兮虞兮奈若何！

\medskip
\emph{My strength could uproot mountains, my spirit overmastered the
world. / But the times turned against me, and my horse would not go. /
My horse would not go---what can be done? / Yu, oh Yu---what will
become of you?}
\hfill ---《史记·项羽本纪》
\end{quote}

At the bank of the Wu River, refusing to cross:
\begin{quote}
天之亡我，我何渡为！且籍与江东子弟八千人渡江而西，今无一人还，纵江东父兄怜而王我，我何面目见之？

\medskip
\emph{Heaven has destroyed me---why should I cross? I crossed the river
westward with eight thousand sons of Jiangdong, and not one has
returned. Even if the elders of Jiangdong pitied me and made me king,
with what face could I see them?}
\hfill ---《史记·项羽本纪》
\end{quote}

「天之亡我」(\emph{Heaven has destroyed me}). Not heaven. A single
actuator cannot cover the full state space (\cref{ex:liubang}).
Structural failure, not fate.

\paragraph{韩信 Han Xin (?--196 BCE).}
Military genius. Commanded Liu Bang's armies; conquered more territory
than any other general in the Chu--Han war. Pure knife
(\cref{ex:hanxin}): his execution chain was closed---armies obeyed him,
not Liu Bang.

On Liu Bang's ability:
\begin{quote}
陛下不能将兵，而善将将，此乃信之所以为陛下禽也。

\medskip
\emph{Your Majesty cannot command soldiers, but excels at commanding
commanders. This is why I was captured by Your Majesty.}
\hfill ---《史记·淮阴侯列传》
\end{quote}

「善将将」(\emph{excels at commanding commanders}) $=$ cut vertex
property. Liu Bang does not actuate directly; he routes the actuation
of others.

The incident that sealed his fate---requesting the title King of Qi
during wartime:
\begin{quote}
汉王大怒，骂曰：「吾困于此，旦暮望若来佐我，乃欲自立为王！」张良、陈平蹑汉王足，因附耳语……汉王亦悟……遂遣张良立信为齐王。

\medskip
\emph{The King of Han was furious and cursed: ``I am trapped here,
waiting day and night for you to come help me, and you want to make
yourself king!'' Zhang Liang and Chen Ping stepped on the king's foot
and whispered in his ear\ldots\ The King of Han understood\ldots\ and
sent Zhang Liang to install Han Xin as King of Qi.}
\hfill ---《史记·淮阴侯列传》
\end{quote}

Zhang Liang stepping on Liu Bang's foot $=$ recalibrating the search:
the viable path currently requires Han Xin's actuation (wartime phase),
so the knife cannot be cut yet. Granting the title $=$ extending the
horizon. After the phase transition, the knife was cut.

His final words:
\begin{quote}
果若人言，「狡兔死，良狗亨；高鸟尽，良弓藏；敌国破，谋臣亡。」天下已定，我固当亨！

\medskip
\emph{It is as people said: ``When the cunning hare is killed, the
hunting dog is cooked; when the high-flying birds are gone, the good
bow is stored away; when the enemy state is destroyed, the strategist
perishes.'' The empire is settled---naturally I was to be cooked!}
\hfill ---《史记·淮阴侯列传》
\end{quote}

Han Xin quoted the answer but did not parse its fine structure. See
\cref{app:proverb} for the detailed analysis.

\paragraph{萧何 Xiao He (?--193 BCE).}
Chief administrator. Controlled grain supply and the capital during Liu
Bang's campaigns. Half-knife who self-blunted via deliberate
self-corruption (\cref{ex:xiaohe}).

\begin{quote}
相国何买田宅必居穷处，为家不治垣屋。曰：「后世贤，师吾俭；不贤，毋为势家所夺。」

\medskip
\emph{Chancellor He always bought fields and houses in the poorest
locations, and did not repair the walls of his home. He said: ``If my
descendants are worthy, they will follow my example of frugality. If
they are not, the property will be too poor for powerful families to
bother seizing.''}
\hfill ---《史记·萧相国世家》
\end{quote}

The stated reason (frugality for descendants) is a cover story. The
operational function: signal to the king that $\|\Ur\| \to 0$. An
official this visibly degraded in wealth and reputation cannot
coordinate a revolt. This is path~(a) executed through reputation
rather than resignation---pulling $\|\Ur\|$ toward $\bar{U}$
(\cref{thm:meanfield}).

\paragraph{张良 Zhang Liang (?--189 BCE).}
Strategist. Descendant of five generations of prime ministers of the
state of Han. Devoted his fortune to avenging Han's destruction by Qin.
Not a knife (\cref{ex:zhangliang}): pure advisory function---every
execution chain passed through Liu Bang.

\begin{quote}
留侯乃称曰：「家世相韩，及韩灭，不爱万金之资，为韩报仇强秦，天下振动。今以三寸舌为帝者师，封万户，位列侯，此布衣之极，于良足矣。愿弃人间事，欲从赤松子游耳。」乃学辟谷。

\medskip
\emph{The Marquis of Liu said: ``My family served as ministers of Han
for five generations. When Han was destroyed, I did not begrudge ten
thousand gold to seek vengeance against mighty Qin, and the empire
trembled. Now with my three-inch tongue I have become the emperor's
teacher, been enfeoffed with ten thousand households, and ranked as
marquis. For a commoner, this is the pinnacle---it is enough for me.
I wish to abandon worldly affairs and follow the immortal Chi Songzi.''
He then took up the practice of grain abstinence.}
\hfill ---《史记·留侯世家》
\end{quote}

「三寸舌」(\emph{three-inch tongue}) $=$ pure function. No actuation.
Zhang Liang's retirement (辟谷, grain abstinence) is not path~(a)
(放下)---he had no knife to put down. It is confirmation:
$\Ur = \varnothing$ from the first day, now the function itself
is shut down.

\paragraph{商鞅 Shang Yang (?--338 BCE).}
Architect of the Qin reform system (\cref{sec:qin}). Installed the
submartingale reform sequence (\cref{thm:shangyang}) that transformed
Qin from a peripheral state into the unification engine.

\begin{quote}
令民为什伍，而相牧司连坐。不告奸者腰斩，告奸者与斩敌首同赏，匿奸者与降敌同罚。

\medskip
\emph{He organized the people into groups of five and ten households,
to watch over and be jointly liable for each other. Those who failed to
report wrongdoers were cut in half at the waist. Those who reported
wrongdoers received the same reward as those who beheaded enemy
soldiers. Those who harbored wrongdoers received the same punishment
as those who surrendered to the enemy.}
\hfill ---《史记·商君列传》
\end{quote}

This is $\Obs \to \Obs_{\max}$: neighbors as a distributed sensor
network. The reward structure ensures every agent has positive incentive
to maximize observability.

\begin{quote}
商君相秦十年，宗室贵戚多怨望者。……秦惠王车裂商君以徇，曰：「莫如商鞅反者！」遂灭商君之家。

\medskip
\emph{Lord Shang governed Qin for ten years; many among the royal
house and the powerful clans harbored resentment.\ldots\ King Hui of
Qin had Lord Shang torn apart by chariots and displayed, saying: ``Let
none rebel as Shang Yang did!'' His entire clan was exterminated.}
\hfill ---《史记·商君列传》
\end{quote}

The unitary group acts on all vectors without exception
(\cref{thm:shangyang}). The installer is not in the invariant
subspace of the group he created.

\paragraph{范蠡 Fan Li (536--448 BCE).}
Minister of Yue. After helping King Goujian destroy the state of Wu,
Fan Li immediately left (\cref{sec:dollar}). Accumulated three
fortunes, dispersed two---ensuring $\Ur$ never crossed the knife
threshold.

\begin{quote}
范蠡遂去，自齐遗大夫种书曰：「飞鸟尽，良弓藏；狡兔死，走狗烹。越王为人长颈鸟喙，可与共患难，不可与共乐。子何不去？」

\medskip
\emph{Fan Li then departed. From Qi he sent a letter to Grand Officer
Zhong, saying: ``When the birds are gone, the good bow is stored away;
when the cunning hare is killed, the hunting dog is cooked. The King
of Yue has a long neck and a bird's beak---one can share hardship with
him, but not prosperity. Why do you not leave?''}
\hfill ---《史记·越王勾践世家》
\end{quote}

This passage---written to his colleague Wen Zhong (文种), who did not
leave and was subsequently forced to commit suicide---is the origin of
the proverb Han Xin quoted two centuries later (\cref{rem:hanxin}).
See \cref{app:proverb} for the fine structure.

\begin{quote}
范蠡……乃乘扁舟浮于江湖，变名易姓……止于陶，……十九年之中三致千金，再分散与贫交疏昆弟。

\medskip
\emph{Fan Li\ldots\ took a small boat and drifted on the rivers and
lakes, changing his name\ldots\ He settled at Tao\ldots\ In nineteen
years he amassed a fortune of a thousand gold three times, and twice
distributed it to his poor friends and distant kin.}
\hfill ---《史记·货殖列传》
\end{quote}

Three accumulations, two dispersals. Each time $\Ur$ approached the
threshold, Fan Li reset it. The dollar is a knife precursor
(\cref{sec:dollar}); Fan Li ensured the precursor never converted.

\paragraph{陈胜 Chen Sheng (?--208 BCE).}
Farmer and conscript laborer. When Qin's water reached zero
(\cref{thm:dumu}), the pawn became a knife.

\begin{quote}
陈胜佐之，并杀两尉。召令徒属曰：「公等遇雨，皆已失期，失期当斩。藉第令毋斩，而戍死者固十六七。且壮士不死即已，死即举大名耳，王侯将相宁有种乎！」

\medskip
\emph{Chen Sheng helped him, and together they killed the two officers.
He gathered the conscripts and said: ``You have all been delayed by
rain and missed the deadline. The penalty for missing the deadline is
death. Even if you are not executed, six or seven out of ten who go to
garrison duty will die. Besides, when a true man dies, he dies with
his name known---are kings and nobles born to their station?''}
\hfill ---《史记·陈涉世家》
\end{quote}

「失期当斩」(\emph{miss the deadline, face execution}): $w = 0$.
Every path leads to death. The viability axiom now applies to the pawn
himself (\cref{prop:binary}). 「王侯将相宁有种乎」:
$U_{\text{pawn}}: \varnothing \to \neq\varnothing$. The breakpoint
has dissolved.

\paragraph{冯谖 Feng Xuan (3rd century BCE).}
Retainer of Lord Mengchang of Qi. Architect of the observability
reduction strategy.

\begin{quote}
冯谖曰：「狡兔有三窟，仅得免其死耳。今君有一窟，未得高枕而卧也。请为君复凿二窟。」

\medskip
\emph{Feng Xuan said: ``A cunning hare has three burrows, and barely
manages to escape death. You, my lord, have only one burrow---you
cannot yet sleep with your head high on the pillow. Allow me to dig
two more burrows for you.''}
\hfill ---《战国策·齐策四》
\end{quote}

Three burrows $=$ three alternative positions $=$ reducing the king's
detection function $\Obs$ coverage. If the king cannot observe your
$\Ur$, you move from ``knife'' to ``hidden knife''---still dangerous,
but outside the king's strategy space.

\paragraph{魏征 Wei Zheng (580--643 CE).}
Chief advisor to Emperor Taizong of Tang. Articulated the water
dynamics (\cref{sec:water}) as a political principle.

\begin{quote}
臣闻古语云：「君，舟也；人，水也。水能载舟，亦能覆舟。」陛下以为可畏，诚如圣旨。

\medskip
\emph{Your minister has heard an ancient saying: ``The ruler is a boat;
the people are the water. Water can carry the boat, and water can
capsize the boat.'' That Your Majesty considers this worthy of fear is
truly wise.}
\hfill ---《贞观政要》
\end{quote}

Water carries the boat ($w > 0 \implies$ pawn serves $\implies$ king
sovereign) and capsizes the boat ($w = 0 \implies$ pawn $\to$ knife
$\implies$ king absorbed). Wei Zheng (630~CE) and Du Mu (825~CE) stated
the same theorem (\cref{thm:dumu}). Wei Zheng gave the intuition.
Du Mu gave the proof. This paper gives the formalization.

\subsection{飞鸟尽良弓藏：韩信之误读}\label{app:proverb}

Han Xin's final words (\cref{rem:hanxin}) quote the proverb
originating from Fan Li. The proverb is treated in Chinese historical
tradition as a single lesson: ``after the war, the meritorious are
killed.'' This reading is imprecise. The proverb contains three
structurally distinct resources with three distinct fates:

\begin{center}
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Proverb} & \textbf{Resource type} & $\Ur$ &
\textbf{Fate} \\
\midrule
飞鸟尽，良弓\textbf{藏} & Bow (no autonomous actuation) &
$= \varnothing$ & \textbf{Stored}---not destroyed \\
狡兔死，走狗\textbf{烹} & Dog (autonomous actuator) &
$\neq \varnothing$ & \textbf{Cooked}---eliminated \\
敌国破，谋臣\textbf{亡} & Strategist---depends on $\Ur$ &
? & Depends on classification \\
\bottomrule
\end{tabular}
\end{center}

The bow is \emph{stored} (藏), not \emph{cooked} (烹). The proverb
itself distinguishes between the two fates at the level of the verb.
The dog---an autonomous actuator that can hunt independently---is
killed. The bow---a tool that cannot shoot itself---is merely put away.

Han Xin conflated all three. The corrected version:

\begin{itemize}
  \item \textbf{Bow} ($\Ur = \varnothing$): Zhang Liang. Retired and
  survived. A bow that stores itself.
  \item \textbf{Dog} ($\Ur \neq \varnothing$): Han Xin. Killed. A dog
  that refused to stop hunting.
  \item \textbf{Half-tool} ($\Ur$ partially non-empty): Xiao He.
  Imprisoned, then released. A dog that deliberately blunted its teeth.
\end{itemize}

The survival strategies are also encoded in the proverb:

\begin{center}
\begin{tabular}{@{}lp{5.5cm}l@{}}
\toprule
\textbf{Strategy} & \textbf{Mechanism} & \textbf{Applicable to} \\
\midrule
Don't let the birds disappear & Maintain
$J_{\text{king}}(r) > 0$ (king still needs you) & Bow
($\Ur = \varnothing$) \\
Three burrows (冯谖) & Reduce $\Obs$ coverage &
Dog ($\Ur \neq \varnothing$) \\
Self-blunting / dispersal & $\Ur \to \varnothing$ &
Any type \\
\bottomrule
\end{tabular}
\end{center}

Han Xin needed the third strategy (put down the knife). He chose the
zeroth (do nothing). The answer was in the proverb he quoted---弓
is \emph{stored}, 狗 is \emph{cooked}---but he did not parse the
distinction.

Fan Li understood this. His letter to Wen Zhong describes the bow and
the dog. His advice: become the bow (leave). But Wen Zhong was the
prime minister of Yue---his administrative network made him a dog,
not a bow. He could not leave without first setting $\Ur \to \varnothing$,
and a prime minister's administrative network is not so easily
relinquished. Wen Zhong stayed. Wen Zhong died.

\subsection{阿房宫赋}\label{app:afanggongfu}

Du Mu (杜牧, 803--852~CE) wrote the \emph{Rhapsody on the Epang
Palace} in 825~CE. \Cref{thm:dumu} formalizes its central argument.
The full text follows in traditional characters, with English
translation and framework mapping.

\subsubsection*{第一段：存在性证明的物理实例}

\begin{quote}
六王畢，四海一。蜀山兀，阿房出。覆壓三百餘里，隔離天日。驪山北構而西折，直走咸陽。二川溶溶，流入宮牆。五步一樓，十步一閣。廊腰縵迴，簷牙高啄。各抱地勢，鈎心鬬角。盤盤焉，囷囷焉，蜂房水渦，矗不知其幾千萬落。長橋臥波，未雲何龍？複道行空，不霽何虹？高低冥迷，不知西東。歌臺暖響，春光融融。舞殿冷袖，風雨淒淒。一日之內，一宮之間，而氣候不齊。

\medskip
\emph{The six kings finished, the four seas unified, the Shu mountains
stripped bare, and the Epang Palace rose. It pressed down over three
hundred li, blocking out sun and sky. From the northern foot of
Mt.~Li it turned west, running straight to Xianyang. Two rivers flowed
gently into its walls. Every five paces a tower, every ten paces a
pavilion; corridors wound and turned, eaves rose like pecking beaks;
each structure embraced the terrain, their ridges interlocking.
Spiraling and curving, like beehives and whirlpools, towering---who
knows how many thousands of clusters. The long bridge lay over the
waves---if not clouds, why a dragon? The skyway crossed the air---if
not after rain, why a rainbow? In the haze of heights and depths, one
could not tell west from east. On the singing stages, warm sounds like
spring sunlight; in the dancing halls, cold sleeves like wind and rain.
Within a single day, within a single palace, the seasons differed.}
\end{quote}

All resources converge on a single node (Xianyang). Star-graph dispatch
(\cref{sec:qin}). The palace's scale $=$ the center node's extractable
throughput, made physical.

\subsubsection*{第二段：相变后的掠夺}

\begin{quote}
妃嬪媵嬙，王子皇孫，辭樓下殿，輦來於秦。朝歌夜絃，爲秦宮人。明星熒熒，開粧鏡也。緑雲擾擾，梳曉鬟也。渭流漲膩，棄脂水也。煙斜霧橫，焚椒蘭也。雷霆乍驚，宮車過也。轆轆遠聽，杳不知其所之也。一肌一容，盡態極妍。縵立遠視，而望幸焉，有不得見者，三十六年。

\medskip
\emph{The consorts and attendants, the princes and grandsons of the six
kings, left their towers and descended their halls, riding in carriages
to Qin. Morning songs and evening strings---they became Qin's palace
women. Bright stars glittering---that was opening their mirrors. Green
clouds in disorder---that was combing their morning hair. The Wei River
rising oily---that was discarded cosmetics. Smoke slanting, mist
spreading---that was burning pepper and orchid. Thunder suddenly
startling---a palace carriage passing. Wheels rumbling into the
distance, vanishing beyond knowing. Every curve and every face brought
to its utmost beauty, standing gracefully, gazing far, hoping for the
emperor's favor. Some waited thirty-six years and never saw him.}
\end{quote}

Resources collected at the cut vertex, but the cut vertex's bandwidth
is finite. One node cannot process all chains simultaneously.
``Thirty-six years without being seen'' $=$ star-graph bottleneck.

\subsubsection*{第三段：生存公理的普遍性}

\begin{quote}
燕趙之收藏，韓魏之經營，齊楚之精英，幾世幾年，剽掠其人，倚疊如山。一旦不能有，輸來其間。鼎鐺玉石，金塊珠礫，棄擲邐迤。秦人視之，亦不甚惜。

嗟乎！一人之心，千萬人之心也。秦愛紛奢，人亦念其家。奈何取之盡錙銖，用之如泥沙！使負棟之柱，多於南畝之農夫；架梁之椽，多於機上之工女；釘頭磷磷，多於在庾之粟粒；瓦縫參差，多於周身之帛縷；直欄橫檻，多於九土之城郭；管絃嘔啞，多於市人之言語：使天下之人不敢言而敢怒。獨夫之心，日益驕固。戍卒叫，函谷舉。楚人一炬，可憐焦土。

\medskip
\emph{The treasures of Yan and Zhao, the collections of Han and Wei,
the finest goods of Qi and Chu---plundered from their people over how
many generations, piled up like mountains. One day they could keep them
no more, and all was shipped here. Tripods used as pots, jade treated
as stone, gold discarded in heaps, pearls scattered like gravel---the
Qin people saw these and did not much care.}

\emph{Alas! One man's heart is the heart of ten thousand men. Qin loved
extravagance, yet people also cherish their homes. Why take from them
down to the last coin, and spend it like mud and sand? The pillars
outnumbered the farmers; the rafters outnumbered the weavers; the
nail-heads outnumbered the grain in the granaries; the tile-seams
outnumbered the threads in a bolt of silk; the railings outnumbered the
city walls of the nine provinces; the cacophony of pipes and strings
outnumbered the speech of the marketplace. The people of the empire
dared not speak, but dared to be angry. The tyrant's heart grew daily
more arrogant and obstinate. The garrison soldiers cried out, Hangu
Pass was taken, a torch from Chu, and---alas---scorched earth!}
\end{quote}

「一人之心，千萬人之心也」$=$ the viability axiom is not the king's
exclusive property (\cref{ax:viability}). Every agent has the same
axiom. 「取之盡錙銖」$=$ $dw/dt \ll 0$ (\cref{thm:dumu}).
「不敢言而敢怒」$=$ $\Ur = \varnothing$ still (not speaking $=$ no
autonomous actuation), but energy accumulates. 「獨夫之心，日益驕固」
$=$ the cut vertex receives no feedback---all correction channels have
been eliminated. 「戍卒叫，函谷舉」$=$ $U_{\text{pawn}}: \varnothing
\to \neq\varnothing$ (\cref{prop:binary}). Phase transition fires.

\subsubsection*{第四段：定理}

\begin{quote}
嗚呼！\textbf{滅六國者，六國也，非秦也。族秦者，秦也，非天下也。}嗟夫！使六國各愛其人，則足以拒秦。使秦復愛六國之人，則遞三世可至萬世而爲君，誰得而族滅也。\textbf{秦人不暇自哀，而後人哀之。後人哀之，而不鑑之，亦使後人而復哀後人也。}

\medskip
\emph{Alas! It was the six states that destroyed the six states, not
Qin. It was Qin that destroyed Qin, not the world. Had the six states
each loved their own people, they would have had enough to resist Qin.
Had Qin, in turn, loved the people of the six states, it could have
passed from the third generation to the ten-thousandth and remained
sovereign---who could have destroyed it? The people of Qin had no
leisure to mourn for themselves, and later generations mourned for them.
But if later generations mourn them without learning from them, they
will only cause yet later generations to mourn for the later
generations in turn.}
\end{quote}

This is \cref{thm:dumu} in prose. 「滅六國者六國也」$=$ the six states
depleted their own water (internal knife dynamics, coordination cost
$O(n^2)$). 「族秦者秦也」$=$ $w(t) \to 0 \implies \text{pawn} \to
\text{knife} \implies \text{king absorbed}$. The causal chain is
internal.

「後人哀之而不鑑之」: the theorem is time-invariant. It does not care
about dynasty names, centuries, or regime labels. It checks three
conditions: is $\Ur \neq \varnothing$? Is the loop closed? Is water
being maintained? Du Mu did not say ``you are Qin.'' He said: check
the premises.

\begin{thebibliography}{9}
\bibitem{aubin} J.-P.~Aubin, \emph{Viability Theory}, Birkh\"auser,
1991.
\bibitem{shiji} Sima Qian, \emph{Shiji} (Records of the Grand
Historian), c.~94~BCE.
\bibitem{dumu} Du Mu, \emph{A Fang Gong Fu} (Rhapsody on the Epang
Palace), 825~CE.
\bibitem{diestel} R.~Diestel, \emph{Graph Theory}, 5th ed., Springer,
2017.
\end{thebibliography}

\end{document}
