% sections/discussion.tex
% Note: this paper uses \sfrac{a}{b} from xfrac, NOT \nicefrac{a}{b} (nicefrac.sty unavailable).
\section{Discussion}
\label{sec:discussion}

\paragraph{What this paper does and does not show.}
If the experiments succeed, \textsc{MinPG} is evidence that the
standard policy-gradient tricks are not all individually necessary on
Walker2d-v4, given a heavy-tailed policy and a margin-based
objective. It is not evidence that they are unnecessary in general,
in higher dimensions, or in sparse-reward tasks where the margin
formulation degrades into a constant.

\paragraph{When the mollifier fails.}
The smoothness in Lemma~\ref{lem:smoothness} relies on a bounded
Lipschitz constant $L_Q$. Tasks where the per-step margin can be
unboundedly negative -- for example, an explicit ``fall and reset''
penalty spike -- violate this assumption. The soft-min handles mild
non-smoothness but not unbounded jumps.

\paragraph{Variance vs.\ value functions.}
Lemma~\ref{lem:bounded-score} bounds variance through the policy and
the bounded objective; it does not eliminate variance. At very large
horizons the variance still grows linearly in $H$. A control variate
(state-dependent baseline) would reduce this further; we do not adopt
one because it would reintroduce the value-function machinery that
\textsc{MinPG} is designed to avoid.

\paragraph{Limitations.}
\textsc{MinPG} requires (i) a black-box simulator that supports many
parallel rollouts, (ii) margins definable purely from the simulator
state without learned components, and (iii) bounded torque actuation
(the truncation in~\eqref{eq:mollifier}). It does not address
sim-to-real transfer, sparse-reward exploration, or partial
observability beyond what proprioception provides.
