% sections/experiments.tex
% Note: this paper uses \sfrac{a}{b} from xfrac, NOT \nicefrac{a}{b} (nicefrac.sty unavailable).
\section{Experiments}
\label{sec:exp}

\subsection{Setup}
\label{sec:exp-setup}

\paragraph{Environment.}
Walker2d-v4 from Gymnasium~\cite{gymnasium2023}, MuJoCo
backend~\cite{todorov2012mujoco}: 17-dimensional observation (joint
angles, velocities, torso pose), 6-dimensional torque action,
episode length $H = 1000$, control frequency $25$\,Hz.

\paragraph{Margins.}
We instantiate the soft-min objective~\eqref{eq:soft-margin} with
$k = 3$ margins read from the simulator state at each step:
$m_1 = z_{\mathrm{torso}} - 0.8$ (height above floor in metres),
$m_2 = \cos\theta_{\mathrm{pitch}} - 0.7$ (upright orientation), and
$m_3 = \min_i z_i^{\mathrm{foot}} + \epsilon$ (no foot penetration,
with $\epsilon = 3$\,mm). No reward shaping beyond these three
quantities.

\paragraph{Baselines.}
PPO~\cite{schulman2017ppo} and SAC~\cite{haarnoja2018sac} from
Stable-Baselines3 with default hyperparameters; a vanilla
REINFORCE-Gaussian control. All methods use the same MLP backbone
($L = 3$, $n = 256$) and are run for five random seeds.

\paragraph{Metrics.}
Episodic return at $\{0.5, 1, 2, 5\} \times 10^6$ environment steps.
For \textsc{MinPG} we additionally report the soft-min margin
$J_\beta(\theta)$ and a viability rate (fraction of episodes
reaching $H = 1000$ steps without termination).

\subsection{Main results}
\label{sec:exp-main}

\begin{table}[h]
\centering
\caption{Walker2d-v4 episodic return at four sample budgets, mean
$\pm$ standard deviation over 5 random seeds. \emph{Numbers pending
experiments.}}
\label{tab:main}
\begin{tabular}{lcccc}
\toprule
Method & 0.5M & 1M & 2M & 5M \\
\midrule
PPO (SB3 default) & TBD & TBD & TBD & TBD \\
SAC (SB3 default) & TBD & TBD & TBD & TBD \\
REINFORCE-Gaussian & TBD & TBD & TBD & TBD \\
\textsc{MinPG} (ours) & TBD & TBD & TBD & TBD \\
\bottomrule
\end{tabular}
\end{table}

The hypothesis to be tested: \textsc{MinPG} matches or exceeds PPO at
$\ge\!2$M steps, loses to SAC in sample efficiency at low budgets due
to on-policy variance, and decisively beats REINFORCE-Gaussian thanks
to the mollifier (Lemma~\ref{lem:smoothness}).

\subsection{Ablations}
\label{sec:exp-abl}

We ablate three orthogonal axes: policy family (Student-$t$ vs.\
Gaussian), training signal (soft-min margin vs.\ Walker2d default
return), and capacity clamp (with vs.\ without). Eight configurations,
five seeds each.

\begin{table}[h]
\centering
\caption{Ablation grid: episodic return at 2M steps. \emph{Pending.}}
\label{tab:ablation}
\begin{tabular}{lcc}
\toprule
Configuration & Soft-min reward & Task return \\
\midrule
Student-$t$ + clamp & TBD & TBD \\
Student-$t$, no clamp & TBD & TBD \\
Gaussian + clamp & TBD & TBD \\
Gaussian, no clamp & TBD & TBD \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Sensitivity}
\label{sec:exp-sens}

We sweep the four Tier-II hyperparameters of the
algorithm~(Section~\ref{sec:hyper})
$L \in \{3, 4, 5\}$, $n \in \{128, 256, 512\}$,
$N \in \{256, 1024, 4096\}$, $\beta \in \{1, 10, 100\}$
one at a time around a centre point $(L, n, N, \beta) =
(3, 256, 1024, 10)$. The Tier-I parameters
($\eta$, $H$, $\{c_\ell\}$) are read from the environment XML and
held fixed.
