\begin{thebibliography}{99}

\bibitem{Aubin}
J.-P.~Aubin,
\emph{Viability Theory}, new ed.,
Modern Birkh\"auser Classics, Birkh\"auser, 2009.

\bibitem{Barr1979}
M.~Barr,
\emph{$*$-Autonomous Categories},
Lecture Notes in Mathematics, vol.~752, Springer, 1979.

\bibitem{BarrChuHist}
M.~Barr,
The Chu construction: history of an idea,
\emph{Theory Appl.\ Categ.} \textbf{17} (2006), no.~10, 10--16.

\bibitem{BarrOmatt}
M.~Barr,
$*$-Autonomous categories: once more around the track,
\emph{Theory Appl.\ Categ.} \textbf{17} (2006), no.~1, 1--9.

\bibitem{BQ}
Y.~Benoist and J.-F.~Quint,
\emph{Random Walks on Reductive Groups},
Ergebnisse der Mathematik und ihrer Grenzgebiete, vol.~62, Springer, 2016.

\bibitem{Billingsley}
P.~Billingsley,
\emph{Probability and Measure}, 3rd ed.,
Wiley, New York, 1995.

\bibitem{BottcherSilbermann}
A.~B\"ottcher and B.~Silbermann,
\emph{Analysis of Toeplitz Operators}, 2nd ed.,
Springer Monographs in Mathematics, Springer, 2006.

\bibitem{Burg}
J.~P.~Burg,
\emph{Maximum Entropy Spectral Analysis},
PhD thesis, Stanford University, 1975.

\bibitem{BucyJoseph}
R.~S.~Bucy and P.~D.~Joseph,
\emph{Filtering for Stochastic Processes with Applications to Guidance},
Wiley-Interscience, 1968; 2nd ed., Chelsea, 1987.

\bibitem{Cuturi2013}
M.~Cuturi,
Sinkhorn distances: lightspeed computation of optimal transport,
in \emph{Advances in Neural Information Processing Systems~26} (NIPS),
2013, pp.~2292--2300.

\bibitem{CarpentierMansard}
J.~Carpentier and N.~Mansard.
\newblock Analytical derivatives of rigid body dynamics algorithms.
\newblock In \emph{Robotics: Science and Systems~(RSS)}, 2018.

\bibitem{Davenport}
H.~Davenport,
\emph{Multiplicative Number Theory}, 3rd ed.,
Graduate Texts in Mathematics, vol.~74, Springer, 2000.

\bibitem{Diaconis}
P.~Diaconis,
Group representations in probability and statistics,
\emph{IMS Lecture Notes---Monograph Series}, vol.~11, 1988.

\bibitem{Douglas}
R.~G.~Douglas,
\emph{Banach Algebra Techniques in Operator Theory}, 2nd ed.,
Graduate Texts in Mathematics, vol.~179, Springer, 1998.

\bibitem{Durrett}
R.~Durrett,
\emph{Probability: Theory and Examples}, 5th ed.,
Cambridge University Press, 2019.

\bibitem{Featherstone}
R.~Featherstone,
\emph{Rigid Body Dynamics Algorithms},
Springer, 2008.

\bibitem{Folland}
G.~B.~Folland,
\emph{A Course in Abstract Harmonic Analysis}, 2nd ed.,
CRC Press, 2015.

\bibitem{Girard}
J.-Y.~Girard,
Linear logic,
\emph{Theoret.\ Comput.\ Sci.} \textbf{50} (1987), no.~1, 1--101.

\bibitem{Garnett}
J.~B.~Garnett,
\emph{Bounded Analytic Functions}, revised 1st ed.,
Graduate Texts in Mathematics, vol.~236, Springer, 2007.

\bibitem{GrenanderSzego}
U.~Grenander and G.~Szeg\H{o},
\emph{Toeplitz Forms and Their Applications},
University of California Press, 1958.

\bibitem{Hall}
B.~C.~Hall,
\emph{Lie Groups, Lie Algebras, and Representations: An Elementary Introduction}, 2nd ed.,
Graduate Texts in Mathematics, vol.~222, Springer, 2015.

\bibitem{Heyer}
H.~Heyer,
\emph{Probability Measures on Locally Compact Groups},
Ergebnisse der Mathematik und ihrer Grenzgebiete, vol.~94, Springer, 1977.

\bibitem{HewittRoss}
E.~Hewitt and K.~A.~Ross,
\emph{Abstract Harmonic Analysis}, vol.~I,
Grundlehren der mathematischen Wissenschaften, vol.~115, Springer, 1963.

\bibitem{Hoffman}
K.~Hoffman,
\emph{Banach Spaces of Analytic Functions},
Prentice-Hall, 1962; reprinted Dover, 1988.

\bibitem{IwaniecKowalski}
H.~Iwaniec and E.~Kowalski,
\emph{Analytic Number Theory},
AMS Colloquium Publications, vol.~53, 2004.

\bibitem{Kappen}
H.~J.~Kappen,
Path integrals and symmetry breaking for optimal control theory,
\emph{J.\ Stat.\ Mech.}, 2005:P11011, 2005.

\bibitem{Mackey1945}
G.~W.~Mackey,
On infinite-dimensional linear spaces,
\emph{Trans.\ Amer.\ Math.\ Soc.} \textbf{57} (1945), 155--207.

\bibitem{MastalliCrocoddyl}
C.~Mastalli, R.~Budhiraja, W.~Merkt, G.~Saurel, B.~Hammoud,
M.~Naveau, J.~Carpentier, L.~Righetti, S.~Vijayakumar, and N.~Mansard,
Crocoddyl: an efficient and versatile framework for multi-contact
optimal control,
in \emph{IEEE Int.\ Conf.\ Robotics and Automation (ICRA)}, 2020,
pp.~2536--2542.

\bibitem{MarsdenRatiu}
J.~E.~Marsden and T.~S.~Ratiu,
\emph{Introduction to Mechanics and Symmetry}, 2nd ed.,
Texts in Applied Mathematics, vol.~17, Springer, 1999.

\bibitem{Maslov}
V.~P.~Maslov,
\emph{M\'ethodes Op\'eratorielles},
Mir, Moscow, 1987.

\bibitem{MenagerCarpentier2025}
E.~M\'enager, A.~Bambade, W.~Jallet, A.~De~Marchi, and J.~Carpentier.
\newblock Contact-implicit inverse dynamics.
\newblock Submitted to \emph{IEEE Trans.\ Robotics}, 2025.

\bibitem{MochizukiIII}
S.~Mochizuki,
Inter-universal Teichm\"uller theory~III: Canonical splittings of the log-theta-lattice,
\emph{Publ.\ Res.\ Inst.\ Math.\ Sci.} \textbf{57} (2021), 723--1185.

\bibitem{MurrayLiSastry}
R.~M.~Murray, Z.~Li, and S.~S.~Sastry,
\emph{A Mathematical Introduction to Robotic Manipulation},
CRC Press, 1994.

\bibitem{Nikolski}
N.~K.~Nikolski,
\emph{Operators, Functions, and Systems: An Easy Reading}, vols.~1--2,
Mathematical Surveys and Monographs, AMS, 2002.

\bibitem{Pratt1995}
V.~R.~Pratt,
Chu spaces and their interpretation as concurrent objects,
\emph{J.\ Comput.\ System Sci.} \textbf{50} (1995), 334--350.

\bibitem{PrattHPC}
V.~R.~Pratt,
Harmonic predictive control as a variant of {PID} control,
preprint, Stanford University, 2016.

\bibitem{Parthasarathy}
K.~R.~Parthasarathy,
\emph{Probability Measures on Metric Spaces},
Academic Press, 1967; reprinted AMS Chelsea, 2005.

\bibitem{Raibert}
M.~H.~Raibert,
\emph{Legged Robots That Balance},
MIT Press, 1986.

\bibitem{Rudin}
W.~Rudin,
\emph{Fourier Analysis on Groups},
Wiley-Interscience, New York, 1962.

\bibitem{Seely1989}
R.~A.~G.~Seely,
Linear logic, $*$-autonomous categories and cofree coalgebras,
in \emph{Categories in Computer Science and Logic} (Boulder, 1987),
Contemp.\ Math., vol.~92, AMS, 1989, pp.~371--382.

\bibitem{ScholzeStix}
P.~Scholze and J.~Stix,
Why abc is still a conjecture,
preprint, 2018.

\bibitem{Simon}
B.~Simon,
\emph{Orthogonal Polynomials on the Unit Circle}, 2 vols.,
AMS Colloquium Publications, vol.~54, 2005.

\bibitem{SzNF}
B.~Sz.-Nagy, C.~Foias, H.~Bercovici, and L.~K\'erchy,
\emph{Harmonic Analysis of Operators on Hilbert Space}, 2nd ed.,
Universitext, Springer, 2010.

\bibitem{TassaErezTodorov}
Y.~Tassa, T.~Erez, and E.~Todorov,
Synthesis and stabilization of complex behaviors through online
trajectory optimization,
in \emph{IEEE/RSJ Int.\ Conf.\ Intelligent Robots and Systems (IROS)},
2012, pp.~4906--4913.

\bibitem{Todorov}
E.~Todorov,
Linearly-solvable Markov decision problems,
in \emph{Advances in Neural Information Processing Systems~19} (NIPS),
2006, pp.~1369--1376.

\bibitem{TodorovGDD}
E.~Todorov,
General duality between optimal control and estimation,
in \emph{IEEE Conf.\ Decision and Control (CDC)}, 2008, pp.~4286--4292.

\bibitem{TodorovDSO}
E.~Todorov,
Dynamical system optimization,
arXiv:2506.08340, 2025.

\bibitem{Villani}
C.~Villani,
\emph{Optimal Transport: Old and New},
Grundlehren der mathematischen Wissenschaften, vol.~338, Springer, 2009.

\bibitem{Willerton2015}
S.~Willerton,
Tight spans, Isbell completions and semi-tropical modules,
\emph{Theory Appl.\ Categ.} \textbf{28} (2013), no.~22, 696--732.

\bibitem{VDN}
D.~V.~Voiculescu, K.~J.~Dykema, and A.~Nica,
\emph{Free Random Variables},
CRM Monograph Series, vol.~1, AMS, 1992.

\bibitem{HighamExpm}
N.~J.~Higham,
\emph{Functions of Matrices: Theory and Computation},
SIAM, 2008.

\bibitem{WilliamsMPPI}
G.~Williams, A.~Aldrich, E.~A.~Theodorou,
\emph{Model Predictive Path Integral Control: From Theory to Parallel
Computation},
J.~Guidance, Control, and Dynamics \textbf{40}(2), 344--357, 2017.

\bibitem{WuOctonion}
J.~Wu, L.~Xu, Y.~Kong, L.~Senhadji, H.~Shu,
\emph{Deep Octonion Networks},
Neurocomputing \textbf{397}, 179--191, 2020.
arXiv:1903.08478.

\bibitem{BaezOctonions}
J.~C.~Baez,
\emph{The Octonions},
Bull.\ Amer.\ Math.\ Soc.\ \textbf{39}(2), 145--205, 2002.

\bibitem{Varadhan1967}
S.~R.~S.~Varadhan,
\emph{Diffusion processes in a small time interval},
Comm.\ Pure Appl.\ Math.\ \textbf{20}(4), 659--685, 1967.

\bibitem{Nemirovski-Yudin}
A.~Nemirovski and D.~Yudin,
\emph{Problem Complexity and Method Efficiency in Optimization},
Wiley, 1983.

\bibitem{Shalev-Shwartz}
S.~Shalev-Shwartz,
Online learning and online convex optimization,
\emph{Foundations and Trends in Machine Learning} \textbf{4}(2), 107--194, 2012.

\bibitem{BakryEmery}
D.~Bakry and M.~\'Emery,
Diffusions hypercontractives,
in \emph{S\'eminaire de Probabilit\'es~XIX},
Lecture Notes in Math., vol.~1123, Springer, 1985, pp.~177--206.

\bibitem{SrinivasGPUCB}
N.~Srinivas, A.~Krause, S.~M.~Kakade, and M.~W.~Seeger,
Gaussian process optimization in the bandit setting: no regret and
experimental design,
in \emph{Proc.\ 27th Int.\ Conf.\ Machine Learning (ICML)}, 2010,
pp.~1015--1022.

\bibitem{SuiSafeOpt}
Y.~Sui, A.~Gotovos, J.~W.~Burdick, and A.~Krause,
Safe exploration for optimization with Gaussian processes,
in \emph{Proc.\ 32nd Int.\ Conf.\ Machine Learning (ICML)}, 2015,
pp.~997--1005.

\bibitem{CalandrielloMINIMETA}
D.~Calandriello, L.~Carratino, A.~Lazaric, M.~Valko, and L.~Rosasco,
Near-linear time Gaussian process optimization with adaptive batching
and resparsification,
in \emph{Proc.\ 39th Int.\ Conf.\ Machine Learning (ICML)}, 2022,
pp.~2451--2467.

\bibitem{LiuSVGD}
Q.~Liu and D.~Wang,
Stein variational gradient descent: a general purpose Bayesian inference
algorithm,
in \emph{Advances in Neural Information Processing Systems~29} (NeurIPS),
2016, pp.~2378--2386.

\bibitem{Ledoux}
M.~Ledoux,
\emph{The Concentration of Measure Phenomenon},
Mathematical Surveys and Monographs, vol.~89, AMS, 2001.

\bibitem{ManchesterUDP}
Z.~Manchester and S.~Kuindersma,
Derivative-free trajectory optimization with unscented dynamic
programming,
in \emph{IEEE Conf.\ Decision and Control (CDC)}, 2016,
pp.~3642--3647.

\bibitem{PlancherCUDP}
B.~Plancher, Z.~Manchester, and S.~Kuindersma,
Constrained unscented dynamic programming,
in \emph{IEEE/RSJ Int.\ Conf.\ Intelligent Robots and Systems (IROS)},
2017, pp.~5674--5680.

\bibitem{BerkenkampSafeMPC}
F.~Berkenkamp, M.~Turchetta, A.~P.~Schoellig, and A.~Krause,
Safe model-based reinforcement learning with stability guarantees,
in \emph{Advances in Neural Information Processing Systems~30} (NeurIPS),
2017, pp.~908--918.

\bibitem{FreyDiffMPC}
J.~Frey, K.~Baumg\"artner, G.~Frison, D.~Reinhardt,
J.~Hoffmann, L.~Fichtner, S.~Gros, and M.~Diehl,
Differentiable nonlinear model predictive control,
\emph{arXiv preprint arXiv:2505.01353}, 2025.

\end{thebibliography}
