\chapter{Applications}\label{sec:applications}

\section{The Qin operating system}\label{sec:qin}

The Qin state (356--207~BCE) provides the first complete engineering
implementation of the framework. Shang Yang's reforms map directly to
the formal vocabulary:

\begin{center}
\begin{tabular}{@{}lp{5cm}p{5cm}@{}}
\toprule
\textbf{Policy} & \textbf{Framework equivalent} & \textbf{Effect} \\
\midrule
Abolish well-field system & Remove aristocratic $\Ur$ & Nobles
$\to$ commoners \\
Military merit ranks & $\Ur$ controlled by state (revocable) &
Actuation is rented, not owned \\
Mutual surveillance (\emph{lianzuo}) & Maximise $\Obs$ &
Neighbors $=$ distributed sensor network \\
Standardise weights \& measures & Increase $\Obs$ precision &
Higher observational resolution \\
Commandery-county system & All chains through capital &
King $=$ cut vertex \\
\bottomrule
\end{tabular}
\end{center}

Qin's unification of the six states was a graph-theoretic outcome:
Qin's star graph (center $=$ Xianyang, $O(1)$ dispatch) vs.\ the six
states' mesh graphs (multiple aristocratic centers, $O(n^2)$
coordination cost).

\begin{remark}[Temporal linearity]\label{rem:temporal}
The mesh-to-star compression is not an efficiency optimisation.  It is
a survival requirement forced by temporal linearity.

Time is real and linear: no pause, no speedup, no frame drops
(\cref{rem:deployment,rem:di-temporal}).  Every system that maintains
viability must complete a full sense--compute--act cycle within one
time step $\Delta t$:
\[
  \underbrace{|\phi| > 0}_{\text{do not fall}}
  \quad\wedge\quad
  \underbrace{t_{\pi} \leq \Delta t}_{\text{do not lag}}.
\]
The first condition is physical (the viability axiom).  The second is
computational: the response function must fit inside $\Delta t$.
Both must hold simultaneously.  Violating either kills the system
by different mechanisms: the first is a fall, the second is a lag,
and the outcome is the same.

This binds computational complexity to survival.  A mesh topology
dispatches in $O(n^2)$; a star topology in $O(1)$.  As the system
grows (more territory, more agents), $\Delta t$ does not grow with
it---time does not slow down for larger empires.  At critical $n$,
mesh dispatch exceeds $\Delta t$ and the system can no longer respond
in time.  商鞅's reform---mesh to star---is latency engineering:
compressing $t_\pi$ to fit within a fixed $\Delta t$ at any scale.

The isomorphism between robotic deployment and historical deployment
is exact:
\begin{center}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Deployment (robotics)} & \textbf{Deployment (history)} \\
\midrule
policy $\pi$ (neural network) & emperor (cut vertex) \\
sensor reading $o_t$ & intelligence reports \\
torque command $\tau_t$ & edicts, military orders \\
$\Delta t = 20$\,ms & dynastic response window \\
ReLU (piecewise linear, $O(n)$) & star graph ($O(1)$ dispatch) \\
$|\phi| = 0$ (robot falls) & dynasty collapses \\
$t_\pi > \Delta t$ (compute lags) & cut vertex non-functional \\
\bottomrule
\end{tabular}
\end{center}

Every structural choice in the framework---ReLU as activation
function, star as topology, binary threshold as detection
criterion---is forced by the requirement that computation fit inside
real, linear, non-pausable time.  These are not design preferences;
they are the only structures fast enough to survive.

The two historical failure modes---the processor that runs an
irrelevant subroutine (latency explosion) and the processor that is
removed entirely (see \cref{sec:errorlog} for the full
analysis)---are the two ways the computational constraint breaks:
\begin{itemize}
\item 嘉靖: cut vertex present but $t_\pi \to \infty$.
  The emperor computes Daoist alchemy for twenty years while the
  system requires real-time response.  Every year without a routing
  decision is a dropped frame.  Gravity does not pause while the
  processor runs an irrelevant subroutine.
\item 元顺帝: cut vertex removed.  No processor.  Immediate crash
  ($|\phi| = 0$).
\end{itemize}
Temporal linearity is not a property of one remark in one chapter.
It is the medium in which the entire theory operates.
\end{remark}

Shang Yang's fate validates the framework's robustness: his reform
network itself became an autonomous actuator ($\Ur \neq \varnothing$).
The system eliminated its creator---not irony, but a robustness test.
A sword-detection system that exempts its designer is not robust
(\cref{sec:second} formalises this as second-mover viability).

\subsection{The submartingale-induced unitary group}

More precisely, Shang Yang's reform sequence is a
\emph{submartingale-induced unitary group}.

\begin{definition}[Submartingale reform]\label{def:submartingale}
A reform sequence $X_0, X_1, \ldots, X_n$ is a \emph{submartingale}
if each step strictly increases the centralisation index:
$\mathbb{E}[X_{k+1}] \geq X_k$ (deterministic analogue: the
sequence is non-decreasing in expectation under any re-ordering,
with the added constraint of irreversibility), and each step is
irreversible (reversal requires undoing all subsequent steps).
\end{definition}

Shang Yang's five reforms form such a sequence:
\[
  \underbrace{\text{abolish well-fields}}_{X_0}
  \to \underbrace{\text{merit ranks}}_{X_1}
  \to \underbrace{\text{mutual surveillance}}_{X_2}
  \to \underbrace{\text{standard measures}}_{X_3}
  \to \underbrace{\text{commandery-county}}_{X_4}.
\]
Each step burns the entropy of the old feudal order irreversibly.

Once installed, the system forms a \emph{unitary group}: a
structure-preserving transformation that maintains its invariants
(king $=$ cut vertex, all swords eliminated) at every cycle, applied
to every vector in the state space without exception.

\begin{theorem}[Shang Yang's paradox]\label{thm:shangyang}
Any agent who installs a sword-elimination system via submartingale
reform must possess $\Ur \neq \varnothing$ to execute the installation.
But the submartingale is monotone: each step lowers the detection
threshold. The installer's own actuation becomes increasingly visible
with each reform step. Upon completion, the unitary group acts on the
installer:
\[
  \underbrace{\text{installer drives submartingale}}_{\text{requires }
  \Ur \neq \varnothing}
  \;\longrightarrow\;
  \underbrace{\text{unitary group forms}}_{\text{detects all }
  \Ur \neq \varnothing}
  \;\longrightarrow\;
  \underbrace{\text{group acts on installer}}_{\text{elimination}}.
\]
The system has no creator exemption.
\end{theorem}

\begin{proof}
The installer is not in the invariant subspace of the group he
created (because $\Ur \neq \varnothing$). A unitary group does not
preserve non-invariant elements---it decomposes them. But unitary
means \emph{norm-preserving}: the installer is destroyed, but his
contributions are fully conserved. The reforms---commandery-county
system, merit ranks, mutual surveillance, standard measures---persist
uniformly across the empire. The person is eliminated; not one bit of
the contribution is lost. This is not destruction but
\emph{delocalisation}: a localised power entity is scattered across
the full state space by the unitary action. Norm conserved, structure
zeroed.
\end{proof}

\subsection{The existence proof}

Qin unified the six states, then collapsed (207~BCE, lasting only
15~years). But Qin left something more powerful than any army: an
\emph{existence proof}.

\begin{theorem}[Qin existence theorem]\label{thm:qin}
\[
  \exists\;\text{centralised state}\;\text{s.t.}\;
  \text{autonomous actuation} = 0
  \;\wedge\;
  \text{state functions.}
\]
\end{theorem}

Before Qin, no one knew a state without feudal aristocrats was
possible. Qin proved it was---and proved it functioned \emph{better}
(star-graph dispatch vs.\ mesh-graph coordination). You cannot refute
an existence proof. You can burn the paper, but the theorem persists.

After Qin fell, two men entered the ruins. Liu Bang, then a minor
official, had once seen the First Emperor's procession and sighed:
``A great man should be like this!''---\emph{I want to BE this system's
cut vertex}. Xiang Yu, upon conquering Xianyang, burned the palaces
and said: ``Who would be emperor in the dark where no one can
see?''---he wanted the \emph{trophy}, not the \emph{theorem}. One read
the existence proof. The other burned it. The one who read it founded
a dynasty that lasted four centuries.

\subsection{The wisdom of being second}\label{sec:second}

\Cref{thm:shangyang} has a direct corollary that inverts the usual
narrative of primacy.

\begin{corollary}[Second-mover viability]\label{cor:second}
Let $X_0, \ldots, X_n$ be a submartingale reform
(\cref{def:submartingale}) installed by agent~$\alpha$,
and let $\beta$ be an agent who inherits the resulting
infrastructure without participating in the installation.
Then $\beta$ has strictly higher viability than $\alpha$:
\[
  \Ur(\alpha) \neq \varnothing
  \;\;\text{(installation requires actuation)},\qquad
  \Ur(\beta) = \varnothing
  \;\;\text{(inheritance does not)}.
\]
The installed unitary group eliminates $\alpha$ and preserves $\beta$.
\end{corollary}

\begin{proof}
By \cref{thm:shangyang}, the installer $\alpha$ necessarily possesses
$\Ur \neq \varnothing$ and is detected upon completion.  The
inheritor~$\beta$ uses the infrastructure but did not create it:
$\Ur(\beta) = \varnothing$ with respect to the installation.  The
detection function tests $\Ur$, not provenance.  $\beta$ passes;
$\alpha$ does not.
\end{proof}

The existence proof (\cref{thm:qin}) has an installation cost, paid
exactly once by the first mover.  The cost is non-transferable and
non-refundable.  After installation the landscape is permanently
shifted; second movers operate in the post-shift regime without
needing to shift anything.

The first mover's contribution is \emph{entangled} with $\Ur$---they
needed $\Ur$ to install.  The second mover's usage is
\emph{factorised}---they use the infrastructure without the
installer's $\Ur$.  The second mover's path to recognition is smooth;
the first mover's is not.

\begin{center}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{@{}llp{3.5cm}p{4.5cm}@{}}
\toprule
\textbf{First} & \textbf{Second} & \textbf{Inherited} &
\textbf{Outcome} \\
\midrule
商鞅 & 秦's successors & commandery-county, merit ranks &
  商鞅 torn apart; Qin unifies \\
秦 & 汉 (刘邦) & existence proof & Qin 15\,yr; Han 400\,yr \\
项羽 & 刘邦 & Xianyang (capital) & 项羽 suicide; 刘邦 founds Han \\
王安石 & 蔡京 & fiscal infrastructure & 王安石 exiled; 蔡京 20\,yr
  chancellor \\
\bottomrule
\end{tabular}
\end{center}

范蠡 (5th century BCE) is the purest instance.  He helped 勾践 destroy
吴 but let the king remain the visible cut vertex.  Upon victory he
vanished---accumulated three fortunes and dispersed two, ensuring
$\Ur \approx \varnothing$ at every stage.  He was always second.
张良 followed the same strategy: after 刘邦's victory he retired to
``follow 赤松子,'' reducing $\Ur$ to $\varnothing$ before the
detection threshold reached him.  Both understood \cref{cor:second}
before it was stated: read the existence proof, do not write it; use
the infrastructure, do not install it.

\section{The dollar as sword precursor}\label{sec:dollar}

Money is not a sword (it cannot actuate independently). Money is a
\emph{sword precursor}: a universal voucher exchangeable for
$\Ur \neq \varnothing$ on the market. The king detects not how much
money you have, but whether your money has \emph{already converted}
into autonomous execution capability.

Fan Li (5th century BCE) understood this: he accumulated fortunes
three times and dispersed them twice, ensuring $\Ur$ never crossed
the threshold. Shen Wansan and Hu Xueyan did not---their wealth
formed closed execution loops, and they were eliminated.

\section{The Atlantic triangular trade: a sword with no cut
vertex}\label{sec:triangle}

The Atlantic Triangular Trade (16th--19th century) is a three-leg
closed execution loop: manufactured goods (Europe $\to$ West Africa),
enslaved human beings (West Africa $\to$ Americas), raw materials
(Americas $\to$ Europe). This loop satisfies \cref{def:sword}: it
actuates autonomously (multiple nations operate independent instances;
remove any one and the loop continues) and is observable (300~years of
ships, ledgers, and auction blocks).

The loop has a property stronger than condition~(1): it is
\emph{self-financing}. The raw materials extracted on leg~3 fund the
goods shipped on leg~1, which purchase the enslaved labour on leg~2,
who produce the raw materials on leg~3. A sword that funds its own
actuation cannot be starved from outside.

\paragraph{Where the isomorphism breaks.}
In \cref{sec:water}, the king--pawn relationship is sustained by water
(viability bargain: $\text{water} > 0$). In the Triangular Trade,
enslaved people had $\text{water} = 0$ from day one. Force~$F$
substituted for water---temporarily. The substitution cost: every unit
of force burns resources drawn from the extracted water on leg~3. The
loop becomes a self-financing violence machine. \Cref{sec:water}
predicts $\text{water} = 0 \implies \text{pawn} \to \text{sword}
\implies \text{collapse}$; force delays but does not prevent this.
The longer the delay, the more violent the eventual resolution.

\paragraph{Abolition as relabelling.}
Applying \cref{prop:phase}: did abolition (1807--1888) change the
\emph{physics} or only the \emph{label}? The labour force moved from
slavery to sharecropping on the same plantations, the resource flow
(raw materials $\to$ Europe) persisted, the force mechanism shifted
from slave codes to Jim Crow and convict leasing, and the laborer
still had no breakpoint. Verdict: phase transition in the sense of
\cref{sec:phase}---label changed, topology unchanged. This is
path~(c), which \cref{thm:fixedpoint} proves is not a resolution.

\begin{theorem}[Closed-loop sword]\label{thm:triangle}
Any closed execution loop that is self-financing, has no cut vertex
(mesh topology, multiple independent operators), and substitutes
force for water, has exactly two exit paths: physical dismantlement
of the loop topology [path~(a)], or system-level collapse when
force-substitution fails [path~(b)]. Relabeling [path~(c)] preserves
$\Ur \neq \varnothing$ and resolves nothing.
\end{theorem}

\paragraph{Missing cut vertex.}
The essay's framework (\cref{sec:cutvertex}) assumes a single cut
vertex. The Triangular Trade has none: its topology is a mesh, not a
star. Removing any one colonial power does not disconnect the loop.
This makes path~(a) structurally harder---there is no single point to
dismantle---but does not create a third option.

\begin{remark}[Path~(a) by mean-field shift]\label{rem:chaplin}
Chaplin's 1925 song \emph{With You, Dear, in Bombay}---composed during
\emph{The Gold Rush}, a film about extractive economics---takes the same
maritime route and replaces the extractive payload with a romantic one.
The singer sails to Bombay not to trade but to reunite. This is not
path~(c) (the route is not relabeled; the self-financing loop is broken
because love does not fund the next leg). It is path~(a) by mean-field
shift: the infrastructure persists, but the mean changes, and therefore
the sword changes.
\end{remark}

\section{The corrupted detection function}\label{sec:nazi}

\Cref{def:sword} uses a structural detection function
$\Obs_{\mathrm{structural}}$ that tests $\Ur$: \emph{what can the
resource do independently?} This function is identity-blind. Zhang
Liang is safe because $\Ur = \varnothing$, not because he is Zhang
Liang.

\begin{definition}[Detection function corruption]\label{def:corruption}
A detection function is \emph{corrupted} when $\Obs_{\mathrm{structural}}$
(tests $\Ur$) is replaced by $\Obs_{\mathrm{identity}}$ (tests group
membership):
\[
  \Obs_{\mathrm{identity}}:\;
  r \;\mapsto\;
  \begin{cases}
    \textbf{sword} & \text{if } \mathrm{agent}(r) \in \text{Group } X, \\
    \textbf{not sword} & \text{otherwise.}
  \end{cases}
\]
\end{definition}

The corruption produces a \emph{false positive catastrophe}: every
member of Group~$X$ with $\Ur = \varnothing$ is classified as a sword
and eliminated. Every agent outside Group~$X$ with $\Ur \neq \varnothing$
is a false negative. Six million false positives is the output of a
corrupted detection function running to completion.

\begin{definition}[Nazi structure]\label{def:nazi}
A system exhibits \emph{Nazi structure} if and only if: (1)~it performs
viability maintenance (eliminates perceived threats to $\Viab(K)$),
(2)~its detection criterion is identity-based, not structure-based,
and (3)~it executes path~(b) against agents classified by identity
who do not satisfy \cref{def:sword} ($\Ur = \varnothing$ for the
individual). These conditions are necessary and sufficient.
\end{definition}

\begin{corollary}[Identity-invariance]\label{cor:nazi}
The Nazi structure is identity-invariant on both sides: it depends
neither on the identity of the perpetrator nor on the identity of the
target. Replacing Group~$X$ with any other identity group preserves
all three conditions.
\end{corollary}

The operator that maps $\Obs_{\mathrm{structural}} \mapsto
\Obs_{\mathrm{identity}}$ is \emph{Hate}. In this framework, Hate is
not an emotion---it is an operator on detection functions with a
precise signature (structural $\to$ identity) and a precise output
(mass false positives). The motivation is irrelevant; the mapping
determines the output.

\section{The parasitic cut vertex}\label{sec:parasite}

\Cref{thm:cutvertex} establishes the cut vertex as the optimal
survival strategy: route all execution chains through yourself.
A \emph{parasitic cut vertex} has the same topology but opposite flow
direction: it routes access and information for extraction rather than
coordination.

\begin{definition}[Parasitic cut vertex]\label{def:parasite}
A parasitic cut vertex $v_p$ satisfies:
(1)~$G \setminus \{v_p\}$ is disconnected (cut vertex),
(2)~$v_p$ holds no formal authority ($\Ur \approx \varnothing$
officially),
(3)~$v_p$ extracts resources from both sides of the partition by
routing access across the cut.
\end{definition}

The parasitic cut vertex turns \emph{others} into pawns via the
routing function (e.g., blackmail: ``I route your secret; comply or I
re-route it publicly''). This is \cref{thm:cutvertex} inverted:
instead of building the system, the parasite extracts from it.

\begin{proposition}[Self-termination]\label{prop:parasite}
A parasitic cut vertex is self-terminating. When the secrecy that
maintains the cut ($\text{water} = \text{secrecy}$) fails, the graph
reconnects, the cut vertex property vanishes, and the former parasite
faces the combined action of all previously partitioned nodes.
\end{proposition}

Unlike the productive cut vertex (which may persist for centuries via
institutional embedding), the parasitic variant stores no structural
contribution. Upon delocalisation, norm is conserved but there is
nothing to conserve: the extraction leaves no invariant.

\section{Militarism and the net-positive ask}\label{sec:militarism}

Militarism is the structural type where the pawn ($\Ur = \varnothing$,
the military apparatus) captures the cut vertex position, producing
$\Ur \neq \varnothing$ for the military and $\Ur \to \varnothing$ for
civilian institutions. In the framework's language: the pawn becomes the king.

\begin{example}[The Kniefall: path~(a) in 30 seconds]\label{ex:kniefall}
On December~7, 1970, West German Chancellor Willy Brandt knelt before
the Warsaw Ghetto Uprising memorial. This act satisfies four conditions
that make it a pure instance of path~(a):
\emph{physical} (a bodily act, not a speech act---cannot be retracted
by reinterpretation),
\emph{unconditional} (no negotiation, no demand for reciprocity),
\emph{performed by the cut vertex} (the head of state, the node
through which all institutional chains pass), and
\emph{irretractable} (a photograph is a permanent record).
Brandt's Kniefall is an existence proof that path~(a) is available to
any system.
\end{example}

After the Kniefall, Germany installed institutional breakpoints:
Article~1 of the \emph{Grundgesetz} (``Human dignity is inviolable''),
the Federal Constitutional Court as an independent $\Obs$, EU and NATO
membership as external breakpoints. These institutionalise the phase
transition: the system moved from $U = \Umax$ (Nazi regime) through
path~(a) (Kniefall) to a constrained system with structural breakpoints.

\begin{proposition}[Net-positive theorem]\label{prop:netpositive}
Path~(a) is strictly net-positive for all parties. Comparative evidence:
Germany (path~(a), Kniefall 1970) vs.\ Japan (path~(c), relabelling
without structural change, 80~years and counting). Every measurable
outcome---diplomatic relations, regional stability, economic
integration, soft power, domestic constitutional health---favors the
path~(a) system. The cost of path~(a) is pride. The return is
everything else.
\end{proposition}
