\chapter{Discussion}\label{sec:discussion}

\section{Domain of applicability}\label{sec:domain}

The framework is a compressed model that degrades under the following
conditions:

\begin{center}
\begin{tabular}{@{}lp{5.5cm}p{5cm}@{}}
\toprule
\textbf{Condition} & \textbf{Failure mode} & \textbf{Example} \\
\midrule
Diffuse sovereignty & No unique king; cut vertex undefined & Late
European feudalism, early federalism \\
External shock dominance & $\Viab(K)$ broken by external force;
internal knife dynamics secondary & Mongol invasion, colonialism \\
Rapid ideological reshaping & $K$ itself is changing faster than
actuation reshapes state & Religious revolution, ideology \\
High-latency detection & $\Obs$ too slow; phase transition completes
before detection & Large empire frontiers \\
Universal $\Ur \approx \varnothing$ & No knives; framework trivial &
Extremely atomised societies \\
\bottomrule
\end{tabular}
\end{center}

The framework applies to systems with a unique sovereign, differentiated
agent capabilities, and sufficient observability. This covers the main
interval of Chinese imperial history but not all political forms.

\section{The water dynamics}\label{sec:water}

The framework so far analyzes one layer: the king--knife interaction. A
complete viability analysis requires the \emph{viability chain}---the
three-level dependency that sustains the system:
\[
  \text{King}
  \xrightarrow{\;\text{needs}\;}
  \text{Pawn}
  \xrightarrow{\;\text{needs}\;}
  \text{Water.}
\]
The \emph{pawn} is any agent with $\Ur \neq \varnothing$ whose actuation
is currently directed by the king (soldiers, administrators, tax
collectors). \emph{Water} is the population's aggregate resource
level---the substrate from which the pawn draws manpower, revenue, and
legitimacy.

\begin{definition}[Water]\label{def:water}
Water $w(t) \in [0, W_{\max}]$ is the population's aggregate extractable
resource level at time $t$. The pawn's actuation is bounded by water:
$\|\Ur\| \leq g(w)$ for some monotone function $g$ with $g(0) = 0$.
\end{definition}

The king needs the pawn to execute viability maintenance (eliminate
knives, administer territory). The pawn needs water to function. If
$w \to 0$, the pawn's actuation capacity collapses regardless of the
king's commands.

\begin{proposition}[Binary action space at $w = 0$]\label{prop:binary}
When $w(t) \to 0$, the pawn's action space collapses to a binary:
\[
  A_{\text{pawn}} = \{\text{submit},\; \text{rebel}\}.
\]
The intermediate options (negotiate, migrate, trade, accumulate) require
$w > 0$. At $w = 0$, the pawn has nothing to lose, and the viability
axiom (\cref{ax:viability}) now applies \emph{to the pawn}: the pawn's
own survival requires a viable path, and submission no longer provides
one. The pawn becomes a knife---$\Ur$ transitions from $\varnothing$ to
$\neq \varnothing$---and the king faces a knife he created by exhausting
the water.
\end{proposition}

\begin{theorem}[Du Mu's theorem]\label{thm:dumu}
Let $w(t)$ be decreasing under extraction. Then:
\[
  w(t) \to 0
  \implies
  \text{pawn} \to \text{knife}
  \implies
  \text{king absorbed.}
\]
The causal chain is internal: the system destroys itself by exhausting
its own substrate.
\end{theorem}

This is the content of Du Mu's \emph{A Fang Gong Fu} \cite{dumu} (825~CE): ``It was
not the Qin who destroyed the six states, but the six states themselves;
it was not the world that destroyed Qin, but Qin itself.'' In our
language: the states depleted their own water, creating the knives that
destroyed them. Qin, having unified, then depleted its own water
(conscription for the Great Wall and Epang Palace), creating the knives
(Chen Sheng, Wu Guang) that destroyed it.

\begin{remark}[Water as viability constraint]
``Water can carry the boat, and water can capsize the boat'' (attributed
to Wei Zheng, Tang dynasty) is not a metaphor. It is a restatement of
\cref{thm:dumu}: the substrate that enables the king's viability
($w > 0$ $\implies$ pawn functions $\implies$ king's path to $\infty$
exists) is the same substrate whose depletion destroys it ($w = 0$
$\implies$ pawn $\to$ knife $\implies$ no path to $\infty$).
\end{remark}
