\chapter{The Sword as Mean Field}\label{sec:meanfield}

The preceding sections defined the sword as a two-condition criterion
applied to individual resources. We now argue that the sword is
fundamentally a \emph{mean-field} phenomenon.

\section{The mean actuation field}

Consider $n$ agents with autonomous actuation levels
$\|U_1\|, \ldots, \|U_n\|$, where $\|U_i\|$ measures the
\emph{degrees of freedom} of agent~$i$'s autonomous control set
(the dimension of $\mathrm{span}(U_i)$ when $U_i$ is a linear
subspace; $|U_i|$ when discrete).
Define the \emph{mean actuation field}:
\[
  \bar{U} = \frac{1}{n} \sum_{i=1}^{n} \|U_i\|.
\]

The king's detection function $\Obs$ has finite bandwidth: it cannot
monitor all agents with equal precision.

\begin{proposition}[Detection as optimal monitoring]
\label{prop:detection-derivation}
Let the king have finite monitoring bandwidth~$B$: he can
observe at most~$B$ agents simultaneously.  Under the viability
axiom (\cref{ax:viability}), the optimal monitoring policy
$\Obs^*$ is a threshold rule:
\[
  r \in \mathrm{Im}(\Obs^*)
  \iff
  \|U_r\| - \bar{U} > \tau(B),
\]
where $\tau(B)$ is the threshold that selects the top-$B$
agents by deviation from the mean.
\end{proposition}

\begin{proof}[Proof sketch]
\leavevmode
\begin{enumerate}[label=(\roman*),nosep]
  \item The king must maintain $F(x) \cap T_K(x) \neq
  \varnothing$ (tangential condition,
  \cref{thm:viability-di}).
  \item The threat from agent~$r$ is proportional to
  $\|\Ur\|$ (the dimension of its autonomous control set).
  \item The mean field~$\bar{U}$ is the unique unbiased
  baseline: by the massless axiom, deviations must net to
  zero, forcing $b = \bar{U}$ (\cref{lem:capacity}).
  Agents at or below the mean carry zero bypass capacity
  and pose no threat.
  \item With finite bandwidth~$B$, optimal allocation monitors
  the agents with largest $\|\Ur\| - \bar{U}$, i.e., the
  most deviant.
  \item The threshold $\tau(B)$ is determined by~$B$: it is
  the $(n - B)$-th order statistic of
  $\{\|U_i\| - \bar{U}\}$.
\end{enumerate}
This is the evaluation operator $T^{\Obs}$: given the current
mean field~$\bar{U}$, it evaluates which agents are monitored.
The detection equation~\eqref{eq:detection} is the fixed point
of this operator, self-consistent when both~$\bar{U}$ and the
detected set are stable.
\end{proof}

By \cref{prop:detection-derivation}, $\Obs$ triggers on agents
whose actuation deviates significantly from the mean:
\begin{equation}\label{eq:detection}
  r \in \mathrm{Im}(\Obs)
  \iff
  \|U_r\| - \bar{U} > \tau(\Obs),
\end{equation}
where $\tau(\Obs)$ is the detection threshold determined by the king's
observational capacity.  This reduces the set-valued observability
concept ($r \in \mathrm{Im}(\Obs)$ or not) to a scalar comparison,
valid when the king's bandwidth is finite and detection is
threshold-based.

\begin{theorem}[The sword is the mean]\label{thm:meanfield}
The sword threshold is determined by the system's mean actuation field.
A resource $r$ is a sword if and only if:
\begin{enumerate}[label=(\roman*)]
  \item $\Ur \neq \varnothing$ (autonomous actuation exists), and
  \item $\|U_r\|$ exceeds the mean field by more than the detection
  threshold: $\|U_r\| > \bar{U} + \tau(\Obs)$.
\end{enumerate}
Consequently, the phase transition (\cref{prop:phase}) is a shift in
$\bar{U}$, not a change in any individual $\Ur$.
\end{theorem}

\begin{proof}
Condition~(i) is unchanged from \cref{def:sword}.
For condition~(ii): by the detection model~\eqref{eq:detection},
$r \in \mathrm{Im}(\Obs)$ if and only if
$\|U_r\| - \bar{U} > \tau(\Obs)$.  Substituting into the sword
criterion (\cref{def:sword}) yields the stated biconditional.

For the phase-transition claim: let $\bar{U}_1$ denote the mean
before the transition and $\bar{U}_2 < \bar{U}_1$ the mean after.
An agent with fixed $\|U_r\|$ satisfying
$\bar{U}_2 + \tau < \|U_r\| \leq \bar{U}_1 + \tau$
is not a sword in regime~1 but becomes a sword in regime~2.
The shift is in $\bar{U}$, not in any individual $\|U_r\|$.
\end{proof}

\begin{remark}[Historical illustration]\label{rem:meanfield-history}
The formal mechanism is visible at every phase transition.
In wartime, many agents carry high actuation; the mean $\bar{U}$ is
high; few exceed the threshold. At demobilization, the mean drops
sharply. Agents who retain wartime-level actuation now exceed the
peacetime threshold---the same $\|U_r\|$ that was unremarkable before
is now a deviation. The sword is not created by the agent but by the
shift in the system's reference frame.
\end{remark}

\begin{remark}[Connection to statistical mechanics]\label{rem:statmech}
This is precisely the mechanism of a phase transition in statistical
mechanics: the order parameter (mean actuation) shifts, and
configurations that were typical in one phase become atypical---and
therefore detectable---in the other. The viability axiom plays the role
of the free energy: the system minimises threats to $\Viab(K)$, just
as a thermodynamic system minimises free energy.
\end{remark}

\begin{remark}[Mechanism consistency: mother--daughter distributions]
\label{rem:mother-daughter}
The fixed-point structure of $T^{\Obs}$ admits a distributional
interpretation.  Let $q_j$ denote the current belief about
agent~$j$'s state (the \emph{mother} distribution), and let
$q_j^{\mathrm{da}}$ denote the pushforward of $q_j$ through
the mechanism---the state distribution that \emph{would} result
if agent~$j$ executed its autonomous actuation for one step
(the \emph{daughter} distribution).  The detection operator
$T^{\Obs}$ is self-consistent when
$\mathrm{KL}(q_j \| q_j^{\mathrm{da}})$ is small for all
monitored agents: the king's belief about what each agent
\emph{will} do matches what it \emph{would} do.

A phase transition (\cref{prop:phase}) breaks this consistency:
$\bar{U}$ shifts, the daughter distribution under the new mean
field diverges from the mother, and the detection threshold
must re-equilibrate.  The feedback loop of
\cref{thm:paradox}---elimination $\to$ lower $\bar{U}$ $\to$
new swords---is precisely the transient before the
mother--daughter distributions re-align at the new fixed point.
\end{remark}

\begin{remark}[From parent--child to mother--daughter]
\label{rem:parent-daughter-bridge}
In any DAG the \emph{parent set}
$\operatorname{pa}(v) = \{u : (u, v) \in E\}$ encodes
which nodes may influence~$v$.  In a dynamical system
supported on the DAG, each node updates via a local
\emph{causal scan kernel}
\[
  X_v(t{+}1) \;\sim\;
  Q_v\!\bigl(\,\cdot \mid X_{\operatorname{pa}(v)}(t{+}1),\;
  X_v(t)\bigr),
\]
where parents are scanned in topological order so that
$X_{\operatorname{pa}(v)}(t{+}1)$ is already available when
$v$ updates.  This is the micro-level dynamics.

When the DAG has a layered structure and nodes within each
layer are exchangeable, the dependence on individual parent
states collapses to a dependence on the parent-layer
\emph{external field} $\eta = \varphi(\mu^{(\ell-1)})$
(e.g.\ the mean $\bar{U}$ of \cref{prop:phase}).  The kernel
becomes the \emph{mother--daughter kernel}
\[
  Q_\ell\!\bigl(\,\cdot \mid x,\;\eta\bigr),
\]
and the layer-wise measure update
$\mu^{(\ell)}_{t+1} = Q_\ell \sharp\, \mu^{(\ell)}_t$
is the mean-field flow.  In short: parent--child is the
graph-theoretic relation; mother--daughter is its
distributional lift under exchangeability.
\end{remark}

\begin{remark}[Cut-cycle and the causal mass gap]
\label{rem:causal-mass-gap}
Feedback loops---such as the elimination--resurgence cycle of
\cref{thm:paradox}---appear as cycles in the dependency graph.
The \emph{cut-cycle principle} resolves this: delay one edge
by one time step, so that $v$ at time $t{+}1$ depends on $u$
at time~$t$ rather than $t{+}1$.  The resulting \emph{unrolled
DAG} (one copy of~$V$ per time step, edges pointing strictly
forward in time) is acyclic, and the causal scan of
\cref{rem:parent-daughter-bridge} applies.

On this unrolled DAG, the spectral gap
$\lambda_1 > 0$ of \cref{thm:massgap} acquires a causal
interpretation: it is the rate at which correlations decay
along the causal direction.  Precisely, for local observables
$F_a, G_b$ separated by causal distance~$d$,
\[
  |\operatorname{Cov}(F_a, G_b)|
  \;\leq\;
  C\,e^{-\lambda_1\, d}.
\]
The mass gap $\lambda_1 > 0$ means finite causal correlation
length; its closure ($\lambda_1 \to 0$) is the phase
transition of \cref{prop:phase}---critical slowing down,
divergent correlation length, and the onset of long-range
order.
\end{remark}

\begin{remark}[Validity conditions]\label{rem:meanfield-validity}
The arithmetic mean~$\bar{U}$ is the operative threshold for
homogeneous systems where agents draw from comparable control
sets.  In hierarchical systems, clique amplification
(see \cref{app:zhongxian}) may require a modified threshold
that accounts for correlated actuation within subgroups.
\end{remark}

\section{Implications}

The mean-field interpretation resolves several puzzles:

\begin{enumerate}
  \item \textbf{Why identical resources have different fates.} Two
  generals with identical $\Ur$ can have opposite outcomes if one
  operates in a high-$\bar{U}$ environment (wartime coalition) and
  the other in a low-$\bar{U}$ environment (consolidated empire).
  The sword is relative to the mean.

  \item \textbf{Why the paradox is a feedback loop.} As the king
  eliminates swords, $\bar{U}$ drops, lowering the threshold. Agents
  who were below the old threshold now exceed the new one $\to$ new
  swords $\to$ more elimination $\to$ lower $\bar{U}$ $\to$ \ldots
  This is the positive feedback of \cref{thm:paradox}, now given a
  statistical mechanism.

  \item \textbf{Why self-blunting works.} Xiao He's strategy
  (self-corruption to signal low $\|U_r\|$) works precisely because
  the detection function triggers on \emph{deviation from the mean}.
  By visibly degrading his own actuation, Xiao He pulled $\|U_r\|$
  toward $\bar{U}$, falling below the detection threshold.

  \item \textbf{Why breakpoints prevent swords.} A breakpoint in the
  execution chain reduces $\|U_r\|$ (effective autonomous actuation)
  to below $\bar{U} + \tau(\Obs)$, since the king controls part of
  the chain. The resource remains capable but not \emph{independently}
  capable---it does not deviate from the mean.
\end{enumerate}

\begin{remark}[Why the axioms suffice]\label{rem:is-ising}
The viability axiom, the mass-gap equivalence
(\cref{thm:massgap}), and real linear time are the only
primitives the theory requires.  Their operational content
decomposes as:
\begin{enumerate}[label=(\alph*),nosep]
  \item \emph{Lattice structure:} agents occupy sites;
  interactions are nearest-neighbour in the execution graph
  (the Ising skeleton).
  \item \emph{Measure:} the state is a probability measure on
  configurations; the viability kernel $\Viab(K)$ is its
  support.
  \item \emph{Mass gap $=$ openness:} $\lambda_1 > 0$ implies
  the viable phase is an open set in parameter space---robust
  to small perturbations without fine-tuning.
  \item \emph{Uniform law of large numbers:} as
  $n \to \infty$, the empirical mean $\bar{U}$ converges
  uniformly to its population value; the detection threshold
  $\tau(\Obs)$ becomes sharp.
  \item \emph{Real linear time:} the filtration
  $(\mathcal{F}_t)_{t \ge 0}$ is the causal envelope; it
  forbids closed time-like loops and makes the forcing lemma
  (\cref{lem:forcing}) irreversible.
\end{enumerate}
No force law is postulated.  What appears as a restoring force
(the king's reaction to a sword) is the system's regression
toward the maximum-measure configuration under ULLN---the
\emph{entropic} pressure of the viable set.  The following
proposition quantifies the locking precision that these
primitives guarantee.
\end{remark}

\begin{proposition}[Event-locking bound]\label{prop:event-lock}
Let the system satisfy the viability axiom (\cref{ax:viability})
with finite monitoring bandwidth~$B$ and control frequency~$M$.
Let $E$ denote a target event on the external
$\sigma$-algebra, and let $\varepsilon$ parametrise the
perturbation strength (noise, wind, model error).  If the
following four conditions hold:
\begin{enumerate}[label=(\roman*),nosep]
  \item \emph{Viability:} the viable set $\Viab(K)$ is
  non-empty and the tangential condition is satisfiable;
  \item \emph{Projected controllability:} the input vector
  fields, projected onto the viable directions, satisfy the
  Lie-algebraic rank condition (LARC) near the event surface;
  \item \emph{Frequency-aligned realisability:} the required
  control trajectory is realisable within the bandwidth~$M$
  (the action roughness $\sum \|u_{t+1} - u_t\|^2$ remains
  bounded under the Nyquist constraint);
  \item \emph{Event closure:} the event~$E$ is closed under
  the pushforward of the closed-loop policy~$\pi_\theta$
  on the external $\sigma$-algebra;
\end{enumerate}
then the failure probability satisfies
\begin{equation}\label{eq:event-lock}
  1 - p_{\varepsilon,M}(\theta)
  \;\leq\;
  C_1\,r(\varepsilon) + \frac{C_2}{M},
\end{equation}
where $r(\varepsilon) \to 0$ as $\varepsilon \to 0$
(perturbation vanishes) and $C_1, C_2 > 0$ depend on the
system geometry.  In particular,
$p_{\varepsilon,M}(\theta) \to 1$ as $\varepsilon \to 0$
and $M \to \infty$.
\end{proposition}

\begin{definition}[Causality fractal]\label{def:causality-fractal}
Let $G_{\mathrm{task}} = (V, E)$ be an execution DAG
(\cref{def:m-dag}) with layered structure
$V = L_0 \sqcup L_1 \sqcup \cdots$, and let
$Q_\ell(\cdot \mid x, \eta)$ be the mother--daughter kernel
of \cref{rem:parent-daughter-bridge}.  The
\emph{causality fractal} is the triple
$\mathfrak{F} = (G_\infty,\, Q,\, \lambda_1)$:
\begin{enumerate}[label=(\roman*),nosep]
  \item $G_\infty = V \times \mathbb{Z}_{\ge 0}$ is the
  \emph{unrolled DAG} (\cref{rem:causal-mass-gap}): one copy
  of~$V$ per time step, edges pointing strictly forward;
  \item $Q$ is the \emph{self-similar kernel}: the same
  mother--daughter update~$Q_\ell$ acts at every layer under
  exchangeability (within each thermodynamic phase;
  at a phase transition the description re-initialises
  at the new fixed point);
  \item $\lambda_1 > 0$ is the spectral gap
  (\cref{thm:massgap}), serving as the contraction ratio.
\end{enumerate}
The \emph{mean-field operator}
$\mathcal{T} \colon \mathcal{P}(\mathcal{X}) \to
\mathcal{P}(\mathcal{X})$ is the nonlinear pushforward
\begin{equation}\label{eq:mf-operator}
  (\mathcal{T}\mu)(A)
  \;=\;
  \int_{\mathcal{X}}
  Q_\ell\!\bigl(A \mid x,\;\varphi(\mu)\bigr)\,\mu(dx),
\end{equation}
where $\varphi(\mu)$ is the external field
(e.g.\ $\bar{U}$ of \cref{prop:phase}).
\end{definition}

\begin{theorem}[Ergodic convergence on the causality fractal]
\label{thm:fractal-convergence}
Let $\mathfrak{F} = (G_\infty, Q, \lambda_1)$ be a causality
fractal (\cref{def:causality-fractal}) with
$\lambda_1 > 0$.  Suppose the kernel~$Q_\ell$ satisfies:
\begin{enumerate}[label=\textup{(A\arabic*)},nosep]
  \item\label{a:feller} \emph{Feller continuity:}
  $x \mapsto Q_\ell(\cdot \mid x, \eta)$ is weakly continuous
  for each~$\eta$;
  \item\label{a:lip} \emph{Lipschitz external field:}
  $\varphi \colon \mathcal{P}(\mathcal{X}) \to \mathbb{R}^d$
  is Lipschitz in the $1$-Wasserstein metric~$W_1$;
  \item\label{a:doeblin} \emph{Doeblin minorisation:} there
  exist $\delta > 0$ and a probability measure~$\nu$ such that
  $Q_\ell(A \mid x, \eta) \ge \delta\,\nu(A)$ for all
  $x, \eta, A$.
\end{enumerate}
Then:
\begin{enumerate}[label=\textup{(\roman*)},nosep]
  \item \emph{Existence and uniqueness.}
  $\mathcal{T}$ has a unique fixed point
  $\pi \in \mathcal{P}(\mathcal{X})$: \;$\mathcal{T}\pi = \pi$.
  \item \emph{Geometric event convergence.}
  For any initial measure~$\mu_0$,
  \begin{equation}\label{eq:event-conv}
    W_1(\mathcal{T}^n \mu_0,\;\pi)
    \;\leq\;
    (1 - \delta)^{n}\; W_1(\mu_0,\;\pi).
  \end{equation}
  \item \emph{Exponential mixing.}
  For local observables $F_a, G_b$ at causal distance~$d$
  on~$G_\infty$,
  \begin{equation}\label{eq:exp-mixing}
    \lvert\operatorname{Cov}_\pi(F_a, G_b)\rvert
    \;\leq\;
    C\,e^{-\lambda_1\, d}.
  \end{equation}
\end{enumerate}
\end{theorem}

\begin{proof}
\ref{a:doeblin} gives Dobrushin's contraction: the
total-variation coupling between
$Q_\ell(\cdot \mid x, \eta)$ and
$Q_\ell(\cdot \mid x', \eta)$ satisfies
$d_{\mathrm{TV}} \leq 1 - \delta$ for every $x, x'$.
Combined with \ref{a:lip}, the composed map
$\mu \mapsto \mathcal{T}\mu$ is a contraction in~$W_1$
with Lipschitz constant $\kappa \leq 1 - c\,\delta$,
where $c > 0$ depends on the Lipschitz constant
of~$\varphi$.
Since $(\mathcal{P}(\mathcal{X}), W_1)$ is complete
(the state space~$\mathcal{X}$ is Polish), the Banach
fixed-point theorem yields~(i) and~(ii).

For~(iii), each time-slice of~$G_\infty$ inherits the
graph Laplacian of~$G_{\mathrm{task}}$ with the same
spectral gap~$\lambda_1$; since the inter-layer coupling
is via the self-similar kernel~$Q$ with uniform bounds,
the Poincar\'{e} inequality transfers to~$G_\infty$
(cf.\ \cref{thm:cheeger}).
Standard results (Martinelli~\cite{martinelli1999},
Liggett~\cite{liggett2005}) then promote the
Poincar\'{e} inequality to exponential decay of
correlations along the causal direction.
\end{proof}

\begin{corollary}[Event controllability]
\label{cor:event-controllability}
Under the hypotheses of \cref{thm:fractal-convergence},
let $\alpha \in \mathcal{A}$ parametrise the kernel:
$Q_\ell = Q_\ell^\alpha$, and write $\pi(\alpha)$ for the
corresponding fixed point.  If
$\alpha \mapsto \pi(\alpha)$ is Lipschitz with constant
$L_\pi$ (structural stability), then:
\begin{enumerate}[label=\textup{(\roman*)},nosep]
  \item \emph{Steerability.}  For any two viable
  parameters $\alpha_0, \alpha_1$ and any
  $\varepsilon > 0$, there exists a parameter
  trajectory $\alpha(t)$ such that
  $W_1(\mu_t,\, \pi(\alpha(t))) < \varepsilon$ for
  all~$t$.
  \item \emph{Tracking bound.}  If $\alpha(t)$ varies
  at rate $\|\dot{\alpha}\| \leq \rho$, the tracking
  error satisfies
  \begin{equation}\label{eq:tracking}
    W_1\!\bigl(\mu_t,\;\pi(\alpha(t))\bigr)
    \;\leq\;
    \frac{L_\pi\,\rho}{\delta}
    \;+\;
    (1 - \delta)^{t}\,W_1(\mu_0,\;\pi(\alpha_0)).
  \end{equation}
  \item \emph{Loss at phase transition.}  As
  $\delta \to 0$ the contraction weakens and the
  tracking bound~\eqref{eq:tracking} diverges:
  the system cannot follow control inputs faster than
  the vanishing mixing rate.  Since $\lambda_1 \to 0$
  forces $\delta \to 0$ (a disconnected graph cannot
  satisfy Doeblin minorisation), the phase transition
  of \cref{prop:phase} entails loss of event
  controllability.
\end{enumerate}
\end{corollary}

\begin{proof}
At each step the error decomposes as
\[
  W_1(\mu_{t+1},\, \pi(\alpha(t{+}1)))
  \;\leq\;
  \underbrace{(1 - \delta)\,
    W_1(\mu_t,\, \pi(\alpha(t)))}_{\text{contraction}}
  \;+\;
  \underbrace{L_\pi\,
    \|\alpha(t{+}1) - \alpha(t)\|}_{\text{target drift}}.
\]
Setting $e_t = W_1(\mu_t, \pi(\alpha(t)))$ and
$\rho = \sup_t \|\alpha(t{+}1) - \alpha(t)\|$, we get
$e_{t+1} \leq (1 - \delta)\,e_t + L_\pi \rho$.
Iterating:
$e_t \leq (1-\delta)^t e_0 + L_\pi \rho
\sum_{k=0}^{t-1}(1-\delta)^k
\leq (1-\delta)^t e_0 + L_\pi \rho / \delta$.
This gives~(ii); (i) follows by choosing
$\rho < \varepsilon\,\delta / L_\pi$;
(iii) is immediate from $\delta \to 0$.
\end{proof}

\begin{remark}[The renormalisation-group interpretation]
\label{rem:rg}
The mean-field operator~$\mathcal{T}$
(\cref{eq:mf-operator}) is a renormalisation-group (RG)
transformation: each application coarse-grains one causal
layer of~$G_\infty$.  The fixed point
$\pi = \mathcal{T}\pi$ is the RG fixed point;
$\lambda_1$ is the leading relevant eigenvalue; and the
phase transition $\lambda_1 \to 0$
(\cref{prop:phase}) is the approach to a critical RG
fixed point where the system becomes scale-invariant.
The causality fractal is, precisely, the RG orbit in
measure space---and \cref{thm:fractal-convergence} says
this orbit converges whenever the mass gap is open.
\end{remark}
