\section{The five-phase cycle}
\label{sec:phase-cycle}

\subsection{Phase structure}
\label{sec:phase-structure}

The two-crossing lemma (\cref{lem:two-crossing}) reduces every
admissible pick--place trajectory to a fixed combinatorial type:
between the upward crossing at $t_{1}$ and the downward crossing
at $t_{4}$, the contact order parameter satisfies $\rcontact > 1$
(the manipulator supports the object); outside the closed interval
$[t_{1}, t_{4}]$, $\rcontact < 1$ (the robot is out of contact
with the object). The two crossings themselves, the in-contact
interior, and the two out-of-contact intervals together form
five distinguishable arcs of $[0, T]$, and the present subsection
labels them. This labeling is forced---no implementation choice
enters---and it is the combinatorial scaffold against which the
operator algebra of \cref{sec:operator-algebra} and the
$\Greflect$-equivariance of \cref{sec:G-equivariance} are stated.

\begin{definition}[Five-phase decomposition]\label{def:five-phase}
A pick--place trajectory satisfying the boundary conditions
\eqref{eq:pick-place-BCs} of \cref{lem:two-crossing} decomposes
into five phases, indexed $k = 1, \ldots, 5$, with the following
arc types and order-parameter regimes:

\begin{center}
\small
\begin{tabular}{@{}clccc@{}}
\toprule
$k$ & Phase & Time interval & $\rcontact$ & $\specgap$ \\
\midrule
1 & pick (approach) & $[0, t_{1}]$ & $< 1$ & --- \\
2 & load (contact establishment) & $\{t_{1}\}$ & $\uparrow 1$ & $0 \to \varepsilon$ \\
3 & use (transport) & $(t_{1}, t_{4})$ & $> 1$ & $\ge \varepsilon$ \\
4 & unload (release) & $\{t_{4}\}$ & $\downarrow 1$ & $\varepsilon \to 0$ \\
5 & place (retreat) & $[t_{4}, T]$ & $< 1$ & --- \\
\bottomrule
\end{tabular}
\end{center}

Phases 1 and 5 are singular arcs of the OCP \eqref{eq:ocp};
phases 2 and 4 are bang arcs concentrated at the
$\swsurf$-crossings; phase 3 is a singular arc subject to the
spectral-gap maintenance condition $\specgap \ge \varepsilon$.
\end{definition}

\paragraph{Per-phase reading of the three-term decomposition.}
The three-term decomposition (\cref{thm:three-term}) lights up
in different combinations across the five phases of
\cref{def:five-phase}. On phase 1 (pick) the trajectory is a
pure singular arc with no $\swsurf$-crossing in its interior, so
the spectral-kick term vanishes ($u_{s} = 0$) and the control is
$u = u_{g} + u_{a}$: gravity compensation plus the least-action
pre-contact reach. On phase 2 (load) the spectral-kick term
$u_{s}$ is active at $t_{1}$, producing the impulsive
contribution that drives $\rcontact$ across $\swsurf$ upward and
brings $\specgap$ from $0$ up to the threshold $\varepsilon$. On
phase 3 (use) the trajectory is again a singular arc on the open
interior $(t_{1}, t_{4})$, so $u_{s} = 0$ and the control returns
to $u = u_{g} + u_{a}$, now subject to the spectral-gap
maintenance constraint $\specgap \ge \varepsilon$ enforced
through the inner level of \cref{def:bilevel}. On phase 4
(unload) the spectral-kick term is active again at $t_{4}$,
driving $\rcontact$ back across $\swsurf$ downward and releasing
the contact graph spectral gap. On phase 5 (place) the
trajectory is once more a pure singular arc with $u = u_{g} +
u_{a}$ and $u_{s} = 0$: the post-contact retreat to the home
configuration.

\paragraph{Hybrid-systems reading.}
\Cref{def:five-phase} is the hybrid-automaton instantiation of
the pick--place OCP. Each phase $k$ is a discrete mode of the
automaton, the in-contact and out-of-contact regimes of $\rcontact$
are the continuous-time invariants of the corresponding modes,
and the two transitions $\swsurf$-upward at $t_{1}$ and
$\swsurf$-downward at $t_{4}$ are guard conditions in the sense
of \citet{Lygeros2003} and \citet{GoebelSanfeliceTeel2012}. The
structural force of \cref{lem:two-crossing} is that this hybrid
automaton is determined: any admissible trajectory factors
through these five modes in the prescribed order, with the two
guard crossings exactly once each. There is no admissible
manipulation episode that visits the five modes in a different
sequence.

\paragraph{Forward references.}
\Cref{sec:operator-algebra} associates a structural operator
$\phaseop{k-1}$ to each phase $k$, and \cref{sec:G-equivariance}
has already shown that the resulting family
$\{\phaseop{0}, \ldots, \phaseop{4}\}$ is $\Greflect$-equivariant.
\Cref{sec:hierarchical} will specialize the task-space /
null-space decomposition (\cref{def:TS-NS}) per phase.

\subsection{Operator algebra}
\label{sec:operator-algebra}

To each phase of \cref{def:five-phase} we associate a structural
operator on the joint-space state of the manipulator--object
system. The collection $\{\phaseop{0}, \phaseop{1}, \phaseop{2},
\phaseop{3}, \phaseop{4}\}$ forms a closed loop under composition
in the following precise sense: the composition returns the
manipulator to its home configuration, but it does not return the
object to its initial pose. This asymmetry between the
manipulator (which closes) and the object (which is transported)
is the structural source of irreversibility in manipulation; the
present subsection states the operators, the loop theorem, and
the irreversibility paragraph that names the asymmetry.
\Cref{fig:operator-loop} depicts the cycle and the corresponding
$\rcontact$ trace.

\paragraph{State space.}
Throughout this subsection the operating state is
\begin{equation}
\label{eq:operator-state}
  s \;=\; (\jointvec, \dot{\jointvec}, x_{\mathrm{obj}})
  \;\in\; \mathbb{R}^{n} \times \mathbb{R}^{n} \times SE(3),
\end{equation}
with $\jointvec$ the manipulator joint configuration,
$\dot{\jointvec}$ the joint velocity, and
$x_{\mathrm{obj}} \in SE(3)$ the object pose. The contact graph
$G_{c}(t)$ of \cref{sec:specgap} is not a state component: it is
\emph{observed} from the per-timestep convex contact solve and
the perception pipeline, in the sense made precise by
\cref{sec:smooth-contact}. Consequently $\rcontact$, $\specgap$,
and $G_{c}$ enter the operator definitions below as derived
quantities along the trajectory rather than as carried state.

\begin{definition}[Phase operators]\label{def:phase-operators}
Each phase of \cref{def:five-phase} has an associated structural
operator on the joint-space state $s = (\jointvec, \dot{\jointvec},
x_{\mathrm{obj}})$:
\begin{enumerate}[label=(\roman*),nosep]
  \item $\phaseop{0}$ (\emph{pick}, phase 1):
        $s_{0} \mapsto s_{t_{1}^{-}}$. Moves the manipulator from
        a home configuration $\jointvec_{0}$ to the pre-grasp pose
        $\jointvec_{t_{1}^{-}}$. The object pose $x_{\mathrm{obj}}$
        is unchanged (no robot--object contact during phase 1).
  \item $\phaseop{1}$ (\emph{load}, phase 2):
        $s_{t_{1}^{-}} \mapsto s_{t_{1}^{+}}$. Crosses $\swsurf$
        upward ($\rcontact: <1 \to >1$). Establishes the contact
        graph: $\specgap$ rises from $0$ to at least $\varepsilon$.
        Joint configuration may change by an amount within the
        precision allowed by the absolutely continuous
        spectral-kick term $u_{s}$ of \cref{thm:three-term},
        concentrated on a small interval surrounding $t_{1}$.
  \item $\phaseop{2}$ (\emph{use}, phase 3):
        $s_{t_{1}^{+}} \mapsto s_{t_{4}^{-}}$. Singular arc of
        the OCP \eqref{eq:ocp} under the spectral-gap
        maintenance constraint $\specgap \ge \varepsilon$. The
        object pose evolves from the initial support
        configuration $x_{\mathrm{obj}}^{A}$ to the destination
        configuration $x_{\mathrm{obj}}^{B}$; the joint
        configuration tracks the object trajectory through the
        bi-level scaffold of \cref{def:bilevel}.
  \item $\phaseop{3}$ (\emph{unload}, phase 4):
        $s_{t_{4}^{-}} \mapsto s_{t_{4}^{+}}$. Crosses $\swsurf$
        downward ($\rcontact: >1 \to <1$). Releases the contact
        graph: $\specgap$ drops from $\ge \varepsilon$ to $0$.
  \item $\phaseop{4}$ (\emph{place}, phase 5):
        $s_{t_{4}^{+}} \mapsto s_{T}$. Returns the manipulator
        from the post-release pose $\jointvec_{t_{4}^{+}}$ toward
        the home configuration $\jointvec_{T} = \jointvec_{0}$.
        The object pose $x_{\mathrm{obj}}$ is unchanged in phase
        5 (no robot--object contact).
\end{enumerate}
\end{definition}

\input{figures/operator-loop}

\begin{theorem}[Joint-space loop with object transport]
\label{thm:joint-loop}
Let $\phaseop{0}, \ldots, \phaseop{4}$ be the phase operators of
\cref{def:phase-operators}. Their composition acts on the initial
state
$s_{0} = (\jointvec_{0}, \dot{\jointvec}_{0},
x_{\mathrm{obj}}^{A})$
as
\begin{equation}
\label{eq:joint-loop}
  (\phaseop{4} \circ \phaseop{3} \circ \phaseop{2} \circ
   \phaseop{1} \circ \phaseop{0})(s_{0})
  \;=\; (\jointvec_{0}, \dot{\jointvec}_{0},
         x_{\mathrm{obj}}^{B}),
\end{equation}
where $x_{\mathrm{obj}}^{B} \neq x_{\mathrm{obj}}^{A}$ is the
destination pose of the manipulated object. The manipulator
forms a closed loop in joint space; the object does not.
\end{theorem}

\begin{proof}
By the boundary conditions \eqref{eq:pick-place-BCs} of
\cref{lem:two-crossing}, $\rcontact(0) < 1$ and $\rcontact(T) <
1$, so phases 1 and 5 are out-of-contact arcs by
\cref{def:five-phase}. On those phases the object is not
coupled to the manipulator and $x_{\mathrm{obj}}$ is unchanged
across $\phaseop{0}$ and $\phaseop{4}$. The home-configuration
specification $\jointvec_{T} = \jointvec_{0}$ is built into
$\phaseop{4}$ by \cref{def:phase-operators}~(v); together with
the corresponding velocity boundary condition
$\dot{\jointvec}_{T} = \dot{\jointvec}_{0}$, this returns the
manipulator coordinates to their initial values. The object
coordinate, by contrast, changes only on phase 3, where
\cref{def:phase-operators}~(iii) carries $x_{\mathrm{obj}}^{A}$
to $x_{\mathrm{obj}}^{B}$ along the singular arc with $\rcontact
> 1$; no subsequent operator reverses this transport, since
phases 4 and 5 act trivially on $x_{\mathrm{obj}}$ (phase 4
releases the contact instantaneously at $t_{4}$ without object
motion; phase 5 is out of contact). The composition therefore
satisfies \eqref{eq:joint-loop}. The associated hybrid-automaton
factorization is the one identified in \citet{Lygeros2003}.
\end{proof}

\paragraph{Irreversibility lives in the object.}
\Cref{thm:joint-loop} elevates the asymmetry between manipulator
and object to a structural claim. The manipulator returns to
its home configuration $\jointvec_{0}$: the joint-space cycle is
closed, and in this sense the manipulator's trajectory is
reversible (the same operator family run in time-reversed order
on a time-reversed trajectory would return both endpoints). The
object, on the other hand, does not return: it is transported
from $x_{\mathrm{obj}}^{A}$ to $x_{\mathrm{obj}}^{B} \neq
x_{\mathrm{obj}}^{A}$, and no operator in
\cref{def:phase-operators} reverses this. The irreversibility
of a manipulation episode is therefore not in the manipulator
but in the object: the cycle does work on the object precisely
because the manipulator's joint-space loop closes around an
object-space displacement. This is the structural reason a
pick--place episode is a unit of useful work, and it is the
property that the perception, planning, and control layers of
\cref{sec:hierarchical,sec:perception,sec:plan-training}
must collectively preserve.

\paragraph{Forward references.}
\Cref{sec:G-equivariance} has already shown that the family
$\{\phaseop{0}, \ldots, \phaseop{4}\}$ is $\Greflect$-equivariant;
\cref{sec:hierarchical} will specialize the task-space /
null-space decomposition (\cref{def:TS-NS}) per operator,
identifying for each $\phaseop{i}$ what the task variable is
and what the null-space freedom encodes.

% ---------------------------------------------------------------
\subsection{$G$-equivariance of the five-phase cycle}
\label{sec:G-equivariance}

The framework so far has treated each manipulation episode as a
single trajectory: one robot, one object, one ground. The same
operator algebra in fact applies---without modification---to a
family of dynamical configurations related by a discrete symmetry.
Two configurations make this concrete. In a basketball dribble,
the hand strikes the ball downward at the apex of a bounce; in
ball juggling, the hand strikes the ball upward from below. Both
admit the same five-phase decomposition (free-flight, approach,
impact, separation, free-flight) with the same control structure
(gravity compensation plus least-action interior plus a
spectral-kick at each $\swsurf$-crossing). The present subsection
identifies the symmetry group $\Greflect$ that relates such
configurations, formalizes its action on the extended state, and
proves that the operator algebra of \cref{sec:operator-algebra}
is $\Greflect$-equivariant. The corollary at the end of the
subsection provides the structural reason a single controller
realizes both branches of the orbit; Stage~III of the experimental
staircase (\cref{sec:exp-stage-III}) is the corresponding
deployment.

\begin{definition}[Reflection group $\Greflect$]
\label{def:G-action}
Let $\Greflect = \{e, \sigma\}$ be the discrete reflection group
of order two, with $\sigma^{2} = e$. The group $\Greflect$ acts
on the $z$-component of the extended state $(z, \dot z, g_{z},
F_{z})$ by simultaneous sign-reversal:
\[
  \sigma \cdot (z, \dot z, g_{z}, F_{z})
  \;=\;
  (-z, -\dot z, -g_{z}, -F_{z}),
\]
where $z, \dot z$ are the vertical position and velocity, and
$g_{z}, F_{z}$ are the $z$-components of the gravitational
acceleration and the relevant external force, respectively. The
horizontal components $(x, y, \dot x, \dot y, g_{x}, g_{y},
F_{x}, F_{y})$ are fixed by $\sigma$.
\end{definition}

\begin{remark}[$\Greflect$ is a coordinate symmetry, not a physical symmetry]
\label{rem:G-coord-sym}
The action of $\sigma$ in \cref{def:G-action} is a symmetry of
the \emph{equations of motion under coordinate relabeling}, not a
physical symmetry of the laboratory frame: gravitational
acceleration in the laboratory always points downward. The claim
of $\Greflect$-equivariance is that the dynamics written in
$(z, \dot z, g_{z}, F_{z})$ coordinates is identical to the
dynamics written in $(-z, -\dot z, -g_{z}, -F_{z})$ coordinates,
so that configurations related by $\sigma$ admit the same
controller modulo a deterministic sign-flip on the control
input. This is the formal counterpart to the empirical
observation that a basketball dribble and a ball-juggle are run
by the same software with a single sign parameter.
\end{remark}

\begin{lemma}[$\Greflect$-equivariance of the five-phase cycle]
\label{lem:G-equivariance}
Let $\Greflect = \{e, \sigma\}$ act on the extended state
$(z, \dot z, g_{z}, F_{z})$ as in \cref{def:G-action}, and let
$\rcontact, \swsurf, \phaseop{0}, \ldots, \phaseop{4}$ denote
the order parameter, switching surface, and phase operators of
\cref{sec:rho,sec:Sigma,sec:operator-algebra}. Then:
\begin{enumerate}[label=(\roman*),nosep]
  \item the order parameter is $\Greflect$-invariant:
        $\rcontact(\sigma \cdot \xi) = \rcontact(\xi)$ for every
        admissible state $\xi$;
  \item the switching surface is $\Greflect$-invariant:
        $\sigma(\swsurf) = \swsurf$;
  \item the phase operators commute with $\sigma$:
        $\phaseop{i} \circ \sigma = \sigma \circ \phaseop{i}$ for
        $i = 0, \ldots, 4$;
  \item the three-term control decomposition of
        \cref{thm:three-term} is linearly $\Greflect$-covariant:
        $\sigma$ acts on each component by sign-reversal,
        $\sigma \cdot u_{g} = -u_{g}$,
        $\sigma \cdot u_{a} = -u_{a}$,
        $\sigma \cdot u_{s} = -u_{s}$.
\end{enumerate}
\end{lemma}

The proof is deferred to \cref{app:proof:G-equivariance}.

\begin{corollary}[Single controller across the $\Greflect$-orbit]
\label{cor:G-controller}
A controller $\pi$ whose feedback law depends on the state only
through $\Greflect$-invariant quantities---in particular through
$\rcontact$, $\specgap$, and the contact gap $\phi$---is
automatically $\Greflect$-equivariant: applying $\pi$ to state
$\xi$ and applying $\pi$ to state $\sigma \cdot \xi$ produce
control inputs related by $\sigma \cdot u = -u$. A single such
controller therefore realizes both branches of the
$\Greflect$-orbit. Stage~III (\cref{sec:exp-stage-III})
instantiates this corollary in the dual-dribble experiment, in
which one controller drives both basketball dribble and ball
juggling without per-mode tuning.
\end{corollary}

The corollary is the structural reason the same code base
handles both Stage~III dual-dribble configurations: the feedback
law depends on the state only through $\Greflect$-invariant
scalars, so the sign-flip prescribed by $\sigma$ is absorbed by
the linear covariance of the three-term decomposition. The
basketball-versus-juggle pair is the simplest non-trivial example
of a $\Greflect$-orbit of manipulation tasks; the same machinery
extends, mutatis mutandis, to richer discrete symmetries that
relate larger families of pick--place configurations under the
same operator algebra. \Cref{sec:exp-stage-III} reports the
dual-dribble deployment of this single-controller architecture.
