\chapter{夏虫运冰}\label{sec:manipulation}

\begin{center}
\itshape
夏虫不可以语于冰者，笃于时也。\\[4pt]
\upshape ---庄子，《秋水》
\end{center}

\bigskip

The summer insect cannot \emph{discuss} ice---it has never observed
ice, and its causal domain excludes winter.  But the summer insect can
\emph{transport} ice, if three conditions hold: (i)~it can observe the
ice within its field of view (causal observability), (ii)~it can reach
the ice (kinematic feasibility), and (iii)~it can grasp the ice before
it melts (information timeliness).  This chapter formalises robotic
manipulation as 夏虫运冰: the agent (robot arm) transports objects
(blocks) through the five-phase cycle of the sword lifecycle
(\cref{thm:lifecycle}), subject to the cognitive boundary of its
sensors.

The argument proceeds in eight stages:
\begin{enumerate}
  \item \textbf{Phase structure} (\cref{sec:m-phases}): every
  manipulation task decomposes into five phases forced by
  $\rho$ crossing $\Sigma$ twice.
  \item \textbf{Operator algebra} (\cref{sec:m-operators}): the
  phase operators form a loop in robot joint space; irreversibility
  lives in the object.
  \item \textbf{RANK/NULL decomposition} (\cref{sec:m-ranknull}):
  task forces are the real part, self-motion is the imaginary part,
  and their commutator measures dexterity.
  \item \textbf{Bi-level structure} (\cref{sec:m-bilevel}):
  each pick--place cycle nests force closure maintenance (inner)
  inside trajectory optimisation (outer); grasping is the
  deliberate creation of a contact singularity.
  \item \textbf{Observer} (\cref{sec:m-observer}): the cognitive
  boundary of a wrist-mounted camera, formalised via causal
  observability.
  \item \textbf{Plan generator} (\cref{sec:m-plangen}): the upper
  level of the two-level architecture---discrete, stochastic,
  and the only tractable approach.
  \item \textbf{Training} (\cref{sec:m-training}): measure matching
  (MMD) is the only reward; conditional flow matching trains the
  velocity field; the network is a phase-conditioned MLP.
  \item \textbf{Curriculum} (\cref{sec:m-curriculum}): four
  orthogonal complexity axes---stack layers, block count, block
  types, and adversarial wind---each testing a different
  capability.
\end{enumerate}


% ═══════════════════════════════════════════════════════════
\section{Phase structure: the five-phase cycle}
\label{sec:m-phases}

\subsection{The contact order parameter in manipulation}

As in the three-body system (\cref{sec:3b-ocp}), the state of the
robot--object system is characterised by the contact order parameter
$\rho$ and the spectral gap $\lambda_1$ of the contact graph
Laplacian.  The manipulation setting differs from the celestial
setting in one respect: contact is \emph{binary} (finger either
touches object or does not), whereas gravitational coupling is
continuous.  Consequently, $\rho$ transitions sharply at the
switching surface $\Sigma = \{\rho = 1\}$.

\begin{definition}[Manipulation phases]\label{def:m-phases}
A single pick--place cycle consists of five phases determined by
the trajectory of $\rho(t)$:
\begin{center}
\begin{tabular}{@{}clccl@{}}
\toprule
\textbf{Phase} & \textbf{Robot act} & $\rho$ & $\lambda_1$
  & \textbf{Arc type} \\
\midrule
1 & pick (approach) & $< 1$ & --- & singular \\
2 & load (contact)  & $\uparrow 1$ & $0 \to \ge\epsilon$
  & bang \\
3 & use (transport)  & $> 1$ & $\ge \epsilon$
  & singular \\
4 & unload (release) & $\downarrow 1$ & $\ge\epsilon \to 0$
  & bang \\
5 & place (retreat)  & $< 1$ & --- & singular \\
\bottomrule
\end{tabular}
\end{center}
This decomposition is forced: contact is established exactly once
($\rho$ crosses $\Sigma$ upward) and released exactly once
($\rho$ crosses $\Sigma$ downward).  No valid manipulation
trajectory can avoid both crossings.
\end{definition}

\begin{remark}[Connection to the three-term decomposition]
\label{rem:m-three-term}
The three-term structure of \cref{thm:3b-damper} instantiates at
each phase: Term~(I) is gravity on the object, Term~(II) is the
minimum-energy trajectory of the robot, and Term~(III) is the
spectral kick at $\Sigma$.  Phases 1 and 5 are purely
singular (Term~III $= 0$); phases 2 and 4 are bang arcs
(Term~III active at $\Sigma$); phase 3 is a singular arc with
spectral maintenance ($\lambda_1 \ge \epsilon$ enforced).
\end{remark}


% ═══════════════════════════════════════════════════════════
\section{Operator algebra: the joint-space loop}
\label{sec:m-operators}

\subsection{Phase operators}

Each phase has an associated operator on the state space:

\begin{definition}[Phase operators]\label{def:m-operators}
\leavevmode
\begin{enumerate}[label=(\roman*),nosep]
  \item $T_0$ \textbf{(pick):} approach operator. Moves robot from
  home configuration $q_0$ to pre-grasp pose.  Object unchanged.
  \item $T_1$ \textbf{(load):} contact-building operator. Crosses
  $\Sigma$ upward ($\rho \colon {<}\,1 \to {>}\,1$).  Establishes
  $\lambda_1 \ge \epsilon$.  This is the bang arc.
  \item $T^*$ \textbf{(use):} the adjoint of $T_1$.  With contact
  established, the robot transports the object.
  $T_1$ maps forces to motions (building the grasp); $T^*$ maps
  motions to forces (exploiting the grasp).  In Kalman terms
  (\cref{thm:3b-kalman}): $T_1$ is observability, $T^*$ is
  controllability.
  \item $T_1^{-1}$ \textbf{(unload):} inverse of loading.
  Crosses $\Sigma$ downward.  $\lambda_1$ drops below $\epsilon$.
  \item $T_0^{-1}$ \textbf{(place):} return operator.  The robot
  deposits the object and retracts to $q_0$.
\end{enumerate}
\end{definition}

\begin{theorem}[Joint-space loop]\label{thm:m-loop}
The robot cycle closes in joint space:
\[
  q_0 \xrightarrow{T_0} q_\mathrm{grasp}
  \xrightarrow{T_1} q_\mathrm{ready}
  \xrightarrow{T^*} q_\mathrm{done}
  \xrightarrow{T_1^{-1}} q_\mathrm{release}
  \xrightarrow{T_0^{-1}} q_0.
\]
The composite operator is the identity on robot state but not on
object state:
\begin{align}
  \pi_\mathrm{robot} \circ
    (T_0^{-1} \circ T_1^{-1} \circ T^* \circ T_1 \circ T_0)
    &= \mathrm{Id}_\mathrm{robot}, \label{eq:m-robot-loop} \\
  \pi_\mathrm{obj} \circ
    (T_0^{-1} \circ T_1^{-1} \circ T^* \circ T_1 \circ T_0)
    &\neq \mathrm{Id}_\mathrm{obj}. \label{eq:m-obj-irrev}
\end{align}
\end{theorem}

\begin{proof}
Equation~\eqref{eq:m-robot-loop}: $T_0^{-1}$ is the kinematic
inverse of $T_0$ (same path reversed in joint space), and
similarly $T_1^{-1}$ inverts $T_1$ on the robot coordinates.
The adjoint $T^*$ acts on the object's $SE(3)$ coordinates via
the contact Jacobian; on the robot's joint coordinates it is a
diffeomorphism that depends on the grasped object but preserves
the robot's reachable set.  The composition returns the robot to
$q_0$.

Equation~\eqref{eq:m-obj-irrev}: $T^*$ transports the object from
$g_\mathrm{init}$ to $g_\mathrm{target}$, and no subsequent
operator reverses this.  The object's pose change is the
irreversible residue of the cycle.
\end{proof}

\begin{remark}[Separation of concerns]\label{rem:m-separation}
The robot is cyclic ($q_0 \to q_0$), the object is irreversible
($g \to g'$), and the $\rho$ trajectory is cyclic
($0 \to 1 \to {>}1 \to 1 \to 0$).  This separation makes each
per-block cycle self-contained: the upper-level planner issues
``do block $k$'' and the cycle executes without residual state
in the robot.  This is the manipulation analogue of temporal
irreversibility (\cref{rem:dumu-irreversibility}): the
pawn-to-sword transition is hysteretic, and in manipulation the
object's pose change is the irreversible record.
\end{remark}


% ═══════════════════════════════════════════════════════════
\section{RANK/NULL decomposition: Re and Im}
\label{sec:m-ranknull}

\subsection{The task Jacobian}

Let $q \in \R^n$ be the robot's joint configuration ($n = 7$ for
the Franka Emika Panda) and $x \in SE(3) \cong \R^6$ the task-space
pose of the end-effector.  The \emph{task Jacobian}
$J(q) \in \R^{6 \times n}$ maps joint velocities to task-space
velocities: $\dot{x} = J(q)\,\dot{q}$.

\begin{definition}[Re and Im of manipulation]\label{def:m-reim}
\leavevmode
\begin{enumerate}[label=(\roman*),nosep]
  \item \textbf{Re} $= \mathrm{RANK}(J^\top)$: the range of $J^\top$
  in joint space.  Joint torques in $\mathrm{RANK}(J^\top)$ produce
  task-space forces---they \emph{do work} on the object.  This is
  the self-adjoint (real) part: $S = \tfrac{1}{2}(T + T^*)$.
  \item \textbf{Im} $= \mathrm{NULL}(J)$: the null space of $J$.
  Joint velocities in $\mathrm{NULL}(J)$ produce no task-space
  motion---they reconfigure the robot internally (elbow motion,
  wrist flip) without moving the object.  This is the
  skew-self-adjoint (imaginary) part: $A = \tfrac{1}{2}(T - T^*)$.
\end{enumerate}
\end{definition}

For the Franka Panda: $n = 7$, $\dim(\mathrm{RANK}) = 6$,
$\dim(\mathrm{NULL}) = 1$.  The single null-space direction is the
\emph{elbow self-motion}: the elbow swivels without changing the
hand's pose.

\begin{theorem}[Commutator and dexterity]\label{thm:m-commutator}
Let $S = \mathrm{RANK}(J^\top)$ and $A = \mathrm{NULL}(J)$.
Then $[S, A] \neq 0$ if and only if the robot can modulate
task-space forces by reconfiguring its internal state.
\begin{enumerate}[label=(\roman*),nosep]
  \item $[S, A] \neq 0$: \textbf{dexterous manipulation.}  The
  null-space motion changes the arm's inertia tensor, which
  changes the torques produced by the same joint commands.
  The elbow position affects the force at the fingertips.
  \item $[S, A] = 0$: \textbf{power grasp.}  Internal
  reconfiguration does not affect task forces.  The arm is
  kinematically redundant but dynamically non-dexterous.
\end{enumerate}
\end{theorem}

\begin{proof}
The inertia matrix $M(q)$ couples Re and Im: a null-space velocity
$\dot{q}_\mathrm{null} \in \mathrm{NULL}(J)$ changes $q$, hence
$M(q)$, hence $J^\top(q) F$ for the same task force $F$.  The
commutator $[S, A]$ is the linearisation of this coupling:
\[
  [S, A]_q = \left.\frac{\partial}{\partial \epsilon}\right|_0
  \left( J^\top(q + \epsilon\,\dot{q}_\mathrm{null}) F
  - J^\top(q) F \right).
\]
This vanishes if and only if $M(q)$ is independent of the
null-space coordinates---i.e., the arm's inertia is insensitive
to elbow position.  For the Franka Panda, $M(q)$ depends on all
7 joint angles, so $[S, A] \neq 0$ generically.
\end{proof}

\begin{remark}[Connection to the paper's $T = S + A$ decomposition]
\label{rem:m-cartesian}
The Cartesian decomposition $T = S + A$ (self-adjoint plus
skew-self-adjoint) is standard operator theory.  The physical
content is: $S$ corresponds to dissipation/friction (time-reversal
even), $A$ corresponds to inertia/Coriolis (time-reversal odd),
and $[S, A] \neq 0$ means friction couples to rotation---the
same mechanism that maintains the spectral gap in
\cref{rem:3b-superconductor}.
\end{remark}


% ═══════════════════════════════════════════════════════════
\section{Bi-level structure: grasping nested within transport}
\label{sec:m-bilevel}

The five-phase cycle contains an inner optimisation problem that
is invisible in the operator notation
$T_0 \to T_1 \to T^* \to T_1^{-1} \to T_0^{-1}$ but dominates
the physics.

\begin{definition}[Bi-level pick--place]\label{def:m-bilevel}
A single pick--place cycle has two coupled optimisation levels:
\begin{enumerate}[label=(\roman*),nosep]
  \item \textbf{Outer (trajectory):} choose a velocity field
  $v_\theta(x,t)$ that transports the object from $g^A$ to
  $g^B \in SE(3)$, minimising the OT cost
  (\cref{sec:3b-ocp}).
  \item \textbf{Inner (force closure):} at every instant
  $t \in [t_1, t_4]$ (phases 2--4), choose finger forces
  $f(t) = (f_L, f_R)$ that maintain the contact graph Laplacian's
  spectral gap: $\lambda_1(L_G(f)) \geq \varepsilon$.
\end{enumerate}
The outer level sets the desired acceleration $\ddot{x}_\mathrm{des}(t)$;
the inner level must produce forces that satisfy Newton's equation at
that acceleration while remaining inside the friction cone.
\end{definition}

\begin{theorem}[Bi-level coupling]\label{thm:m-bilevel}
The outer trajectory and inner force closure are coupled
bidirectionally:
\begin{enumerate}[label=(\roman*),nosep]
  \item \emph{Outer $\to$ inner:} the commanded acceleration
  $\ddot{x}_\mathrm{des}(t)$ enters the Newton constraint
  \begin{equation}\label{eq:m-newton}
    \sum_i f_i = m\bigl(\ddot{x}_\mathrm{des}(t) - g\bigr)
    + F_\mathrm{wind}(t).
  \end{equation}
  Faster trajectories demand larger contact forces, consuming
  friction margin.
  \item \emph{Inner $\to$ outer:} force closure feasibility imposes
  a state-dependent acceleration bound
  \begin{equation}\label{eq:m-accel-bound}
    |\ddot{x}_z(t)|
    \;\leq\;
    \frac{2\mu_s\, N_\mathrm{max} - |F_\mathrm{wind}^\perp(t)|}{m}
    - g,
  \end{equation}
  where $\mu_s$ is the friction coefficient, $N_\mathrm{max}$ is
  the maximum actuator normal force, and $F_\mathrm{wind}^\perp$
  is the wind component perpendicular to the contact normal.
  This constrains the set of feasible outer trajectories.
\end{enumerate}
\end{theorem}

\begin{proof}
For a parallel-jaw grasp with two symmetric frictional contacts,
force closure requires that the total friction force span the
required wrench:
$\sqrt{f_{t1}^2 + f_{t2}^2} \leq \mu_s f_n$ per contact
(Coulomb cone), with unilateral constraint $f_n \geq 0$.
The total vertical friction capacity is $2\mu_s N$, where
$N = f_n$ is the clamping force.  The vertical load is
$m|\ddot{x}_z + g| + |F_\mathrm{wind}^\perp|$.  Force closure
$(\lambda_1 \geq \varepsilon)$ holds iff
$2\mu_s N \geq m|\ddot{x}_z + g| + |F_\mathrm{wind}^\perp|$.
Rearranging for the acceleration gives~\eqref{eq:m-accel-bound}.
\end{proof}

\begin{remark}[Force closure as a contact singularity]
\label{rem:m-singularity}
Grasping is the deliberate creation of a \emph{manipulation
singularity}.  Before contact ($\rho < 1$), the object has 6
unconstrained DOF in $SE(3)$: the grasp matrix $G$ has rank~0
(no wrench can be transmitted).  At force closure
($\lambda_1 \geq \varepsilon$), the frictional contacts span
the full wrench space: $\mathrm{rank}(G)$ rises to~6 and the
object's unconstrained DOF drops from~6 to~0.  The switching
surface $\Sigma = \{\rho = 1\}$ is the bifurcation point where
this rank transition occurs---the topology of the configuration
space changes from disconnected (gripper and object independent)
to connected (one coupled system).

The spectral gap $\lambda_1$ measures the \emph{margin} of the
singularity: how deeply the system sits inside the force-closure
manifold.  Wind and inertial loads attempt to restore the
object's free DOF (break the singularity); the inner
optimisation maintains it.
The five-phase cycle is a controlled creation (Phase~2),
maintenance (Phase~3), and dissolution (Phase~4) of this
contact singularity.

In the language of the sword lifecycle (\cref{thm:lifecycle}):
creating the singularity = creating the sword (the object acquires
autonomous actuation through contact---condition~(1) of
\cref{def:sword}).  The locked state (Phase~3) = absorption (the
object is absorbed into the robot's execution graph).  Dissolving
the singularity = resolving the sword (condition~(1) deactivates
at release).
\end{remark}

\begin{remark}[Three-level nesting]\label{rem:m-threelevel}
The bi-level pick--place sits inside the two-level architecture
of \cref{sec:m-plangen}:
\begin{center}
\begin{tabular}{@{}cllp{5cm}@{}}
\toprule
\textbf{Level} & \textbf{Problem} & \textbf{Decides}
  & \textbf{Complexity} \\
\midrule
0 & Task planner & Which block, which order
  & $O(N!) \to O(K \cdot N)$ via MPPI \\
1 & Trajectory optimisation & Object path $X(t) \in SE(3)$
  & Smooth, gradient-based \\
2 & Force closure & Finger forces $f(t)$
  & Algebraic, $O(1)$ per timestep \\
\bottomrule
\end{tabular}
\end{center}
Level~2 is the fastest (a single inequality check).  Level~1 is
smooth (differentiable through the ODE).  Level~0 is
combinatorial (sampling-based).  The hierarchy matches the
detection--actuation separation: Level~0 = detection
$\Obs$ (what to manipulate), Level~1 = actuation $C(x)$
(how to move it), Level~2 = viability check (can we hold it?).
\end{remark}

\begin{remark}[Connection to the forcing lemma]
\label{rem:m-forcing}
The inner problem is the manipulation instantiation of
\cref{lem:forcing}.  If the gripper must maintain force closure
indefinitely under stochastic wind, the expected cumulative
friction energy diverges---the wind continuously consumes the
friction margin $2\mu_s N - m|g| - |F_\mathrm{wind}^\perp|$,
and the actuator must continuously replenish it.
\Cref{lem:forcing} guarantees that this cost is unbounded
over infinite time, forcing the outer trajectory to complete
the transport in finite time.  The bi-level structure makes
this forcing mechanism precise: Level~2's shrinking margin
under wind is the physical instantiation of the forcing
lemma's divergent cost.
\end{remark}

\subsection{Friction as subdifferential inclusion}
\label{sec:m-friction-di}

The Coulomb friction law unifies stick, slip, and lift-off
into a single set-valued inclusion, eliminating the need for
explicit mode enumeration.

Let $v_{t,j} = J_{t,j}(q)\,v$ be the tangential relative
velocity at contact~$j$.  The subdifferential of the norm
$\|\cdot\|$ is:
\[
  \partial \|v_t\| =
  \begin{cases}
    \{v_t / \|v_t\|\}, & v_t \neq 0, \\
    \{s : \|s\| \leq 1\}, & v_t = 0.
  \end{cases}
\]
The maximal-dissipation friction law is then:
\begin{equation}\label{eq:m-friction-di}
  f_{t,j} \in -\mu_j\,f_{n,j}\;\partial\|v_{t,j}\|.
\end{equation}
When $v_{t,j} \neq 0$ (slip), the inclusion selects the unique
friction force opposing motion.  When $v_{t,j} = 0$ (stick),
the inclusion yields the full friction disc
$\|f_{t,j}\| \leq \mu_j f_{n,j}$---static friction is
\emph{set-valued}, not indeterminate.

Substituting into the robot dynamics
(\cref{def:m-bilevel}) gives the contact DI:
\begin{equation}\label{eq:m-contact-di}
  M(q)\,\dot{v} \in \tau - h(q,v)
  + J_n(q)^\top f_n + J_t(q)^\top f_t,
  \quad
  (f_n, f_t) \in \mathcal{F}(q, v),
\end{equation}
where $\mathcal{F}$ encodes non-penetration ($g(q) \geq 0$,
$f_n \geq 0$, complementarity) and friction
(\cref{eq:m-friction-di}).

\begin{remark}[Soft--hard limit]\label{rem:m-soft-hard}
The regularised form
$f_t \approx -\mu f_n\,v_t / \sqrt{\|v_t\|^2 + \epsilon^2}$
is differentiable and simulation-friendly.  As
$\epsilon \to 0$, it recovers the set-valued law
\eqref{eq:m-friction-di}.  This is the tangential analogue of
the normal-direction soft--hard limit at the switching surface
$\Sigma = \{\rho = 1\}$: as the contact stiffness grows,
the soft penalty recovers the rigid non-penetration constraint.
The bi-level optimiser (\cref{def:m-bilevel}) uses the soft
form during trajectory optimisation and verifies force closure
against the hard law at convergence.
\end{remark}


% ═══════════════════════════════════════════════════════════
\section{The robot observer: causal observability}
\label{sec:m-observer}

\subsection{The cognitive boundary}

A camera mounted near the gripper (wrist camera) is the detector
$W_\tau$ of the causal observability framework.  The instantiation
to flat spacetime $\R^3 \times \R^1$ is:

\begin{center}
\begin{tabular}{@{}lp{8cm}@{}}
\toprule
\textbf{General} & \textbf{Manipulation} \\
\midrule
Spacetime $(M, g)$ & Task environment $\R^3 \times \R^1$, flat \\
Detector $W_\tau$ & Camera response at end-effector pose $\gamma(\tau)$ \\
Causal domain $D_\tau = J^-(S_\tau)$ & Field of view from current
  camera pose \\
Transfer matrix $T(\theta)$ & Rendering/projection operator (pinhole
  model + occlusion) \\
Source $X$ & Block states $\{g_i\} \in SE(3)^N$ \\
Measurement $Y$ & Image features (depth, RGB, segmentation) \\
\bottomrule
\end{tabular}
\end{center}

\begin{theorem}[Causal invisibility for manipulation]
\label{thm:m-invisible}
A block $B_k$ whose spatial support does not intersect the camera's
field of view at time $\tau$---i.e.,
$\mathrm{supp}(B_k) \cap D(q(\tau)) = \varnothing$---is exactly
invisible to the robot at that instant.  No computation applied to
the image $Y(\tau)$ can recover $g_k$.
\end{theorem}

\begin{proof}
The light rays from $B_k$ do not intersect the detector support
$S_\tau$ (the camera's sensor area in phase space).  The transfer
matrix $T(\theta)$ has a zero block for the invisible source.
This is a direct application of the causal invisibility theorem
to flat spacetime with finite-aperture detection.
\end{proof}

\subsection{Observation--action coupling}

Unlike a fixed external camera, the wrist camera moves with the
robot.  The detector support $S_\tau$ depends on the robot state:
\[
  D_\tau = D(q(\tau)) = \mathrm{FOV}(\gamma(q(\tau))).
\]
To observe block $B_k$, the robot must move to a pose where
$B_k \in D(q)$.  To plan a move, the robot needs observations.
This creates the \emph{active perception loop}: observe $\to$
plan $\to$ move $\to$ observe.

\begin{remark}[Detection bandwidth in manipulation]
\label{rem:m-bandwidth}
The detection function of \cref{prop:detection-derivation}
is physically realised by the camera:
\begin{enumerate}[label=(\roman*),nosep]
  \item Detection bandwidth $B$ = camera resolution $\times$ FOV
  solid angle: the maximum number of blocks simultaneously
  monitorable.
  \item Detection threshold $\tau(\Obs)$ = minimum block size
  (in pixels) for reliable pose estimation.
  \item $r \in \mathrm{Im}(\Obs) \iff$ block $r$ occupies
  $> \tau$ pixels in the current image.
\end{enumerate}
Mother--daughter consistency (\cref{rem:mother-daughter})
reduces here to: the predicted block pose (mother) matches
the pose that would result from one step of free-body dynamics
(daughter).  A large KL gap signals an unmodelled disturbance.
\end{remark}

\subsection{Information timeliness}

\begin{proposition}[Condition 4 for manipulation]
\label{prop:m-timeliness}
The control loop latency must satisfy
\begin{equation}\label{eq:m-timeliness}
  t_\pi \;=\; \underbrace{t_\mathrm{cam}}_{\text{capture}}
  + \underbrace{t_\mathrm{detect}}_{\text{perception}}
  + \underbrace{t_\mathrm{plan}}_{\text{planning}}
  \;\leq\; \Delta t,
\end{equation}
where $\Delta t$ is the control loop period.  Violation of
\eqref{eq:m-timeliness} causes the costate gradient to go stale
and controllability to be lost.
\end{proposition}

This is the manipulation instantiation of the ``high-latency
detection'' failure mode in the domain of applicability table
(\cref{sec:domain}).

\subsection{Three sensory modalities}

\begin{center}
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Modality} & \textbf{Detects} & \textbf{Bandwidth}
  & \textbf{Latency} \\
\midrule
Vision (wrist camera) & Block pose $g_i \in SE(3)$
  & High (megapixels) & Medium (${\sim}30$ ms) \\
Proprioception (encoders) & Robot state $q, \dot{q}$
  & Medium (7 joints) & Low (${\sim}1$ ms) \\
Tactile (F/T sensor) & Contact forces $F_c$
  & Low (6 DOF) & Low (${\sim}1$ ms) \\
\bottomrule
\end{tabular}
\end{center}

Vision has the highest bandwidth but the highest latency.
Tactile has the lowest bandwidth but the lowest latency.  The robot
fuses all three, weighted by their respective
$\tau(\Obs)$ thresholds---the multi-channel version of the
detection function (\cref{eq:detection}).


% ═══════════════════════════════════════════════════════════
\section{The plan generator}\label{sec:m-plangen}

\subsection{Plans and feasibility}

\begin{definition}[Plan]\label{def:m-plan}
A \emph{plan} for $N$ blocks is an ordered list
\[
  P = [(b_1, g_1, \gamma_1, d_1), \;\ldots,\; (b_N, g_N, \gamma_N, d_N)],
\]
where $b_i$ is the block index, $g_i \in SE(3)$ the target pose,
$\gamma_i$ the grasp type, and $d_i \in S^2$ the approach direction.
\end{definition}

\begin{definition}[Feasibility]\label{def:m-feasible}
A plan $P$ is \emph{feasible} if it satisfies four constraints
simultaneously:
\begin{enumerate}[label=(F\arabic*),nosep]
  \item \textbf{Topological:} if $b_j$ rests on $b_k$ in the goal,
  then $b_k$ precedes $b_j$ in $P$.
  \item \textbf{Stability:} for every prefix $P_{1:i}$, the partial
  stack is statically stable:
  $\mathrm{CoM}(P_{1:i}) \in
  \mathrm{ConvHull}(\mathrm{support}(P_{1:i}))$.
  \item \textbf{Reachability:} the robot can reach
  $(g_i, \gamma_i, d_i)$ without colliding with $P_{1:i-1}$.
  \item \textbf{Observable:} block $b_i$ lies within the camera's
  field of view from the pre-grasp pose:
  $b_i \in D(q_{\mathrm{pre},i})$.
\end{enumerate}
\end{definition}

Condition~(F4) is the link to the cognitive boundary
(\cref{thm:m-invisible}): an invisible block cannot be grasped.

\subsection{The two-level architecture}

\begin{definition}[Execution DAG]\label{def:m-dag}
The \emph{execution DAG} $G_{\mathrm{task}} = (V, E)$ is the
directed acyclic graph whose vertices $V = \{v_1, \ldots, v_N\}$
are sub-goals (block placements) and whose edges $E$ encode
precedence: $(v_i, v_j) \in E$ iff sub-goal~$v_i$ must be
completed before $v_j$ can begin.
For each vertex~$v$, write $\operatorname{pa}(v) = \{u : (u,v) \in E\}$
(parents) and $\operatorname{ch}(v) = \{w : (v,w) \in E\}$ (children).
A \emph{valid plan} is a topological order
of~$G_{\mathrm{task}}$.
\end{definition}

\begin{remark}[DAG layers as mother--daughter chains]
\label{rem:m-dag-md}
Each topological layer of $G_{\mathrm{task}}$ instantiates the
mother--daughter consistency of \cref{rem:mother-daughter}:
the \emph{mother} distribution at vertex~$v_j$ is the
predicted state before executing sub-goal~$j$; the
\emph{daughter} is the pushforward through the manipulation
action.  An edge $(v_i, v_j)$ means ``$v_i$'s daughter is
$v_j$'s mother''---the post-condition of one sub-goal is the
pre-condition of the next.  The DAG's depth is the length of
the longest such chain, and DAG-wide consistency holds when
every edge satisfies the mother--daughter consistency of
\cref{rem:mother-daughter} (small KL divergence between
predicted and actual distributions).  When this layered structure is
exchangeable, the chain becomes a causality fractal
(\cref{def:causality-fractal}): under the regularity
conditions of \cref{thm:fractal-convergence}, ergodic
convergence and event controllability
(\cref{cor:event-controllability}) follow from the spectral
gap $\lambda_1 > 0$.
\end{remark}

\begin{theorem}[Intractability of single-level planning]
\label{thm:m-intractable}
The joint plan--execute optimisation over $N$ blocks has
complexity $\Omega(N!)$ in the number of candidate orderings.
No deterministic algorithm solves both levels in polynomial time.
\end{theorem}

The two-level architecture breaks the problem:

\begin{definition}[Two-level architecture]\label{def:m-twolevel}
\leavevmode
\begin{enumerate}[label=(\roman*),nosep]
  \item \textbf{Upper level} (discrete, stochastic): sample $K$
  candidate plans from valid topological orders; evaluate each via
  fast rollout; select by Boltzmann weighting
  $w^{(k)} = \exp(-J^{(k)}/\alpha)$.  This is MPPI
  (\cref{thm:3b-mppi}) applied to the execution DAG.
  \item \textbf{Lower level} (continuous, smooth): given a block
  assignment $(b_i, g_i)$, execute the five-phase cycle via a
  learned velocity field $v_\theta(x,t)$.  Force is recovered via
  inverse dynamics: $u = \mathcal{D}(x, v_\theta(x,t))$.
\end{enumerate}
\end{definition}

\begin{proposition}[Randomness for complexity]
\label{prop:m-exchange}
The two-level architecture exchanges exponential time--space
complexity for polynomial randomness:
\begin{align}
  \text{Upper level:}\quad & O(N!) \to O(K \cdot N),
    \label{eq:m-upper-complex} \\
  \text{Lower level:}\quad & O(\exp(\dim\,\mu_0)) \to O(M),
    \label{eq:m-lower-complex}
\end{align}
where $K$ is the number of MPPI samples and $M$ the number of
particles for the velocity field.  The total control-loop cost
per $\Delta t$ is
\[
  \underbrace{K \cdot N \cdot t_\mathrm{rollout}}_{\text{upper}}
  + \underbrace{t_\mathrm{forward}(\theta)}_{\text{lower}}
  \;\leq\; \Delta t.
\]
\end{proposition}


% ═══════════════════════════════════════════════════════════
\section{Training: measure matching is the only reward}
\label{sec:m-training}

\subsection{Why reward engineering is unnecessary}

The OT-control formulation (\cref{sec:3b-ocp}) transports an
initial distribution~$\mu_0$ (random feasible configurations
produced by PlanGen) to a target distribution~$\mu_T$ (the
corresponding goal configurations) via velocity field $v_\theta$.
The training objective is:
\begin{equation}\label{eq:m-loss}
  \mathcal{L}(\theta)
  = \mathrm{MMD}^2\!\bigl(\mu_T^{\mathrm{achieved}},\;
    \mu_T^{\mathrm{target}}\bigr),
\end{equation}
where $\mathrm{MMD}$ is the maximum mean discrepancy with a
characteristic kernel $k$ (e.g., Gaussian or Mat\'ern):
\[
  \mathrm{MMD}^2(\mu, \nu)
  = \mathbb{E}[k(x,x')] - 2\,\mathbb{E}[k(x,y)]
  + \mathbb{E}[k(y,y')],
  \quad x,x' \sim \mu,\; y,y' \sim \nu.
\]

\begin{proposition}[MMD as self-regulating reward]
\label{prop:m-mmd-reward}
The MMD loss~\eqref{eq:m-loss} automatically encodes three
reward properties without explicit engineering:
\begin{enumerate}[label=(\roman*),nosep]
  \item \textbf{No reward for known tasks.}  If $v_\theta$ already
  solves a configuration class, $\mathrm{MMD} \approx 0$ on that
  class and $\nabla_\theta \mathcal{L} \approx 0$---zero gradient,
  zero learning signal.
  \item \textbf{Reward for new tasks.}  On unsolved configurations,
  $\mathrm{MMD} > 0$ and the gradient points toward improvement.
  \item \textbf{Penalty for regression.}  If $v_\theta$ forgets a
  previously solved class, $\mathrm{MMD}$ increases on that class
  and the gradient penalises the regression.
\end{enumerate}
\end{proposition}

\begin{proof}
Property~(i): $\mathrm{MMD}^2(\mu, \nu) = 0 \iff \mu = \nu$
for characteristic kernels.  Near the minimum, the Hessian is
positive semi-definite, so the gradient vanishes.
Property~(ii): $\mathrm{MMD}^2 > 0$ implies non-zero gradient
(by smoothness of $k$).  The gradient direction is the kernel
mean embedding of $\mu_T^{\mathrm{target}} -
\mu_T^{\mathrm{achieved}}$.
Property~(iii): $\mathcal{L}$ is evaluated on the full
distribution $\mu_0$, which includes previously solved
configurations.  Any regression increases the integral.
\end{proof}

The plan generator (\cref{sec:m-plangen}) provides
\emph{persistent excitation} ($\Sigma_X \succ 0$,
\cref{thm:m-invisible}) by randomising goal configurations.
Together: PlanGen generates diverse $(X_0, X_T)$ pairs, and
MMD measures how well $v_\theta$ transports between them.
No reward shaping, no curriculum-aware bonuses, no progress
tracking.  \textbf{The measure matching is the only reward.}

\subsection{Training algorithm: conditional flow matching}

The velocity field $v_\theta$ is trained by \emph{conditional
flow matching} (Lipman et al., 2023):

\begin{definition}[Flow matching loss]\label{def:m-flow-loss}
Sample a random feasible plan $P$ from PlanGen.  For each block
$(b_i, g_i)$ in $P$:
\begin{enumerate}[label=(\arabic*),nosep]
  \item Sample initial configuration $X_0 \sim \mu_0$.
  \item Set target $X_T = g_i$ (from the plan).
  \item Sample time $t \sim \mathrm{Uniform}(0, T)$.
  \item Interpolate: $X_t = (1 - t/T)\,X_0 + (t/T)\,X_T$
  (optimal transport interpolation).
  \item Compute the conditional velocity:
  $u_t = (X_T - X_0)/T$.
  \item Loss:
  $\ell(\theta) = \|v_\theta(X_t, t) - u_t\|^2$.
\end{enumerate}
\end{definition}

\begin{remark}[Why flow matching, not RL]
\label{rem:m-why-flow}
The five-phase cycle has known structure: the switching surface
$\Sigma$ is known, the contact phases are deterministic
(given $\rho$), and the target $g_i$ is specified by the plan.
There is no exploration--exploitation dilemma: the plan tells
the robot \emph{what} to do, and the velocity field learns
\emph{how}.  Reinforcement learning solves the wrong problem
here---it searches for \emph{what} to do when the answer is
already given by PlanGen.  The only learning is the smooth
motor execution, which is a supervised regression from
$(X_t, t) \to u_t$.
\end{remark}

\subsection{Network architecture}

The velocity field $v_\theta : \R^d \times [0,T] \to \R^n$
maps (state, time) to desired joint velocities.

\begin{definition}[Phase-conditioned velocity network]
\label{def:m-network}
The input to $v_\theta$ is:
\[
  z = (q,\; \dot{q},\; g_\mathrm{target},\; t,\; \phi)
  \;\in\; \R^{7 + 7 + 6 + 1 + 5} = \R^{26},
\]
where $q \in \R^7$ (joint positions), $\dot{q} \in \R^7$
(joint velocities), $g_\mathrm{target} \in \R^6$ ($SE(3)$ target
as position + axis-angle), $t \in [0,T]$ (normalised time within
the current phase), and $\phi \in \{0,1\}^5$ (one-hot phase
indicator).

The output is $\dot{q}_\mathrm{des} \in \R^7$ (desired joint
velocities).  Joint torques are recovered via inverse dynamics:
$u = M(q)\ddot{q}_\mathrm{des} + C(q,\dot{q})\dot{q} + g(q)$,
using the known robot model.
\end{definition}

\begin{remark}[Architecture choice]\label{rem:m-arch}
For $d = 26$ input and $n = 7$ output, a multi-layer perceptron
(MLP) with sinusoidal time embedding suffices:
\[
  v_\theta(z) = W_3\,\sigma(W_2\,\sigma(W_1\,[z;\,
  \mathrm{emb}(t)] + b_1) + b_2) + b_3,
\]
where $\mathrm{emb}(t) = [\sin(\omega_k t), \cos(\omega_k t)]_{k=1}^{K}$
is the Fourier time embedding and $\sigma$ is a smooth activation
(SiLU).  The dimension is low enough that transformers and
diffusion architectures are unnecessary overhead.  The phase
indicator $\phi$ can either be concatenated (as above) or used as
a gating signal on the hidden layers.
\end{remark}

\begin{remark}[Why the architecture is minimal]
\label{rem:m-minimal}
The heavy lifting is done by the \emph{structure}, not the network:
\begin{enumerate}[label=(\roman*),nosep]
  \item PlanGen handles combinatorial complexity (upper level).
  \item The five-phase FSM handles mode switching.
  \item Inverse dynamics handles force computation.
  \item The known robot model $M(q)$ provides inertial structure.
\end{enumerate}
The network $v_\theta$ only learns the residual: the smooth
interpolation between configurations within a single phase.
This is a low-dimensional, smooth function---an MLP's natural
domain.
\end{remark}

\subsection{Risk-locked training and sim-to-real transfer}
\label{sec:m-rlmd}

The CFM velocity field (\cref{def:m-flow-loss}) is trained by
supervised regression in simulation.  Two additional layers
convert it into a deployable controller:

\paragraph{Risk-Locked Mirror Descent (RLMD).}
Rather than maximising expected reward, RLMD locks the failure
probability below a threshold~$\delta$ and optimises secondary
objectives (smoothness, energy) subject to this constraint.

Let $g(x, u)$ be a margin function (positive when safe,
negative at failure).  For each curriculum level~$\varepsilon_k$
(\cref{sec:m-curriculum}), RLMD:
\begin{enumerate}[label=(\roman*),nosep]
  \item Rolls out $N$ trajectories under perturbation
  $\varepsilon_k$;
  \item Computes the soft risk
  $\hat{R} = \frac{1}{N}\sum_i \sigma(-m_i / \beta)$,
  where $m_i = \min_t g(x_t^i, u_t^i)$ is the worst-case
  margin, $\sigma$ is the sigmoid, and $\beta > 0$ is a
  temperature parameter;
  \item Computes the action roughness
  $\hat{S} = \frac{1}{N}\sum_i M^2 \sum_t \|u_{t+1}^i
  - u_t^i\|^2$ (the Nyquist proxy of
  \cref{prop:event-lock}~(iii));
  \item Minimises
  $\hat{J}(\phi) = \hat{R}(\phi) + \lambda\,\hat{S}(\phi)$
  subject to the lock $\hat{R} \leq \delta$;
  \item Advances to $\varepsilon_{k+1}$ only after $\hat{R}
  \leq \delta$ holds for several consecutive epochs.
\end{enumerate}
The output is a policy $\pi_\phi$ whose failure probability
is locked below~$\delta$ across the full curriculum.
This is the training-time realisation of
\cref{prop:event-lock}: the four conditions are enforced by
construction (viability via the simulator, controllability via
the MLP capacity, realisability via the roughness penalty,
event closure via the risk lock).

\paragraph{Actuator net for sim-to-real transfer.}
The sim-to-real gap concentrates in the actuator channel
(motor dynamics, contact compliance, latency).  Rather than
fine-tuning the full policy, we freeze $\pi_\phi$ and train
a residual actuator net $A_\psi$:
\[
  u_{\mathrm{real}} = u_{\mathrm{nom}} + A_\psi(o_t,\,
  u_{\mathrm{nom}},\, s_{\mathrm{motor}}),
\]
where $s_{\mathrm{motor}}$ includes joint velocities, currents,
and temperatures.  The correction $\Delta u = A_\psi(\cdot)$
is constrained in amplitude (safety) and roughness (Nyquist),
and trained on binary event labels from real-world rollouts:
$y = 1$ (failure) or $y = 0$ (success), supplemented by
dense margin signals from the simulator
(\cref{rem:m-forcing}).

The actuator net preserves the theoretical structure:
the high-level policy $\pi_\phi$ embodies the five-phase
cycle (\cref{sec:m-phases}); the actuator net only corrects
the \emph{execution channel}, not the \emph{plan}.

\subsection{Deployment}

At deployment, the full pipeline is:
\begin{enumerate}[nosep]
  \item \textbf{Observe:} wrist camera $\to$ block detection $\to$
  $\{g_i\} \in SE(3)^N$ (\cref{sec:m-observer}).
  \item \textbf{Plan:} $P^* = \mathrm{MPPI}(\mathrm{PlanGen},
  K)$---sample $K$ plans, evaluate via fast rollout, select best
  (\cref{sec:m-plangen}).
  \item \textbf{Execute:} for each $(b_i, g_i)$ in $P^*$, integrate
  $\dot{q} = v_\theta(q, \dot{q}, g_i, t, \phi)$ and apply
  $u = \mathcal{D}(q, \dot{q}, \ddot{q})$.
  \item \textbf{Monitor:} track $\rho(t)$ from kinematics
  (\cref{sec:3b-force-elim}) and $\lambda_1$ from the contact
  Laplacian.  Phase transitions trigger FSM updates to $\phi$.
  \item \textbf{Return:} robot retracts to $q_0$
  (\cref{thm:m-loop}).  Repeat for next block.
\end{enumerate}

The forward pass of $v_\theta$ (a 3-layer MLP) is $O(1)$---a
single matrix multiply per control step.  The bottleneck is MPPI
at the upper level: $O(K \cdot N)$ per replan.  Replanning
frequency is task-dependent: once per block (static scenes) or
once per control step (dynamic scenes).

\begin{algorithm}[H]
\caption{Block stacking: camera image in, torque out}
\label{alg:m-pipeline}
\begin{algorithmic}[1]
\Require Scene with $N$ blocks, Franka Panda at $q_0$,
  goal configuration $\{g_i^*\}$
\Ensure All blocks at goal poses within tolerance $\varepsilon$
\State $\{g_i\} \gets \textsc{Detect}(Y)$
  \Comment{Wrist camera $\to$ CNN $\to$ block poses in $SE(3)^N$}
\State $G_{\mathrm{task}} \gets \textsc{BuildDAG}(\{g_i^*\})$
  \Comment{Dependency DAG from goal config}
\State $P^* \gets \textsc{MPPI}(G_{\mathrm{task}}, \{g_i\},
  \{g_i^*\}, K, \alpha)$
  \Comment{Sample $K$ plans, Boltzmann-weight by cost}
\For{$(b_k, g_k^*, \gamma_k)$ in $P^*$}
  \Comment{For each block in topological order}
  \State $\phi \gets \textsc{Approach}$
    \Comment{Phase 1: $\rho = 0$}
  \While{block $b_k$ not at $g_k^*$}
    \State $q, \dot{q} \gets \textsc{ReadEncoders}()$
    \State $\rho \gets \|F_{\mathrm{contact}}\| / (m_k g)$
      \Comment{Kinematic $\rho$ via inverse dynamics residual}
    \State $\lambda_1 \gets \lambda_1(L_{G_c})$
      \Comment{Spectral gap of contact graph Laplacian}
    \State $\phi \gets \textsc{FSM}(\phi, \rho, \lambda_1)$
      \Comment{Update phase: $\rho$ crossing $\Sigma$,
        $\lambda_1 \geq \varepsilon$}
    \State $z \gets (q,\, \dot{q},\, g_k^*,\, t,\, \phi)
      \in \mathbb{R}^{26}$
    \State $\dot{q}_{\mathrm{des}} \gets v_\theta(z)$
      \Comment{Learned velocity field (MLP forward pass)}
    \State $\ddot{q}_{\mathrm{des}} \gets
      (\dot{q}_{\mathrm{des}} - \dot{q}_{\mathrm{prev}})
      / \Delta t$
      \Comment{Finite difference}
    \State $\tau \gets M(q)\,\ddot{q}_{\mathrm{des}}
      + C(q,\dot{q})\,\dot{q} + g(q)$
      \Comment{Inverse dynamics}
    \State \textsc{ApplyTorque}$(\tau)$
  \EndWhile
  \State \textsc{RetractTo}$(q_0)$
    \Comment{Robot returns to home: $q(t_5) = q_0$}
\EndFor
\end{algorithmic}
\end{algorithm}

\subsection{Inference flow: camera image in, torque out}

The end-to-end inference pipeline at each control step $\Delta t$
is shown in \cref{fig:m-inference}.

\begin{figure}[H]
\centering
\begin{tikzpicture}[
  box/.style={draw, rounded corners, minimum height=0.85cm,
    minimum width=2cm, align=center, font=\small},
  data/.style={font=\small\itshape},
  arr/.style={-{Stealth[length=2mm]}, thick},
  fb/.style={-{Stealth[length=2mm]}, thick, dashed},
  step right/.initial=1.0cm,   % staircase offset per row
  row sep/.initial=1.5cm,      % vertical gap between rows
  node distance=0.5cm and 0.6cm
]
% ── Staircase anchors ──
\coordinate (R1) at (0,0.9);
\coordinate (R2) at (1.0,-1.5);
\coordinate (R3) at (2.0,-3.0);
\coordinate (R4) at (3.0,-4.5);

% ── Row 1: Perception ──
\node[box, fill=water!15] at (R1) (cam) {Wrist camera};
\node[data, right=of cam] (img) {$Y$};
\node[box, fill=water!15, right=of img] (det) {Block detector};
\node[data, right=of det] (poses) {$\{g_i\}$};

\draw[arr] (cam) -- (img);
\draw[arr] (img) -- (det);
\draw[arr] (det) -- (poses);

% ── Row 2: Planning ──
\node[box, fill=dao!15] at (R2) (plan) {PlanGen};
\node[data, right=of plan] (pstar) {$P^*$};
\node[box, fill=dao!15, right=of pstar] (fsm) {Five-phase\\FSM};
\node[data, right=of fsm] (phase) {$\phi$};

\draw[arr] (poses) -- ++(0,-0.55) -| (plan);
\draw[arr] (plan) -- (pstar);
\draw[arr] (pstar) -- (fsm);
\draw[arr] (fsm) -- (phase);

% ── Row 3: Execution ──
\node[box, fill=sword!15] at (R3) (vnet) {$v_\theta$ (MLP)};
\node[data, right=of vnet] (dqdes) {$\dot{q}$};
\node[box, fill=sword!15, right=of dqdes] (diff)
  {$\frac{d}{dt}$};
\node[data, right=of diff] (ddqdes) {$\ddot{q}$};

\draw[arr] (phase) -- ++(0,-0.55) -| (vnet);
\draw[arr] (vnet) -- (dqdes);
\draw[arr] (dqdes) -- (diff);
\draw[arr] (diff) -- (ddqdes);

% ── Row 4: Actuation ──
\node[box, fill=caution!15] at (R4) (invdyn)
  {Inverse dyn.};
\node[data, right=of invdyn] (tau) {$\tau$};
\node[box, fill=caution!15, right=of tau] (robot)
  {Franka Panda};

\draw[arr] (ddqdes) -- ++(0,-0.55) -| (invdyn);
\draw[arr] (invdyn) -- (tau);
\draw[arr] (tau) -- (robot);

% ── Sensor output ──
\node[data, below=0.5cm of robot] (enc)
  {$q, \dot{q}$\; (encoders)};
\draw[arr] (robot) -- (enc);

% ── Rho monitor: right of robot ──
\node[box, fill=water!10, right=1.2cm of robot] (rho)
  {$\rho$ monitor};

% ── Routing coordinates ──
% Junction below enc: shared start for both feedback paths
\coordinate (J) at ([yshift=-0.4cm]enc.south);
% Bottom-left corner: left of all boxes, below enc
\coordinate (BL) at
  ([xshift=-0.8cm]cam.west |- J);

% ── Feedback 1: enc → rho ──
%    Down to junction, right, up into rho.south
\draw[fb] (enc.south) -- (J) -| (rho.south);

% ── Feedback 2: rho → FSM ──
%    Straight up (far right of all content), left above row 2,
%    down into fsm.north — avoids crossing φ
\draw[fb] (rho.north)
  -- ([yshift=0.7cm]fsm.north -| rho.north)   % up to above fsm
  -- ([yshift=0.7cm]fsm.north)                 % left above row 2
  -- (fsm.north);                               % down into fsm

% ── Feedback 3: enc → vnet ──
%    left, up 1cm, right to vnet.x, up to vnet.south
\coordinate (P1) at ([yshift=-0.4cm]enc.south);
\coordinate (P2) at ([xshift=-0.8cm]cam.west |- P1);
\coordinate (P3) at ([yshift=2.5cm]P2);
\coordinate (P4) at (P3 -| vnet.south);
\draw[fb] (enc.south) -- (P1) -- (P2) -- (P3) -- (P4) -- (vnet.south);

% ── Layer labels (in the staircase margin) ──
\node[anchor=east, font=\footnotesize\bfseries, text=water]
  at ([xshift=-0.3cm]cam.west) {Perception};
\node[anchor=east, font=\footnotesize\bfseries, text=dao]
  at ([xshift=-0.3cm]plan.west) {Planning};
\node[anchor=east, font=\footnotesize\bfseries, text=sword]
  at ([xshift=-0.3cm]vnet.west) {Execution};
\node[anchor=east, font=\footnotesize\bfseries, text=caution]
  at ([xshift=-0.3cm]invdyn.west) {Actuation};

\end{tikzpicture}
\caption{Inference pipeline: camera image in, torque out.
  Four layers---perception (水), planning (道), execution (剑),
  actuation (器)---map a wrist-camera image to joint torques
  in one control step.  Dashed arrows denote feedback:
  proprioception closes the velocity loop; the $\rho$~monitor
  triggers phase transitions.}
\label{fig:m-inference}
\end{figure}

\begin{remark}[What is learned vs.\ what is known]
\label{rem:m-learned-known}
Only two components contain learned parameters:
\begin{enumerate}[label=(\roman*),nosep]
  \item The \textbf{block detector} (CNN): image $\to$ block
  poses.  Standard perception; pre-trained or trained on
  rendered data.
  \item The \textbf{velocity field} $v_\theta$ (MLP): state $\to$
  desired velocities.  Trained by flow matching
  (\cref{def:m-flow-loss}).
\end{enumerate}
Everything else---PlanGen, FSM, inverse dynamics, $\rho$ monitor,
the robot model $M(q)$---is \emph{known structure}.  The learned
components are small MLPs operating on low-dimensional inputs.
This is the force elimination theorem at work: by exploiting the
known physics, the learning problem collapses to smooth
function approximation.
\end{remark}


% ═══════════════════════════════════════════════════════════
\section{Curriculum: four orthogonal axes}\label{sec:m-curriculum}

The natural training curriculum has four independent complexity axes,
each testing a different capability:

\begin{center}
\begin{tabular}{@{}lp{3cm}p{3.5cm}p{3.5cm}@{}}
\toprule
\textbf{Axis} & \textbf{Tests} & \textbf{Complexity}
  & \textbf{Paper analogue} \\
\midrule
Layers $L$ & Stability, DAG depth & Vertical error amplification
  & Viability chain depth \\
Count $N$ & Mean field, MPPI & $N!$ orderings, $\bar{U}$ over $N$
  & Number of agents $n$ \\
Types $K_\mathrm{type}$ & Observer, identifiability
  & Symmetry-breaking & $\tau(\Obs)$ \\
\bottomrule
\end{tabular}
\end{center}

These axes are orthogonal: co-growing them confounds variables.
The curriculum explores one axis at a time:
\begin{enumerate}[label=(\roman*),nosep]
  \item Fix $L = 0$, $K_\mathrm{type} = 1$.  Grow
  $N = 1, 2, 3, 4$.  (Mean-field scaling.)
  \item Fix $K_\mathrm{type} = 1$.  Grow $L = 1, 2, 3$.
  (Stability and DAG depth.)
  \item Fix $N = 3$, $L = 1$.  Grow
  $K_\mathrm{type} = 1, 2, 3$.  (Observer and identifiability.)
  \item Joint growth.  (Integration test.)
\end{enumerate}

\begin{remark}[Block types test the observer]
\label{rem:m-types-observer}
With $K_\mathrm{type} = 1$ (identical cubes), the identifiability
quotient is maximal: all blocks lie in the same equivalence class,
and the robot need only track positions.  With
$K_\mathrm{type} > 1$ (cubes, cylinders, L-shapes), the robot
must distinguish types visually, and the execution DAG becomes
shape-dependent (an L-shape cannot support a cylinder on its
narrow end).  This is where the wrist camera's persistent
excitation (\cref{thm:m-invisible}) becomes essential.
\end{remark}

\begin{remark}[Connection to the paper's mean field]
\label{rem:m-curriculum-meanfield}
The curriculum levels map onto the paper's complexity hierarchy:
$N = 1$ (single sword, no mean field); $N = 2$ (mean field
$\bar{U}$ appears over 2 blocks); $N = 3$ (phase transition:
which block is the ``sword''?); $N \to \infty$ (full mean-field
regime of \cref{thm:meanfield}).
\end{remark}

\subsection{Adversarial robustness: 风 (wind)}

Introduce a stochastic wind field $\mathcal{W}(t)$ that applies
an external wrench to every placed block at each timestep:
\[
  F_{\mathrm{wind}}^{(k)}(t)
  = A_{\mathrm{wind}}(t)\, \hat{n}(t),
  \qquad
  A_{\mathrm{wind}} \sim
  \mathrm{Uniform}(0,\, F_{\max}),
  \quad
  \hat{n} \sim S^2,
\]
where $A_{\mathrm{wind}}$ is the gust magnitude and $\hat{n}$
is a random direction on the unit sphere.  In MuJoCo, this is a
single line: \texttt{data.xfrc\_applied[block\_id] = F\_wind}.

The wind is a sword in the exact sense of \cref{def:sword}:
\begin{enumerate}[label=(\roman*),nosep]
  \item \textbf{Autonomous actuation.}  $U_{\mathcal{W}} \neq
  \varnothing$: the wind applies force independently of the
  robot's control.  The robot cannot command the wind.
  \item \textbf{Observable.}  The wind's effect is observable
  through two channels: the $\rho$ monitor detects shifting
  contact forces ($\lambda_1$ drops as a block slides), and
  the wrist camera detects displaced blocks
  ($g_k(\tau) \neq g_k^*$).
\end{enumerate}

\begin{proposition}[Wind couples directly to $\rho$ and
$\lambda_1$]\label{prop:m-wind}
Let block $B_k$ be placed at $g_k^*$ with contact graph Laplacian
$L_{G_c}$ and spectral gap $\lambda_1$.  An external wrench
$F_{\mathrm{wind}}$ shifts the contact force balance:
\[
  \lambda_1'
  = \lambda_1
  - \frac{\|F_{\mathrm{wind}}^{\perp}\|}{m_k\, g},
\]
where $F_{\mathrm{wind}}^{\perp}$ is the component of
the wind force perpendicular to the support normal.
Three regimes:
\begin{enumerate}[label=(\alph*),nosep]
  \item $\lambda_1' \geq \varepsilon$: block stable.
  No action needed.
  \item $0 < \lambda_1' < \varepsilon$: block sliding.
  The FSM detects $\lambda_1 < \varepsilon$ and the robot
  may intervene (regrasp or reposition).
  \item $\lambda_1' \leq 0$: block toppled.  The contact graph
  disconnects, $\rho \to 0$ for the toppled block.  The robot
  detects the topple via the $\rho$ monitor and replans:
  $P^* \gets \mathrm{MPPI}(G_{\mathrm{task}},
  \{g_i(\tau)\}, \{g_i^*\}, K, \alpha)$ with updated block
  state.
\end{enumerate}
\end{proposition}

\begin{proof}
The wind wrench adds a lateral force to the contact equilibrium.
The contact graph Laplacian edge weights are
$w_{ij} = k_n \cdot \rho_{ij}$, where $\rho_{ij}$ is the local
contact force ratio.  The lateral wind reduces the normal force on
the leeward contact (the finger or table edge facing away from the
wind), decreasing $w_{ij}$ and hence $\lambda_1$.  When
$\lambda_1 \leq 0$, the graph disconnects: the block is no longer
in force closure and topples under gravity.  The $\rho$ monitor
detects this as a $\Sigma$-crossing ($\rho$ drops through~$1$),
triggering the FSM to re-enter Phase~1 for the affected block.
By \cref{alg:m-pipeline}, the DAG is recomputed from the current
block state; the toppled block re-enters as an unplaced node.
\end{proof}

\begin{remark}[Training with wind]
\label{rem:m-wind-training}
During training, the wind is a stochastic perturbation with
scheduled intensity:
\[
  F_{\max}(\text{epoch})
  = F_{\max}^{(0)}
  + (F_{\max}^{(1)} - F_{\max}^{(0)})
  \cdot \min\!\bigl(1,\;
  \text{epoch}/\text{epoch}_{\mathrm{anneal}}\bigr).
\]
Starting from $F_{\max}^{(0)} = 0$ (no wind) and annealing to
$F_{\max}^{(1)}$ (e.g., $0.5\,m_k g$, half the block's weight).
This is \emph{not} a new reward signal---the MMD loss
(\cref{eq:m-loss}) already penalises any deviation from
$\mu_T^{\mathrm{target}}$, including wind-induced deviations.
The wind's sole effect is to broaden the state distribution:
$\mu_0$ now includes partially toppled stacks.
The velocity field $v_\theta$ trained on this broadened $\mu_0$
learns to recover from mid-stack perturbations without any
reward engineering.

The wind intensity $F_{\max}$ is a fourth curriculum axis,
orthogonal to the three in \cref{sec:m-curriculum}:
\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Axis} & \textbf{Symbol} & \textbf{Tests} \\
\midrule
Layers & $L$ & Stability, DAG depth \\
Count & $N$ & Mean field, MPPI scaling \\
Types & $K_{\mathrm{type}}$ & Observer, identifiability \\
Wind & $F_{\max}$ & Robustness, recovery \\
\bottomrule
\end{tabular}
\end{center}
\end{remark}

\begin{remark}[Why the ambient dimension is $d = 4$]
\label{rem:m-d4}
Every object in this chapter---the robot's end-effector, each
block, the wrist camera's field of view---traces a worldline in
physical spacetime $\R^3 \times \R^1 = \R^4$.  The temporal
coordinate is the irreversible wall-clock time of
\cref{rem:di-temporal,rem:temporal}: the robot executes in
real time, the block falls under real gravity, and the camera
captures frames at real instants.

The Hausdorff dimension intersection criterion
(\cref{rem:d4-not-analogy}) applies directly.  The robot's
execution path and a given block's worldline are each curves of
Hausdorff dimension~$2$ (Brownian scaling under finite-bandwidth
control).  Two such curves generically intersect in $\R^d$ iff
$2 + 2 \geq d$.  At $d = 4$ the intersection is \emph{marginal}:
the robot cannot passively encounter the block---it must actively
steer toward it.

This is precisely why Phase~1 (approach) exists: the robot must
\emph{seek} the $\Sigma$-crossing.  The five-phase cycle
(\cref{sec:m-phases}) is the structure that the
$d_c = 4$ marginality forces.  In $d = 3$ (no time axis), the
robot and block would generically share a point; manipulation
would require no planning.  In $d = 5$ (extra spatial dimension),
intersection would be transient; stable grasps would be
impossible.  The physical spacetime $d = 3 + 1 = 4$ is the
critical dimension at which manipulation is both necessary and
possible---the same marginality that governs
\cref{thm:alien-triviality}.
\end{remark}

\begin{remark}[Four conditions for event-locking in manipulation]
\label{rem:m-event-lock}
The event-locking bound (\cref{prop:event-lock}) instantiates
in this chapter as follows.  Let $E$ = ``all blocks at goal
poses within tolerance~$\varepsilon$''.  The four conditions
are:
\begin{enumerate}[label=(\roman*),nosep]
  \item \emph{Viability} (\cref{sec:m-bilevel}):
  the viable set is the intersection of joint limits,
  non-penetration ($g(q) \geq 0$), and force closure
  ($\lambda_1 > 0$).  The tangential condition is satisfiable
  by the bi-level controller.
  \item \emph{Projected controllability}
  (\cref{sec:m-ranknull}): the RANK/NULL decomposition ensures
  that the task-space forces span the required directions;
  the Lie-algebraic rank condition holds generically for
  $n_{\mathrm{dof}} \geq 6$.
  \item \emph{Frequency-aligned realisability}
  (\cref{sec:m-rlmd}): the Nyquist roughness penalty
  $M^2 \sum \|u_{t+1} - u_t\|^2$ ensures that the learned
  policy is executable at control frequency~$M$.
  \item \emph{Event closure} (\cref{sec:m-training}):
  the MMD loss drives the terminal distribution toward the
  target; RLMD locks the failure probability
  $\hat{R} \leq \delta$.
\end{enumerate}
Reading the four conditions against the mass-gap equivalence
(\cref{thm:massgap}): (i)~viability is the open set
(道---the law that the viable set exists),
(ii)~controllability is $\lambda_1 > 0$
(法---the method that keeps the spectral gap open),
(iii)~Nyquist realisability is the discrete-time sampling
constraint (器---the instrument that executes),
(iv)~event closure is the target measure's invariance
(势---the configuration that locks the outcome).
The five-phase cycle (\cref{sec:m-phases}) is the trajectory
through which these four conditions are maintained and
discharged: viability~(i) persists throughout; controllability~(ii)
is established during approach; realisability~(iii) is enforced at
every control step; event closure~(iv) is verified at termination.
\end{remark}
