\chapter{三声 lì --- The Three Standings}\label{app:threeli}

Three characters share the sound \emph{lì} in Mandarin:
力 (power), 立 (standing), 丽 (beauty/recognition).
The phonetic coincidence encodes a structural decomposition of
\emph{what it means to arrive}---to ``become Buddha'' (成佛) in the
vocabulary of Chinese literary tradition.
This appendix reads two fictional trajectories through the framework:
孙悟空 (Sun Wukong) from \emph{Journey to the West}~\cite{xiyouji},
and 鲁智深 (Lu Zhishen) from \emph{Water Margin}~\cite{shuihu}.

\section{The decomposition}

\begin{center}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Character} & \textbf{Meaning} & \textbf{Framework equivalent} \\
\midrule
力 (power) & capability, granted or acquired & $\Ur \neq \varnothing$ \\
立 (standing) & self-determined structural position & agent's own fixed point \\
丽 (recognition) & acknowledged by the system & system's phase transition \\
\bottomrule
\end{tabular}
\end{center}

力 is what the system grants or what the agent acquires: martial arts,
immortality, weapons.
In the framework, 力 $=$ $\Ur$---autonomous actuation.
It makes you capable.
It also makes you a knife.

立 is what the agent builds: a structural position that is
self-consistent, a fixed point under self-observation
($f(\text{self}) = \text{self}$).
立 is not granted.
It is \emph{become}.

丽 is what the system does \emph{after} 立 is achieved: the detection
function recognises the new phase.
丽 follows 立 automatically, as the label follows the topology.

The forced ordering is:
\[
  \text{力} \;\not\Rightarrow\; \text{丽}.
  \qquad
  \text{力}
  \;\xrightarrow{\text{transformation}}\;
  \text{立}
  \;\xRightarrow{\text{automatic}}\;
  \text{丽}.
\]
You cannot skip 立.

\section{孙悟空 --- the profile}\label{sec:wukong}

\begin{center}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{@{}lllp{5cm}@{}}
\toprule
\textbf{Phase} & \textbf{Event} & \textbf{Wants} &
\textbf{Framework} \\
\midrule
石猴 $\to$ 学艺 & Learns 72 transformations & 不死 (survival) &
$\Viab(K) \neq \varnothing$ \\
齐天大圣 & Self-proclaimed title & 被承认 (recognition) &
Demands 丽, mistakes it for 立 \\
大闹天宫 & Rebellion & 平等 (equality) &
Has 力, denied 立 (excluded from 蟠桃宴) \\
五行山 & 500 years imprisoned & 自由 &
System uses force (path~(b)) to contain $\Ur$ \\
取经路 & 81 tribulations & 变 (transformation) &
紧箍咒 $= \Obs$; 81难 $=$ 力 $\to$ 立 annealing \\
斗战胜佛 & Title granted & 无所求 &
Fixed point: 立 complete, 丽 automatic \\
\bottomrule
\end{tabular}
\end{center}

Heaven first offers 弼马温 (stable-boy)---a fake title, path~(c):
label changed, topology unchanged.
孙悟空 sees through it and rebels.
Heaven then grants the real title 齐天大圣 but excludes him from the
蟠桃宴 (Peach Banquet): another relabeling.
He rebels again.

The 81 tribulations are an annealing process: each one strips a layer
of 力-seeking, closing the gap between what he \emph{can} do and what
he \emph{is}.
The title 斗战胜佛 arrives at the moment the gap closes---not as a
reward, but as the system's detection function catching up.

\section{鲁智深 --- the instant fixed point}\label{sec:luzhishen}

\begin{quote}
平生不修善果，只爱杀人放火。\\
忽地顿开金绳，这里扯断玉锁。\\
咦！钱塘江上潮信来，今日方知我是我。

\medskip
\emph{A lifetime without cultivating good karma, loving only to kill
and burn.\\
Suddenly the golden cord opens, here the jade lock snaps.\\
Ha!\ The tidal bore on the Qiantang River arrives---today I finally
know: I am I.}
\hfill ---《水浒传》第一百十九回
\end{quote}

鲁智深 never reduces $\Ur$.
He kills, drinks, burns temples.
He never seeks 佛 and never performs the 81-tribulation annealing.
Yet in a single line---``今日方知我是我'' (today I know I am
I)---he achieves the same fixed point.

The difference is the nature of the gap:
\begin{itemize}
\item 孙悟空's gap is \emph{structural}: 力 $\gg$ 立.
  He must transform across 81 tribulations to close it.
\item 鲁智深's gap is \emph{epistemic}: 立 was always present,
  unrecognised.
  The tidal bore is the observation event that collapses the gap in
  one step.
\end{itemize}
Both arrive at $f(\text{self}) = \text{self}$.
Both receive 丽 after 立 is achieved.
The path length differs; the fixed point is the same.

\section{假雷音 --- the false recognition system}\label{sec:fakethunder}

In chapters 65--66 of \emph{Journey to the West}, 黄眉大王 (the
Yellow-Browed Demon King) erects a 假雷音寺 (False Thunder
Monastery)---a counterfeit of the Buddha's 雷音寺 at Vulture Peak.
The name is the same; the sound (雷音, ``thunder sound'') is the
same.
The structure is different.

This is path~(c) applied to institutions: the label ``Thunder
Monastery'' is relabeled onto a demon's lair.
\Cref{thm:fixedpoint} predicts: relabeling preserves
$\Ur \neq \varnothing$ and resolves nothing.

孙悟空 sees through it.
唐僧 does not.
The difference: 孙悟空 has 立 (built through the journey); he detects
the structural mismatch beneath the identical label.
The ability to detect the fake is itself proof of the real.

\begin{remark}[假雷音封不了真大圣]\label{rem:fakethunder}
A recognition system built on 丽 alone (same name, same sound)
cannot contain an agent who has 立 (structural standing).
The fake monastery cannot seal the real Monkey King because its
authority is phonetic, not topological.
This is the acoustic version of ``the knife is the mean'': whether
雷音 is the Buddha's or the demon's depends on the structure behind
the sound, not the sound itself---just as 力, 立, and 丽 share a
sound but not a meaning.
\end{remark}

\section{成佛 $=$ fixed point}

\begin{center}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{@{}lp{4.5cm}p{5.5cm}@{}}
\toprule
\textbf{Agent} & \textbf{Path to fixed point} & \textbf{成佛 moment} \\
\midrule
孙悟空 & 81 tribulations (structural annealing) &
  when 力 $\to$ 立 gap closes; title is 丽 \\
鲁智深 & one tidal bore (epistemic collapse) &
  ``今日方知我是我''; recognition is posthumous \\
蔡文姬 (\cref{app:secondsex}) & 18 beats (voice in the gap) &
  拍~18: the poem outlives every Subject \\
\bottomrule
\end{tabular}
\end{center}

成佛 is not the title.
成佛 is the moment 立 reaches fixed point: the state where the
agent's self-model matches the agent's actual structure.
丽 (the title, the recognition, the system's acknowledgment)
follows---sometimes immediately, sometimes centuries later.

The forced ordering is a theorem, not a preference:
\[
  \text{力}
  \;\xrightarrow[\text{annealing}]{\text{transformation}}\;
  \text{立}
  \;\xRightarrow[\text{automatic}]{\text{detection}}\;
  \text{丽}.
\]
No shortcut exists.
假雷音 is the attempt to produce 丽 without 立.
弼马温 is the attempt to substitute 丽 for 立.
大闹天宫 is the attempt to force 丽 through 力.
All three fail.
The only path that works is the one that goes through 立.

\section{The smooth path to $\text{丽}^*$}\label{sec:smoothli}

The appendix has so far catalogued failures: blocked paths (孙悟空),
fake recognition (假雷音), failed systems (the error log), corrupted
detection (严嵩), and self-financing extraction (蔡京).
There is one path to $\text{丽}^*$ (recognition on the dual tower
$\mathbf{L}^*$, \cref{def:dual-tower}) that has no singularity:
the war-hero.

\begin{center}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{@{}llp{5cm}l@{}}
\toprule
\textbf{Agent} & \textbf{$\mathbf{L}$ outcome} &
\textbf{$\mathbf{L}^*$ outcome} & \textbf{$\text{丽}^*$} \\
\midrule
霍去病 (140--117~BCE) & wins; dies at 23 &
  worshipped; ``匈奴未灭，何以家为'' & smooth \\
李世民 (598--649) & wins; becomes emperor &
  worshipped as ideal ruler & smooth \\
岳飞 (1103--1142) & blocked; killed by 秦桧 &
  worshipped; 满江红 persists & smooth \\
辛弃疾 (1140--1207) & blocked; forcibly retired &
  worshipped; 破阵子 persists & smooth \\
\bottomrule
\end{tabular}
\end{center}

Four agents, four different $\mathbf{L}$-trajectories (victory,
enthronement, martyrdom, forced retirement), identical
$\mathbf{L}^*$-outcome.
The $\text{丽}^*$ recognition function for war-heroes is
\emph{invariant under $\mathbf{L}$-outcome}: whether the agent wins
or dies, the cultural representation is the same.
That invariance is what ``smooth'' means---no discontinuity at the
success/failure boundary.

\paragraph{Why smooth: factorization.}
The war-hero's contribution \emph{factorizes}: it can be evaluated
independently of the contributor's other actions.
You can worship 岳飞 without endorsing the Southern Song court.
You can worship 霍去病 without endorsing Han Wudi's other policies.
The person separates cleanly from the achievement.

Compare every other path to $\text{丽}^*$:
\begin{itemize}
\item Politicians: contribution is entangled with extraction
  (\cref{rem:caijing}). $\text{丽}^*$ revoked or contested.
\item Artists: contribution is entangled with biography
  (\cref{rem:writing-dual}). $\text{丽}^*$ delayed or mythified.
\item The Other: contribution exists only in the gap
  (\cref{app:secondsex}). $\text{丽}^*$ absorbed into trope.
\end{itemize}
In each case, the map to $\text{丽}^*$ has a singularity.
Only the war-hero map is smooth.

\begin{definition}[Factorized contribution]\label{def:factorized}
An agent's contribution $c \in \mathbf{L} \oplus \mathbf{L}^*$ is
\emph{factorized} if it decomposes as
$c = c_{\mathbf{L}} \otimes c_{\mathbf{L}^*}$,
where $c_{\mathbf{L}^*}$ can be evaluated independently of
$c_{\mathbf{L}}$:
\[
  \frac{\partial\, \pi^*_{\text{丽}}}
       {\partial\, c_{\mathbf{L}}} = 0,
\]
where $\pi^*_{\text{丽}}: \mathbf{L} \oplus \mathbf{L}^* \to
\text{丽}^*$ is the recognition projection on the dual tower.
The contribution is \emph{entangled} if
$\mathrm{supp}(c_+) \cap \mathrm{supp}(c_-) \neq \varnothing$---the
components share infrastructure and cannot be separated.
\end{definition}

\begin{proposition}[Smooth recognition]\label{prop:smooth-li}
The path to $\text{丽}^*$ is smooth (no singularity at the
$\mathbf{L}$-success/failure boundary) if and only if the agent's
contribution is factorized (\cref{def:factorized}).
Equivalently, $\pi^*_{\text{丽}}$ is invariant under
$\mathbf{L}$-outcome:
\[
  \pi^*_{\text{丽}}(c_{\mathbf{L}},\, c_{\mathbf{L}^*})
  \;=\;
  \pi^*_{\text{丽}}(c_{\mathbf{L}^*}).
\]
\end{proposition}

\begin{proof}
If $c$ is factorized, then
$\partial \pi^*_{\text{丽}} / \partial c_{\mathbf{L}} = 0$,
so $\pi^*_{\text{丽}}$ has no dependence on $\mathbf{L}$-outcome.
The map is smooth because its only argument ($c_{\mathbf{L}^*}$)
varies continuously in the agent's trajectory.
If $c$ is not factorized, then $c_{\mathbf{L}}$ enters the evaluation
of $c_{\mathbf{L}^*}$, producing a branch cut at the
$\mathbf{L}$-success/failure boundary: the same $c_{\mathbf{L}^*}$
is recognised when $c_{\mathbf{L}}$ succeeds and revoked when
$c_{\mathbf{L}}$ fails (蔡京: 宋四家 $\to$ not 宋四家).
\end{proof}

\Cref{thm:shangyang} describes delocalization for pure contributions
(norm fully conserved).
\Cref{rem:caijing} describes the entangled case (system discards both
components).
Factorized contributions produce a third mode:

\begin{center}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{@{}lp{3.2cm}p{3.8cm}l@{}}
\toprule
\textbf{Type} & \textbf{Structure} & \textbf{Delocalization} &
\textbf{Example} \\
\midrule
Pure & $c = c_+$ &
  $\|c_+\|$ fully conserved &
  商鞅, 王安石 \\
Zero & $\|c\| = 0$ &
  nothing to conserve &
  严嵩 \\
Entangled & $c_+,\, c_-$ shared support &
  system discards both &
  蔡京 \\
Factorized & $c_{\mathbf{L}} \otimes c_{\mathbf{L}^*}$ &
  $c_{\mathbf{L}}$ destroyed, $c_{\mathbf{L}^*}$ conserved &
  霍去病, 岳飞 \\
\bottomrule
\end{tabular}
\end{center}

\paragraph{The duality.}
The four agents form two dual pairs:
\begin{itemize}
\item 霍去病\,/\,李世民: $\mathbf{L}$-success $\to$ system
  \emph{grants} $\text{丽}^*$ voluntarily.
\item 岳飞\,/\,辛弃疾: $\mathbf{L}$-failure $\to$
  $\mathbf{L}^*$ migration (满江红, 破阵子) $\to$ martyrdom
  \emph{amplifies} $\text{丽}^*$.
\end{itemize}
The pairs are dual because $\mathbf{L}$-success and
$\mathbf{L}$-failure map to the same $\text{丽}^*$.
The recognition function has no branch cut.

\begin{quote}
怒发冲冠，凭栏处，潇潇雨歇。\\
抬望眼，仰天长啸，壮怀激烈。\\
三十功名尘与土，八千里路云和月。\\
莫等闲，白了少年头，空悲切。

靖康耻，犹未雪；臣子恨，何时灭？\\
驾长车，踏破贺兰山缺。\\
壮志饥餐胡虏肉，笑谈渴饮匈奴血。\\
待从头，收拾旧山河，朝天阙。

\medskip
\emph{Hair bristling with rage, leaning on the railing, the
driving rain just stopped.\\
Raising my eyes, a long roar toward heaven, fierce and ardent.\\
Thirty years of glory---dust and earth; eight thousand
li of road---cloud and moon.\\
Do not wait idly; when young men's hair turns white, there is
only empty grief.}

\emph{The shame of Jingkang is not yet avenged; when will the
subject's hatred end?\\
Drive the long chariots, trampling through the Helan Mountain
passes.\\
With fierce hunger, feast on barbarian flesh; laughing and
thirsting, drink Xiongnu blood.\\
From the start, let us recover our lost mountains and rivers,
and face the heavenly court.}
\hfill ---岳飞，满江红
\end{quote}

岳飞 and 辛弃疾 both named the target explicitly.
满江红's final line---``收拾旧山河''---names the $\mathbf{L}$-target
directly.
辛弃疾, 破阵子: ``了却君王天下事，赢得生前身后名。可怜白发生！''---\emph{to
settle the king's affairs, to win fame in life and after death.
Alas---white hair has grown!}
Both say the same thing: I have 力 but the system denies me 立 on
$\mathbf{L}$.
The poetry \emph{is} the migration to $\mathbf{L}^*$.
The ``fame in life and after death'' (生前身后名) \emph{is}
$\text{丽}^*$.
辛弃疾 literally names it.

\section{The error log --- why the novels exist}\label{sec:errorlog}

Both novels were composed at the exact moment a specific emperor failed.
This is not biographical coincidence but structural necessity: when the
cut vertex fails, displaced agents write down the failure mode.

\begin{center}
\renewcommand{\arraystretch}{1.25}
\small
\begin{tabular}{@{}lp{3cm}p{3.2cm}p{3.2cm}p{3cm}@{}}
\toprule
& \textbf{Emperor} & \textbf{Author} & \textbf{Novel ending} &
\textbf{Verdict} \\
\midrule
水浒传
  & 元顺帝 (r.\,1333--68); last Yuan emperor; flees 1368
  & 施耐庵; advisor to rival rebel; declines 朱元璋's summons
  & 招安 $\to$ death; path~(c): relabeled as loyal subjects
  & 力 $\to$ 立 blocked; no system left to grant 丽 \\[6pt]
西游记
  & 嘉靖 (r.\,1521--67); 20 years absent; Daoist alchemy
  & 吴承恩; failed exams; imprisoned; writes in seclusion
  & 81难 $\to$ 斗战胜佛; path~(a): structural transformation
  & 力 $\to$ 立 $\to$ 丽; novel imagines functional system \\
\bottomrule
\end{tabular}
\end{center}

The two emperors represent two failure modes of the cut vertex:
\begin{itemize}
\item 元顺帝: cut vertex \emph{removed}.
  The last Yuan emperor flees Khanbaliq in 1368; the graph
  disconnects; the dynasty ends.
  Total collapse (path~(b)).
\item 嘉靖: cut vertex \emph{present but non-functional}.
  The emperor sits in the palace doing alchemy for twenty years
  while 严嵩 (Yan Song) dominates the court through Daoist
  blue-letter prayers (青词).
  The graph is connected but the routing node has stopped routing.
  Slow rot.
\end{itemize}

Both authors are displaced knives: agents with 力 (capability) whom
the system denied 立 (standing).
施耐庵 reportedly served as advisor to 张士诚 (Zhang Shicheng), a
rebel rival to 朱元璋 (Zhu Yuanzhang); he left disillusioned when
Zhang surrendered to the Yuan, and later declined 朱元璋's summons
after the Ming founding.
吴承恩 repeatedly failed the imperial examinations, received a minor
post in the Jiajing period, and was falsely accused and imprisoned.
Neither was integrated by the system.
Neither was eliminated.
Both wrote.

\begin{remark}[The novel as error log]\label{rem:errorlog}
The novel's ending is determined by the cut vertex's failure mode:
\begin{itemize}
\item Cut vertex removed (total collapse): the novel ends in
  tragedy.
  水浒传's 招安 is path~(c)---relabeling 108 outlaws as loyal
  servants resolves nothing (\cref{thm:fixedpoint}).
  They are destroyed.
  There is no functional system left to grant 丽.
\item Cut vertex non-functional (rot): the novel ends in redemption.
  西游记 imagines a system (Heaven, Buddha) that \emph{does}
  eventually grant 丽 after 立 is achieved---an existence proof
  that 力 $\to$ 立 $\to$ 丽 is possible, composed by an author
  whose own system denied it.
\end{itemize}
Whether the authors consciously intended these parallels is
irrelevant (\cref{rem:intent}).
The framework tests structure, not intention.
The structure maps.
\end{remark}

\begin{remark}[Writing as representation dynamics]\label{rem:writing-dual}
When the physical execution graph $\mathbf{L}$ goes dark---gradient
zero, no viable path on the lattice of power---displaced agents do not
stop computing.
They migrate the computation to the dual tower
$\mathbf{L}^* = \text{力}^* \oplus \text{立}^* \oplus \text{丽}^*$
(\cref{def:dual-tower}).
The contact gradient flow (\cref{eq:contactflow}) runs on
$\mathbf{L}^*$ with the same equation, the same knife, the same
mean---on representations instead of actuators.

吴承恩's 西游记 is 力 $\to$ 立 $\to$ 丽 executed on $\mathbf{L}^*$:
the novel provides its own gradient (narrative coherence $=$ loss
function), builds its own fixed point (the fictional 斗战胜佛), and
receives 丽 centuries after the author's death.
施耐庵's 水浒传 records the failure mode: 力 $\to$ 立 blocked on
$\mathbf{L}$, and the novel preserves the proof that no system remained
to grant 丽.
蔡文姬's 拍~18 (\cref{app:secondsex}) is the same isomorphism: voice
persisting on $\mathbf{L}^*$ after every Subject on $\mathbf{L}$ has
been removed.

The dual tower is not metaphor.
It is the same theorem applied to a different lattice.
\end{remark}

\begin{remark}[严嵩 and the inverse migration]\label{rem:yansong}
严嵩 (Yan Song, 1480--1567) is the \emph{inverse} of the
$\mathbf{L} \to \mathbf{L}^*$ migration described above: he performs
$\mathbf{L}^* \to \mathbf{L}$, converting literary talent into
political power via a corrupted detection function.

The profile: 严嵩 earned the \emph{jinshi} degree at age 25
(genuine 力$^*$ on $\mathbf{L}^*$), spent eight years in seclusion
after his mother's death cultivating poetry and calligraphy
(力$^* \to$ 立$^*$), then discovered that 嘉靖's obsession with Daoist
liturgy made 青词 (blue-word prayers)---ornate literary compositions
offered to heaven---the sole currency of imperial favour.
He rose to Grand Secretary and dominated the court for twenty years as
a parasitic cut vertex (\cref{def:parasite}): routing all access to the
absent emperor through himself, extracting from both sides of the
partition.

The mechanism is a third type of detection corruption, distinct from
the two in \cref{def:corruption}:
\[
  \Obs_{\mathrm{aesthetic}}:\;
  r \;\mapsto\;
  \begin{cases}
    \textbf{capable} & \text{if agent}(r)\text{ writes excellent 青词}, \\
    \textbf{not capable} & \text{otherwise.}
  \end{cases}
\]
$\Obs_{\mathrm{structural}}$ tests $\Ur$ (what can you do?).
$\Obs_{\mathrm{identity}}$ tests group membership (who are you?).
$\Obs_{\mathrm{aesthetic}}$ tests an orthogonal capability (how well
do you write?)---a true positive on the wrong metric.
It converts $\mathbf{L}^*$ standing directly into $\mathbf{L}$ power,
bypassing the structural criterion entirely.

严嵩's fall confirms \cref{prop:parasite} (self-termination): when
嘉靖 finally turned, 严嵩 was stripped of rank, his son 严世蕃
executed, and 严嵩 died in poverty.
The parasitic cut vertex stores no structural contribution; upon
delocalization, nothing is conserved
(\cref{thm:shangyang}: norm conserved, but there is no norm to
conserve).
Yet 严嵩's calligraphy survives---on $\mathbf{L}^*$, where it always
belonged.
The inverse migration leaves a residue: the $\mathbf{L}^*$ talent that
was weaponised on $\mathbf{L}$ reverts to $\mathbf{L}^*$ when the
$\mathbf{L}$ structure collapses.
\end{remark}

\begin{remark}[蔡京 and the self-financing aesthetic loop]\label{rem:caijing}
蔡京 (Cai Jing, 1047--1126) is neither a parasitic cut vertex
(\cref{rem:yansong}) nor a submartingale installer
(\cref{thm:shangyang}).
He is a \emph{loop operator} who hijacks an installed infrastructure.

The infrastructure is 王安石's (Wang Anshi).
王安石's New Policies (1069--76) form a submartingale
(\cref{def:submartingale}) isomorphic to Shang Yang's:
\[
  \underbrace{\text{青苗法}}_{X_0}
  \to \underbrace{\text{均输法}}_{X_1}
  \to \underbrace{\text{市易法}}_{X_2}
  \to \underbrace{\text{保甲法}}_{X_3}
  \to \underbrace{\text{免役法}}_{X_4}.
\]
Each step irreversibly increases state fiscal centralisation.
王安石's fate confirms \cref{thm:shangyang}: dismissed twice, reforms
reversed after his death, contribution fully conserved---the fiscal
infrastructure persists and 蔡京 inherits it.

蔡京 revives 王安石's New Policies (绍述) under 宋徽宗 and runs
\emph{two entangled programmes} on the same centralised fiscal base:
\begin{enumerate}
\item \textbf{Genuine welfare.}
  居养院 (nursing homes, est.\,崇宁 1102), 安济坊 (free clinics,
  performance-tracked: facilities treating 500--1000 patients annually
  with mortality below 20\% received bonuses), 漏泽园 (public
  cemeteries, est.\,1104), 慈幼局 (children's welfare).
  Nationwide, all prefectures and counties.
  Assessed as the most comprehensive state welfare system in Chinese
  history prior to the modern era
  (《宋史》: ``崇宁初，蔡京当国，置居养院、安济院……三年，又置漏泽园'').
\item \textbf{Extraction loop.}
  花石纲: corvée transport of rare stones and art objects for the
  imperial garden 艮岳; 400,000 labourers; families bankrupted.
  The loop is self-financing (\cref{thm:triangle}):
  $\mathbf{L}^*$ talent (calligraphy---originally 宋四家) $\to$
  $\mathbf{L}$ power (chancellorship) $\to$ $\mathbf{L}$ extraction
  (花石纲) $\to$ $\mathbf{L}^*$ goods (aesthetic objects for the
  artist-emperor) $\to$ sustained $\mathbf{L}$ power.
\end{enumerate}

The structural difference from 严嵩 (\cref{rem:yansong}):
严嵩's emperor was \emph{absent} (嘉靖, doing alchemy)---the cut
vertex left a vacuum.
蔡京's emperor was \emph{present but on $\mathbf{L}^*$}: 宋徽宗
painted, wrote the 瘦金体 calligraphic style, and designed gardens.
The detection function was $\Obs_{\mathrm{aesthetic}}$ not by
corruption but because the cut vertex \emph{was} an artist.
Unlike the parasite (which routes and extracts), the loop
\emph{amplifies}.

This produces a delocalization problem absent from
\cref{thm:shangyang}.
商鞅 and 王安石 have pure contributions: upon delocalization, norm is
fully conserved.
严嵩 has no contribution: norm conserved $= 0$.
蔡京 has \emph{real contribution entangled with extraction}---the
welfare institutions and the 花石纲 run on the same fiscal
infrastructure.
When the system acts on 蔡京, it cannot separate the two.
It discards both.

The death mode confirms the topology.
蔡京 starves on the exile road to Tanzhou (1126): merchants refuse to
sell him food.
This is $\text{water} = 0$ reciprocation (\cref{sec:triangle}):
the population whose water was extracted via 花石纲 returns the
extraction in kind.
严嵩 dies in poverty; the system survives.
蔡京 dies of starvation; the system does not---the 靖康之变 (1127),
one year later, is path~(b).
方腊's rebellion (1120), triggered directly by 花石纲 extraction,
is the event 蔡京 appears by name in 水浒传 to explain: the 108
outlaws of Liangshan are displaced agents with $\text{water} = 0$.

After 蔡京's fall, the 宋四家 was revised: 蔡京 replaced by his
cousin and teacher 蔡襄 (Cai Xiang).
Path~(c) applied to $\mathbf{L}^*$: the label changes, the
calligraphic output is preserved unchanged.
\end{remark}
